---
title: "Biologically Informed Model Specification"
author: "Teo Sakel"
date: "2026-08-31"
format:
  html:
    toc: true
    toc-depth: 2
    html-math-method: mathjax
    theme: none
    minimal: true
    embed-resources: true
    df-print: kable
    # The minimal quarto::html vignette engine omits Quarto's layout CSS.
    include-in-header:
      text: |
        <style>
        .quarto-layout-row { display: flex; gap: 1rem; }
        .quarto-layout-cell { min-width: 0; }
        .quarto-layout-cell figure { margin: 0; }
        .quarto-layout-cell img { max-width: 100%; box-sizing: border-box; margin: 0; }
        </style>
vignette: >
  %\VignetteIndexEntry{Biologically Informed Model Specification}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4.5
)
```

```{r}
#| label: libraries
#| message: false

library(batchieR)
library(Matrix)
library(ggplot2)

set.seed(20260827)
```

## Model and biological information

`batchieR` fits Bayesian latent-factor models to combination screens. This
vignette develops six model specifications that use different parts of the same
biological information, then compares their predictions and candidate-plate
scores. For the basic fitting workflow, see
`vignette("introduction", package = "batchieR")`.

For cell $k$ and compound-dose treatments $i$ and $j$, the Gaussian observation
model is

$$
y_{kij} \mid \eta_{kij}, \tau \sim \mathcal{N}(\eta_{kij}, \tau^{-1}),
$$

with observation precision $\tau$ and linear predictor

$$
\eta_{kij} = \alpha + x_{kij}^{\mathsf T}\beta
    + W^{(0)}_k + V^{(0)}_i + V^{(0)}_j
    + \left(W^{(1)}_k\right)^\top\left(V^{(1)}_i + V^{(1)}_j + V^{(2)}_i\odot V^{(2)}_j\right).
$$

The systematic mean $\alpha + x_{kij}^{\mathsf T}\beta$ describes the overall
response level and measured feature effects. Five zero-centered Gaussian
components capture residual cell and treatment variation. Superscripts label
component types, subscripts index entities, and $\odot$ denotes element-wise
multiplication.

| Model argument | Symbol | Dimension | Role |
|:--|:--|:--|:--|
| `mean` | $\alpha + x_{kij}^{\mathsf T}\beta$ | scalar per observation | Overall level and optional metadata regression |
| `cell_offset` | $W^{(0)}$ | $K$ | Baseline differences among cells |
| `cell_factors` | $W^{(1)}$ | $K \times d$ | Latent characteristics of cells |
| `treatment_offset` | $V^{(0)}$ | $T$ | Treatment effects shared across cells |
| `treatment_main_factors` | $V^{(1)}$ | $T \times d$ | Cell-dependent treatment main effects |
| `treatment_interaction_factors` | $V^{(2)}$ | $T \times d$ | Cell-dependent nonadditive combination effects |

Here $K$ is the number of cells, $T$ is the number of compound-dose treatments,
and $d$ is the shared latent dimension. Monotherapies omit the second-treatment
and treatment-interaction terms. Responses in this example are logit-transformed
viabilities; `plogis()` maps predictions back to the viability scale.

We will start with two independent baselines, add measured features and dose
pooling, and then use graph and tree relationships to structure the latent
factors. All six candidates use rank two and a Gaussian response; they differ
in their mean, dose, and prior specifications.

| Model | Global mean | Dose model | Cell-factor structure | Treatment-factor structure | Treatment shrinkage |
|:--|:--|:--|:--|:--|:--|
| `default` | empirical intercept | categorical | IID | IID | horseshoe |
| `iid` | empirical intercept | categorical | IID | IID | global half-Cauchy |
| `features` | feature regression | categorical | IID | IID | global half-Cauchy |
| `nested_dose` | feature regression | nested | IID | IID | global half-Cauchy |
| `gmrf` | feature regression | nested | GMRF | GMRF | global half-Cauchy |
| `tree` | feature regression | nested | tree | tree | global half-Cauchy |

## Running example: biological annotations and hierarchies

Consider a combination screen with

- 8 *cell* lines in two *lineages*,
- 8 *compounds* in 4 *mechanism groups*, and
- 2 *doses*.

Cells from the same lineage may share biological characteristics, and compounds
with a common mechanism may affect similar processes. Besides these group
labels, we include numeric annotations describing individual cells and
compounds. `pathway_activity` represents a cell's activity in a biological
pathway, and `target_signature` summarizes a compound's target-related
characteristics.

Nuisance variables are included to illustrate the effect of irrelevant features
on model selection.  We also include `growth_rate` for cells and `physchem_pc`,
a summary of physicochemical properties, for compounds.

```{r}
#| label: entity-metadata

cells <- sprintf("C%02d", 1:8)
drugs <- sprintf("D%02d", 1:8)

cell_data <- data.frame(
    cell = cells,
    lineage = rep(c("lineage_A", "lineage_B"), each = 4),
    pathway_activity = c(-1.4, -1.0, -0.6, -0.2, 0.2, 0.6, 1.0, 1.4),
    growth_rate = c(-0.7, 0.8, -0.2, 0.4, 0.7, -0.8, 0.2, -0.4),
    stringsAsFactors = FALSE
)
compound_data <- data.frame(
    drug = drugs,
    mechanism = rep(sprintf("MOA_%d", 1:4), each = 2),
    target_signature = c(-1.5, -1.2, -0.6, -0.3, 0.3, 0.6, 1.2, 1.5),
    physchem_pc = c(0.6, -0.7, -0.4, 0.8, -0.8, 0.4, 0.7, -0.6),
    stringsAsFactors = FALSE
)
cell_data
compound_data
```

The hierarchy below organizes the biological setting used to generate the
synthetic screen: cells belong to lineages, compounds belong to mechanism
groups, and each compound is tested at two doses. The cell lineages and compound
mechanism groups each sit beneath a shared root.

```{r}
#| label: fig-example-hierarchies
#| fig-subcap: ["Cells", "Treatments"]
#| echo: false
#| fig-cap: "Generating hierarchy for the synthetic screen. Cells and compounds connect through shared roots; each compound has two nested doses."
#| layout-ncol: 2

draw_nodes <- function(x, y, labels, fill, cex = 0.75) {
    y <- rep_len(y, length(x))
    points(x, y, pch = 21, bg = fill, col = "grey30", cex = 2.5)
    text(x, y, labels, cex = cex)
}

plot.new()
plot.window(xlim = c(0.5, 8.5), ylim = c(0.5, 3.5))
cell_x <- seq_len(8)
lineage_x <- c(2.5, 6.5)
segments(4.5, 3, lineage_x, 2)
segments(rep(lineage_x, each = 4), 2, cell_x, 1)
draw_nodes(4.5, 3, "Cell root", "#FADDB7", 0.7)
draw_nodes(lineage_x, 2, c("Lineage A", "Lineage B"), "#D9EAF7", 0.7)
draw_nodes(cell_x, 1, cells, "white", 0.7)
title("Cells", line = 0.5)

plot.new()
plot.window(xlim = c(0.25, 8.75), ylim = c(0.25, 4.5))
drug_x <- seq_len(8)
mechanism_x <- seq(1.5, 7.5, by = 2)
dose_x <- as.vector(rbind(drug_x - 0.18, drug_x + 0.18))
segments(4.5, 4, mechanism_x, 3)
segments(rep(mechanism_x, each = 2), 3, drug_x, 2)
segments(rep(drug_x, each = 2), 2, dose_x, 1)
draw_nodes(4.5, 4, "Compound root", "#D9EAF7", 0.65)
draw_nodes(mechanism_x, 3, sprintf("MOA %d", 1:4), "#FADDB7", 0.65)
draw_nodes(drug_x, 2, drugs, "white", 0.65)
draw_nodes(dose_x, 1, rep(c("1", "2"), 8), "#E2F0D9", 0.55)
text(8.65, 1, "Dose", adj = 1, cex = 0.65)
title("Treatments", line = 0.5)
```


## Model Specification

### Latent dimension

The cell and treatment factors occupy a shared latent space. The `rank`
argument sets the number of coordinates in that space, and therefore the
number of columns in each factor matrix. We use `rank = 2` to keep the worked
example small and give all six candidates the same latent dimensionality.

### Gaussian components

Each latent component is conditionally Gaussian and can be written schematically as

$$
u \mid \rho,Q \sim \mathcal N(0, (\rho Q)^{-1}).
$$

`combo_gaussian_component()` separates the prior into

- `shrinkage`: which controls the magnitude of component values through $\rho$
- `structure`: which describes relationships among entities through $Q$.

Larger precision means smaller prior variation and stronger shrinkage toward zero.

Magnitude can be controlled globally or locally. A global scale is shared
across entities, with a separate scale for each latent dimension in factor
matrices. Local scales let individual entries vary in magnitude, allowing
some to remain close to zero while others take larger values. Global and local
scales can also be combined.

### `default` and `iid`: independent baselines

The `default` model specification is based on the original BATCHIE prior.
The cell offsets ($W^{(0)}$) use a simple `gamma_precision()` shrinkage.
The cell factors $W^{(1)}$ use multiplicative-gamma shrinkage, with a precision
for latent dimension $h$ given by

$$
\rho_h = \prod_{\ell=1}^h \delta_\ell, \qquad \delta_\ell \sim \operatorname{Gamma}(a,b).
$$

Later dimensions inherit all earlier precision increments. With the default
hyperparameters, this encourages stronger shrinkage in later dimensions,
although individual increments can be below one and precision need not increase
monotonically in every draw. This promotes a lower effective rank when the
response surface needs fewer dimensions than the chosen `rank`. Because
treatment main effects and interactions are multiplied by the corresponding cell
factors, shrinking a cell-factor dimension suppresses its contribution to both terms.
The matrices retain all their columns; shrinkage changes their effective
contribution rather than their dimensions.

The remaining three components (the treatment offsets, treatment main factors,
and treatment interaction factors) use horseshoe shrinkage by default, combining
global and local scales without progressively penalizing later dimensions. The
horseshoe's local scales allow some treatment components to remain large while
others are strongly shrunk.

All five components use `structure = iid()` ($Q=I$), so cells or treatments are
conditionally independent within each component. Lineage and mechanism annotations
do not enter these priors. We construct this default baseline with:

```{r}
#| label: default-model

default_model <- combo_model(rank = 2)
```

The horseshoe's local scales do not generalize trivially to dependent
structures, for which `batchieR` does not support horseshoe shrinkage. This
restriction concerns dependence among latent component values; a
feature-informed response mean can still be combined with independent
horseshoe components.

We therefore define a simpler `iid` model that keeps the independent structure
and both cell-component priors, but uses global half-Cauchy shrinkage for all
treatment components. This lets subsequent candidates change their structure
while retaining the same shrinkage family.

We use `update()` to derive each new specification from an existing model.
Only named arguments change; omitted settings are retained and the original
model is unchanged. Each supplied component replaces its previous specification
in full, so we specify both its structure and shrinkage. The result is validated
by `combo_model()` and can be fitted later with `fit_combo()`.

```{r}
#| label: compatible-iid-model

iid_model <- update(
    default_model,
    treatment_offset = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    )
)
```

### `features`: the systematic mean

Biological annotations provide information about baseline response differences
among cells and compounds. Pathway activity may help explain why some cells are
generally more sensitive, while target-related features may explain systematic
differences among compounds. Besides cell and compound annotations, experiment-level
covariates, such as assay batch or incubation time, can be used to account
for differences in experimental conditions. We collect all selected covariates
in a design matrix $X$, with one row per observation. Writing the row for an
observation as $x_{kij}^{\mathsf T}$, the predictor becomes

$$
\eta_{kij}
  = \alpha + x_{kij}^{\mathsf T}\beta
    + W^{(0)}_k + V^{(0)}_i + V^{(0)}_j + \text{factor terms}.
$$

This decomposition is analogous to a mixed-effects model. The intercept
$\alpha$ and covariate coefficients $\beta$ form the *fixed-effects* part: their
effects are shared across observations, with coefficients estimated from the
data. The cell and treatment offsets act like *random intercepts*, capturing
residual differences that the covariates do not explain. The latent-factor
terms describe cell-dependent treatment effects and interactions.

We regularize the feature coefficients using Gaussian priors, giving a Bayesian
analogue of ridge regression. With a zero prior mean and a common precision $\lambda > 0$,

$$
\beta \sim \mathcal N(0, \lambda^{-1}I).
$$

Larger $\lambda$ shrinks coefficients more strongly toward zero. This allows us
to include potentially informative annotations alongside nuisance features
without treating every apparent association as a large effect. The intercept
$\alpha$ is estimated under a flat prior and is not subject to this penalty.

The `mean` argument of `combo_model()` specifies the systematic part
$\alpha + x_{kij}^{\mathsf T}\beta$. Starting from the `iid` model, we replace
the default empirical mean with `formula_mean()`, which constructs the
regression from an R formula. Here `beta_precision = 1` sets the common ridge
precision $\lambda$ to one for all non-intercept coefficients, whose priors
are centered at zero by default.

The formula can use experiment-level covariates from the observation table,
cell annotations from `cell_data`, and compound annotations from `compound_data`.
Covariates are not automatically centered or scaled, so their units affect the
meaning of the coefficient priors. Scaling continuous covariates to unit
variance can make a common precision more meaningful.  See `?formula_mean` for
the supported prior specifications and formula rules.

```{r}
#| label: feature-model

feature_mean <- formula_mean(
    ~ pathway_activity + growth_rate + target_signature + physchem_pc,
    beta_precision = 1
)
feature_model <- update(iid_model, mean = feature_mean)
```

Cell features contribute once per observation; compound features contribute
once per present compound, so combinations receive the sum of both compounds'
contributions. The offsets remain as residual cell and treatment effects.
When covariates can explain the same differences as the offsets, their
separation depends more strongly on the priors.

The feature coefficients describe effects of the measured annotations on the
response mean. The latent coordinates in $W^{(1)}$, $V^{(1)}$, and $V^{(2)}$
still have sign and permutation symmetries, so they do not acquire individual
biological interpretations. Their zero-centered priors describe residual
variation around the feature-informed mean.

### `nested_dose`: sharing across doses

The `features` model still treats doses as unrelated categories. With
`nested(precision = 4)`, a treatment value at dose $r$ is modeled around a
shared compound parent:

$$
V_{ir} = \bar V_i + e_{ir}, \qquad e_{ir} \mid \rho \sim \mathcal N(0, (4\rho)^{-1} I)
$$

The same construction applies to $V^{(0)}$, $V^{(1)}$, and $V^{(2)}$. It reduces
uncertainty for a sparse or missing dose by borrowing from other doses of that
compound. The value `4` is relative to the component precision $\rho$ and
controls the strength of dose-to-parent pooling. It does not use numerical dose
distance, impose monotonicity, or fit a dose-response curve.

```{r}
#| label: nested-dose-model

nested_dose_model <- update(feature_model, dose = nested(precision = 4))
```

To use concentration distances as well, supply a dose correlation kernel:

```{r}
#| label: correlated-dose-model

correlated_dose_model <- update(
    nested_dose_model,
    dose = nested(
        precision = 4,
        correlation = dose_kernel(kernel = "matern32", length_scale = 1)
    )
)
```

This optional model keeps the same compound parents but replaces the independent
raw dose deviations with covariance $K_i/(4\rho)$ within compound $i$.
`dose_kernel()` obtains concentrations directly from the experiment design;
no additional metadata table is needed. By default it uses log10 concentration,
so a length scale of one means one concentration decade. The exponential kernel
(the default) supplies local pooling with a sparse precision, while `"matern32"`
assumes smoother dose dependence. Both allow nonmonotone responses and
interactions. Compound-parent relationships still come from each component's
structure; tree-backed models retain their path-variance normalization.

The length scale is fixed and shared across compounds and treatment components.
Log10 requires positive concentrations; `transform = "identity"` also accepts
already-transformed coordinates. Numerical dose values are required only when
correlation is enabled. This option retains the nested model's restriction to
shrinkage compatible with non-IID structures, as already used by
`nested_dose_model`. It borrows information for missing responses at treatments
included in the fitting design, but does not add prediction for entirely new
treatments. The comparisons below continue to use the original `nested_dose` model.

### `gmrf`: dependence among entities

The preceding models keep cell and compound residuals independent. We now start
from the `nested_dose` model and ask whether related entities should have
similar latent factors. A Gaussian Markov random field (GMRF) expresses these
relationships through a sparse precision matrix $Q$. For latent dimension $h$,

$$
u_{\cdot h} \mid \rho_h,Q \sim \mathcal N\left(0,(\rho_h Q)^{-1}\right).
$$

Here `structure` supplies $Q$ and `shrinkage` supplies $\rho_h$. We can describe
the dependence encoded by $Q$ as a graph: each node represents an entity, such
as a cell or compound, and an edge joins two entities whose latent factors are
conditionally dependent given the others. A zero off-diagonal entry $Q_{ij}$
means that no edge connects entities $i$ and $j$.

To construct $Q$ from biological information, we first choose which entities
to connect and assign non-negative edge weights describing how strongly their
latent factors should be encouraged to agree. In this example, we connect cells
within each lineage and compounds within each mechanism group. The same
approach can represent pathway or molecular-network relationships.

`batchieR` does not infer these connections from annotations: we specify the
graph and turn it into a matrix operator for `gmrf()`, which adds a diagonal
ridge to obtain $Q$. Here we use the combinatorial graph Laplacian

$$
L = D - A
$$

where the adjacency matrix $A$ stores the edge weights: $A_{ij}$ is the weight
between entities $i$ and $j$, or zero if they are not connected. The graph is
undirected, so $A$ is symmetric, and $D$ is a diagonal matrix containing $A$'s
row sums. For a vector $x$ of latent-factor values across entities, this gives
the penalty

$$
x^\top Lx = \frac{1}{2}\sum_{i,j}A_{ij}(x_i-x_j)^2
$$

Larger edge weights more strongly encourage neighbors to agree. The helper
below connects cells within lineages and compounds within mechanism groups.

```{r}
#| label: gmrf-structures

graph_laplacian <- function(nodes, edges) {
    adjacency <- sparseMatrix(
        i = match(edges[, 1], nodes),
        j = match(edges[, 2], nodes),
        x = 1,
        dims = rep(length(nodes), 2),
        dimnames = list(nodes, nodes)
    )
    adjacency <- adjacency + t(adjacency)
    operator <- Diagonal(
        length(nodes),
        x = rowSums(adjacency)
    ) - adjacency
    dimnames(operator) <- list(nodes, nodes)
    drop0(operator)
}

cell_edges <- cells |>
    split(cell_data$lineage) |>
    lapply(function(group) t(combn(group, 2))) |>
    do.call(what = rbind, args = _)

compound_edges <- drugs |>
    split(compound_data$mechanism) |>
    lapply(function(group) t(combn(group, 2))) |>
    do.call(what = rbind, args = _)

cell_gmrf <- gmrf(8 * graph_laplacian(cells, cell_edges), ridge = 0.5)
compound_gmrf <- gmrf(8 * graph_laplacian(drugs, compound_edges), ridge = 0.5)
```

The GMRF is an alternative to the generating hierarchy. It connects cells
directly within each lineage, without introducing lineage parents or a shared
root. The compound graph uses the same construction within each mechanism group.

```{r}
#| label: fig-gmrf-connectivity
#| echo: false
#| fig-format: "svg"
#| fig-width: 6
#| fig-height: 2.5
#| fig-cap: "Alternative GMRF specification for cell factors. Each lineage forms a fully connected group, with no connections between lineages."

gmrf_nodes <- data.frame(
    cell = cells,
    x = c(0.7, 1.7, 0.7, 1.7, 3.7, 4.7, 3.7, 4.7),
    y = rep(c(1.8, 1.8, 0.6, 0.6), 2)
)
# Draw exactly the edges used to construct cell_gmrf above.
from <- match(cell_edges[, 1], gmrf_nodes$cell)
to <- match(cell_edges[, 2], gmrf_nodes$cell)
gmrf_plot_edges <- data.frame(
    x = gmrf_nodes$x[from], y = gmrf_nodes$y[from],
    xend = gmrf_nodes$x[to], yend = gmrf_nodes$y[to]
)

ggplot(gmrf_nodes, aes(x = x, y = y)) +
    geom_segment(
        data = gmrf_plot_edges, aes(xend = xend, yend = yend),
        colour = "grey55", linewidth = 0.6
    ) +
    geom_point(
        shape = 21, fill = "#E2F0D9", colour = "grey30", size = 10, stroke = 0.4
    ) +
    geom_text(aes(label = cell), size = 3.5) +
    coord_fixed(xlim = c(0.25, 5.15), ylim = c(0.2, 2.2), expand = FALSE) +
    theme_void() +
    theme(plot.margin = margin(4, 4, 4, 4))
```

```{r}
#| label: gmrf-model

gmrf_model <- update(
    nested_dose_model,
    cell_factors = combo_gaussian_component(
        structure = cell_gmrf,
        shrinkage = multiplicative_gamma()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = compound_gmrf,
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = compound_gmrf,
        shrinkage = global_half_cauchy()
    )
)
```

Other operators, such as a symmetric normalized Laplacian, can encode different
prior assumptions about the same graph. The operator must be symmetric, and
$Q = \text{operator} + \text{ridge} I$ must be positive definite to define a
proper Gaussian prior. A positive ridge guarantees this when the operator is
positive semidefinite, as a graph Laplacian is. Here, $Q=8L+0.5I$: the multiplier
controls the edge penalty relative to the ridge, while component `shrinkage`
scales $Q$ as a whole.

Alternatively, `precision(Q)` accepts a complete precision matrix directly,
without adding a ridge; component `shrinkage` still scales the supplied
precision. See `?gmrf` and `?precision` for input requirements.

### `tree`: hierarchical relationships

We now return to the generating hierarchy introduced in the running example.  A
tree is a special graph structure: each non-root node has a single parent, and
each child is a Gaussian increment from that parent. Related leaves share
ancestral increments.  This gives a structured Gaussian prior on all nodes,
including internal latent parents; the leaf covariance reflects their shared
ancestry.

The `tree` candidate uses this hierarchy in place of the direct connections in
`gmrf`. Cell lineages structure $W^{(1)}$, and compound mechanisms structure
$V^{(1)}$ and $V^{(2)}$. Both candidates extend `nested_dose`, keeping its mean,
residual offsets, shrinkage families, and dose pooling.

`tree()` constructs the relative precision from a data frame containing `node`
and `parent` columns and, optionally, a positive `edge_length`. Each child is a
Gaussian increment from its parent. Edge precision is the inverse of edge length,
so shorter edges express stronger similarity, and leaves covary through their
shared ancestral path. Internal parents become latent Gaussian nodes. A valid tree requires:

- every modeled cell or compound must be a terminal node
- all nonmissing parents exist
- no cycles

Here, cells share lineage parents and compounds share mechanism parents:

```{r}
#| label: tree-structures

cell_tree <- data.frame(
    node = c(".cell_root", "lineage_A", "lineage_B", cells),
    parent = c(NA, ".cell_root", ".cell_root", rep("lineage_A", 4), rep("lineage_B", 4)),
    edge_length = c(1, 1, 1, rep(0.05, 8)),
    stringsAsFactors = FALSE
)
cell_tree <- tree(node, parent, cell_tree, edge_length = edge_length)

compound_tree <- data.frame(
    node = c(".drug_root", sprintf("MOA_%d", 1:4), drugs),
    parent = c(NA, rep(".drug_root", 4), rep(sprintf("MOA_%d", 1:4), each = 2)),
    edge_length = c(1, rep(1, 4), rep(0.05, 8)),
    stringsAsFactors = FALSE
)
compound_tree <- tree(node, parent, compound_tree, edge_length = edge_length)
```

The tree model assigns these structures to the three factor components.

```{r}
#| label: tree-model

tree_model <- update(
    nested_dose_model,
    cell_factors = combo_gaussian_component(
        structure = cell_tree,
        shrinkage = multiplicative_gamma()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = compound_tree,
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = compound_tree,
        shrinkage = global_half_cauchy()
    )
)
```

## Reproducible toy responses

We now introduce responses for the annotated screen using the `tree_model` as a generating model.
Its hyperparameters are set as following:

- feature-coefficient priors have means `0.42` for `pathway_activity`, `-0.28` for `target_signature`, and precision `1000`.
- Prior prediction fixes the intercept at `1.15`;
- observation precision has a gamma prior with shape `100` and rate `1.44`.

These are simulation assumptions, distinct from the priors used to fit the six candidates.

The design includes single-agent experiments at both doses and all four dose
combinations for each selected compound pair. Every configuration has three
replicate measurements, sharing the same latent mean but with independent
observation noise. The complete generation recipe is shown below:

<details>
<summary>Show the data-generation recipe</summary>

```{r}
#| label: synthetic-data-generation
#| eval: false

modeled_pairs <- c(
    # within pairs
    "D01:D02", "D03:D04",
    "D05:D06", "D07:D08",
    # cross pairs
    "D02:D03", "D04:D05",
    "D06:D07", "D01:D08",
    "D01:D03", "D03:D05",
    "D05:D07", "D01:D07"
)

screen <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    dose_1 = 1:2,
    drug_2 = c(NA_character_, drugs),
    dose_2 = c(NA_real_, 1:2),
    replicate = 1:3,
    stringsAsFactors = FALSE
)
is_single_drug <- with(screen, is.na(drug_2) & is.na(dose_2))
is_modeled_pair <- with(
    screen,
    !is.na(dose_2) & paste(drug_1, drug_2, sep = ":") %in% modeled_pairs
)
screen <- screen[is_single_drug | is_modeled_pair, , drop = FALSE]
rownames(screen) <- NULL
screen$configuration <- rep(seq_len(sum(screen$replicate == 1)), times = 3)

# Use the tree model with the generating mean and observation precision.
synthetic_model <- update(
    tree_model,
    mean = formula_mean(
        ~ pathway_activity + growth_rate + target_signature + physchem_pc,
        beta_mean = c(
            pathway_activity = 0.42,
            growth_rate = 0,
            target_signature = -0.28,
            physchem_pc = 0
        ),
        beta_precision = 1000
    ),
    family = gaussian_response(
        precision = gamma_precision(shape = 100, rate = 1.44)
    )
)

screen$modeled_response <- as.numeric(
    prior_predict(
        synthetic_model,
        screen,
        cell_data = cell_data,
        compound_data = compound_data,
        draws = 1,
        type = "response",
        seed = 20260828,
        intercept = 1.15
    )
)
screen$viability <- plogis(screen$modeled_response)
```

</details>

For reproducibility, every fit and score below uses a pre-generated screen that
ships with the package.

```{r}
#| label: load-pregenerated-screen

screen <- read.csv(
    system.file("extdata", "biologically_informed_models.csv",
                package = "batchieR", mustWork = TRUE),
    stringsAsFactors = FALSE
)
```

### Evaluation plates

We now define six plates and hide their responses. Later we use these same
plates in two ways:

1. to measure prediction error, and
2. as candidates for future experimental design

| Plate | Four unique configurations | Main question |
|:--|:--|:--|
| `interpolation` | familiar cells, compounds, and doses | Can the model fill an ordinary gap? |
| `dose_transfer` | a held-out dose of `D05+D06` | Does pooling across doses help? |
| `unobserved_cell` | two combinations in `C04` and `C08` | Can annotations and lineage transfer to cells with no responses? |
| `unobserved_compound` | `D07+D08` in four observed cells | Can annotations and mechanism transfer to a compound with no responses? |
| `related_combination` | `D01+D03` in four cells | Can information transfer from related compounds and combinations? |
| `cross_mechanism_discovery` | `D03+D05` in four cells | How does the model treat a new cross-mechanism combination? |

Each plate contains four unique configurations and two wells per configuration.
Training configurations have three observed replicate measurements. All three
measurements of each held-out configuration are masked during fitting;
prediction error uses its first replicate, once per unique configuration.
Candidate plates have two proposed wells per configuration, and PDBAL uses all
eight wells because replication changes the information supplied by the plate.

We add the new *unobserved* plates to the training data without response so that
the model is aware of the entities involved.

<details>
<summary>Show the plate definitions and response masking</summary>

```{r}
#| label: response-mask
# Helpers for identifying compounds and unordered compound pairs.

is_combination <- !is.na(screen$drug_2)
contains_drug <- function(data, selected) {
    data$drug_1 %in% selected |
        (!is.na(data$drug_2) & data$drug_2 %in% selected)
}
pair_key <- function(drug_1, drug_2) {
    ifelse(
        is.na(drug_2),
        drug_1,
        paste(pmin(drug_1, drug_2), pmax(drug_1, drug_2), sep = ":")
    )
}

# Match a proposed experiment to the first replicate of its configuration.
row_matches <- function(cell, drug_1, dose_1, drug_2 = NA_character_, dose_2 = NA_real_) {
    screen$cell == cell & screen$drug_1 == drug_1 & screen$dose_1 == dose_1 &
        if (is.na(drug_2)) {
            is.na(screen$drug_2)
        } else {
            screen$drug_2 == drug_2 & screen$dose_2 == dose_2
        }
}
candidate_indices <- function(configurations) {
    indices <- integer(nrow(configurations))
    for (index in seq_len(nrow(configurations))) {
        matched <- which(screen$replicate == 1 & row_matches(
            configurations$cell[index],
            configurations$drug_1[index],
            configurations$dose_1[index],
            configurations$drug_2[index],
            configurations$dose_2[index]
        ))
        stopifnot(length(matched) == 1)
        indices[index] <- matched
    }
    indices
}

# Build matched-dose combination configurations for candidate plates.
candidate_config <- function(cells, drug_1, drug_2, dose) {
    data.frame(
        cell = cells,
        drug_1 = drug_1,
        dose_1 = dose,
        drug_2 = drug_2,
        dose_2 = dose,
        stringsAsFactors = FALSE
    )
}

# Define each plate once, before adding replicate wells.
prediction_data <- list(
    interpolation = candidate_config(
        c("C01", "C02", "C05", "C06"),
        c("D01", "D03", "D05", "D06"),
        c("D02", "D04", "D06", "D07"),
        1
    ),
    dose_transfer = candidate_config(
        c("C01", "C02", "C05", "C06"), "D05", "D06", 2
    ),
    unobserved_cell = candidate_config(
        c("C04", "C04", "C08", "C08"),
        c("D01", "D03", "D01", "D03"),
        c("D02", "D04", "D02", "D04"),
        c(1, 2, 1, 2)
    ),
    unobserved_compound = candidate_config(
        c("C01", "C02", "C05", "C06"), "D07", "D08", c(1, 2, 1, 2)
    ),
    related_combination = candidate_config(
        c("C01", "C02", "C05", "C06"), "D01", "D03", c(1, 2, 1, 2)
    ),
    cross_mechanism_discovery = candidate_config(
        c("C01", "C02", "C05", "C06"), "D03", "D05", 1
    )
)
prediction_indices <- lapply(prediction_data, candidate_indices)

# The physical plates contain two wells for each unique configuration.
candidate_data <- lapply(prediction_data, function(data) {
    data[rep(seq_len(nrow(data)), each = 2), , drop = FALSE]
})

# Hide extra rows needed to make the cell, compound, dose, and combinations
# genuinely unobserved. The named plate rows are the rows evaluated below.
held_out_cell <- screen$cell %in% c("C04", "C08")
held_out_compound <- contains_drug(screen, "D08")
held_out_dose <-
    (screen$drug_1 %in% c("D05", "D06") & screen$dose_1 == 2) |
    (screen$drug_2 %in% c("D05", "D06") & !is.na(screen$dose_2) & screen$dose_2 == 2)
held_out_pair <- is_combination & pair_key(screen$drug_1, screen$drug_2) %in%
    c("D01:D03", "D03:D05")

screen$role <- "training"
screen$role[held_out_cell | held_out_compound | held_out_dose | held_out_pair] <-
    "supporting_holdout"
for (plate in names(prediction_indices)) {
    configurations <- screen$configuration[prediction_indices[[plate]]]
    screen$role[screen$configuration %in% configurations] <- plate
}
screen$response <- ifelse(
    screen$role == "training",
    screen$modeled_response,
    NA_real_
)

# Retain only the columns required by prediction and plate scoring.
prediction_columns <- c("cell", "drug_1", "dose_1", "drug_2", "dose_2")
prediction_data <- lapply(prediction_data, function(data) data[prediction_columns])
candidate_data <- lapply(candidate_data, function(data) data[prediction_columns])
stopifnot(
    all(vapply(prediction_data, nrow, integer(1)) == 4),
    all(vapply(candidate_data, nrow, integer(1)) == 8),
    all(vapply(prediction_indices, length, integer(1)) == 4)
)

# Inspect two experiments from each role before summarizing the masking design.
screen |>
    split(screen$role) |>
    lapply(head, 2) |>
    do.call(what = rbind, args = _) |>
    `rownames<-`(NULL) |>
    subset(select = c(prediction_columns, "viability", "role"))
table(screen$role)
```

</details>

## Fit and evaluate the six models

We fit all six models to the same observed rows. Sampler mechanics are covered
in the introduction.  For chain diagnostics, posterior predictive checks, and LOO
comparisons, see `vignette("model_checking", package = "batchieR")`. Here we
focus on the consequences of the six model specifications. Each fit uses two
short chains to keep the vignette quick to build. Increase warmup and sampling
and check convergence before using these comparisons for scientific conclusions.

```{r}
#| label: fit-models

models <- list(
    default = default_model,
    iid = iid_model,
    features = feature_model,
    nested_dose = nested_dose_model,
    gmrf = gmrf_model,
    tree = tree_model
)
set.seed(20260900)
fits <- models |>
    lapply(
        fit_combo,
        data = screen[c(prediction_columns, "response")],
        cell_data = cell_data,
        compound_data = compound_data,
        chains = 2,
        parallel_chains = 1,
        iter_warmup = 250,
        iter_sampling = 250,
        thin = 1,
        refresh = 0
    )
```

### Metadata-effect estimates

The metadata formula affects the global response mean, and its coefficients are
on the raw feature scale. We compare their posterior distributions with the
prior means used to generate the synthetic responses: `0.42` for `pathway_activity`,
`-0.28` for `target_signature`, and zero for the two nuisance features. Orange
crosses mark these generating prior means.

```{r}
#| label: feature-coefficients

feature_map <- fits$features |>
    parameter_map() |>
    subset(component == "mean" & parameter == "coefficient")

feature_map$generating_prior_mean <- c(
    pathway_activity = 0.42,
    growth_rate = 0,
    target_signature = -0.28,
    physchem_pc = 0
)[feature_map$feature]

fits$features |>
    posterior_draws(select = feature_map$variable) |>
    bayesplot::mcmc_areas(
        prob = 0.8,
        prob_outer = 0.95,
        point_est = "median"
    ) +
    scale_y_discrete(labels = setNames(feature_map$feature, feature_map$variable)) +
    geom_vline(xintercept = 0, colour = "grey60", linetype = "dashed") +
    geom_point(
        data = feature_map,
        aes(x = generating_prior_mean, y = variable),
        inherit.aes = FALSE, shape = 4, size = 3, stroke = 1, colour = "#D55E00"
    ) +
    labs(x = "Feature coefficient", y = NULL)
```

Look for posterior estimates near the generating prior means, with the two
nuisance effects near zero. The interval widths show how precisely each effect
is estimated. Features can still help prediction even when their individual
coefficients are uncertain.

### Held-out plate prediction

For each plate, RMSE compares the posterior mean prediction with the held-out
observed responses on the fitted logit scale (`modeled_response`). These
responses are hidden during fitting. This evaluates prediction of measured
outcomes, so the error includes observation noise as well as error in the
estimated mean. We evaluate the first held-out replicate of each unique
configuration; all three replicates are excluded from fitting.

```{r}
#| label: prediction-metrics
#| fig-alt: "Held-out response RMSE by model and plate"

prediction_metrics <- data.frame(
    model = rep(names(fits), each = length(prediction_data)),
    plate = rep(names(prediction_data), times = length(fits)),
    n_configurations = rep(vapply(prediction_data, nrow, integer(1)), times = length(fits)),
    response_rmse = NA_real_
)

for (i in seq_len(nrow(prediction_metrics))) {
    model <- prediction_metrics$model[i]
    plate <- prediction_metrics$plate[i]
    expected <- posterior_epred(fits[[model]], newdata = prediction_data[[plate]])
    truth <- screen$modeled_response[prediction_indices[[plate]]]
    prediction_metrics$response_rmse[i] <- sqrt(mean((colMeans(expected) - truth)^2))
}

prediction_metrics$model <- factor(prediction_metrics$model, levels = rev(names(models)))
prediction_metrics$plate <- factor(prediction_metrics$plate, levels = names(prediction_data))

ggplot(prediction_metrics, aes(x = response_rmse, y = model, fill = model)) +
    geom_col(show.legend = FALSE) +
    scale_fill_brewer(palette = "Set2") +
    facet_wrap(vars(plate)) +
    labs(x = "Held-out response RMSE (logit scale)", y = NULL) +
    theme_bw()
```

Read each comparison locally. The feature model can help the two unobserved-
entity plates, nested doses target `dose_transfer`, and the tree and GMRF can
help when lineage or mechanism relationships are useful. These are results from
one synthetic draw, so the plot illustrates model behavior rather than ranking
the priors in general.

## Score candidate plates

To complete the workflow, we score the same six plates with PDBAL. Unlike
RMSE, the score does not use their hidden outcomes. It asks which complete
eight-well plate is expected to reduce posterior uncertainty across the
`r nrow(unique(screen[prediction_columns]))`-configuration reference surface.
Distances are measured after converting expected logit responses back to viability; the candidate likelihood remains
Gaussian on the fitted logit scale. Lower scores are preferred.

```{r}
#| label: score-candidates

score_rows <- vector("list", length(fits))
for (i in seq_along(fits)) {
    scores <- score_plate_pdbal(
        fits[[i]],
        candidate_data = candidate_data,
        reference_grid = unique(screen[prediction_columns]),
        response_transform = plogis,
        max_triplets = 1000
    )
    scores$model <- names(fits)[i]
    scores$normalized_score <- round(scores$score - min(scores$score), 3)
    score_rows[[i]] <- scores
}
plate_scores <- do.call(rbind, score_rows)
rownames(plate_scores) <- NULL

plate_ranks <- reshape(
    plate_scores[c("model", "plate", "normalized_score")],
    idvar = "plate", timevar = "model", direction = "wide"
)
plate_ranks <- plate_ranks[match(names(candidate_data), plate_ranks$plate), , drop = FALSE]
names(plate_ranks) <- sub("normalized_score[.]", "", names(plate_ranks))
plate_ranks
```

The normalized score is zero for the preferred plate within each model. Only
values within a model should be compared because changing the model also
changes its posterior distances and predictive distributions.

```{r}
#| label: selected-plates

subset(plate_scores, rank == 1, select = c(model, plate))
```

## Conclusion

Fixed choices such as edge lengths, graph weights, dose precision, and the GMRF ridge
should be checked with prior-predictive simulation and sensitivity analysis.
They should come from external knowledge or plausible response behavior, rather
than being tuned to the same hidden outcomes used for evaluation. For a fuller treatment, see
[Betancourt's principled Bayesian workflow](https://betanalpha.github.io/assets/case_studies/principled_bayesian_workflow.html)
and [Gabry et al. on visualization in Bayesian workflow](https://academic.oup.com/jrsssa/article/182/2/389/7070184).

## Session information

```{r}
#| label: session-information

sessionInfo()
```
