---
title: "Biologically Informed Model Specification"
author: "Teo Sakel"
date: "2026-08-31"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Biologically informed model specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupg}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4.5
)
library(batchieR)
library(Matrix)
library(ggplot2)
set.seed(20260827)
```

## Model components and biological prior information

`batchieR` models a drug combination response as a linear predictor of cell and
compound latent factors. This vignette explains how biological knowledge can
enter the prior for those factors. It assumes familiarity with the basic fitting
and prediction workflow in `vignette("introduction", package = "batchieR")`;
details of fit objects, printing, and sampler use are intentionally omitted here.

For cell $k$ and compound-dose treatments $i$ and $j$, the linear predictor is

$$
\eta_{kij} = \alpha + W_{0k} + V_{0i} + V_{0j}
    + W_k^\top\left(V_{1i} + V_{1j} + V_{2i}\odot V_{2j}\right)
$$

The five configurable terms have distinct scientific meanings:

| Variable | Role | Information that can be useful |
|:--|:--|:--|
| $W_0$ | cell baseline                   | growth, lineage, or other cell-level annotations |
| $W$   | latent cell state               | relationships among cells |
| $V_0$ | additive treatment effect       | targets, chemical features, compound and dose relationships |
| $V_1$ | cell-dependent treatment effect | compound and dose relationships |
| $V_2$ | nonadditive combination effect  | compound and dose relationships |

The current implementation limits all five configurable terms to conditionally
Gaussian components. Each component can be written schematically as

$$
\theta = X\beta + u, \qquad
u \mid \rho,Q \sim \mathcal N(0, (\rho Q)^{-1})
$$

This separates three questions:

1. **Mean** ($X\beta$): can measured features explain systematic differences?
2. **Magnitude** ($\rho$): how strongly should unexplained deviations $u$ shrink?
3. **Dependence** ($Q$): what is the relationship among entities in $u$?

`combo_gaussian_component()` turns these choices into a component specification
for assignment to one of the five slots in `combo_model()`. Its arguments map
directly onto the decomposition above:

| Argument | Part of the prior |
|:--|:--|
| `mean`           | the systematic term $X\beta$ |
| `beta_precision` | regularization of the coefficients $\beta$ when the mean uses features |
| `shrinkage`      | the overall or entity-specific magnitude of the residual $u$ |
| `structure`      | the relative residual precision $Q$; `iid()` uses $Q=I$ |

For example, a feature-informed offset component with otherwise independent residuals
can be specified as

```{r gaussian-component-template}
example_component <- combo_gaussian_component(
    mean = ~ 0 + feature_1 + feature_2,
    beta_precision = gamma_precision(),
    structure = iid(),
    shrinkage = global_half_cauchy()
)
```

Like most latent-factorization methods, this model does not give individual
latent coordinates an intrinsic interpretation. The components $W$, $V_1$, and
$V_2$ enter the predictor only through products, so coordinated changes to their
representation can leave the likelihood unchanged. Consequently, coordinate-specific
nonzero means for these factors would depend on an arbitrary representation and
require additional identification constraints to be interpretable. It therefore
makes sense to define feature-informed means only for the scalar, identifiable
offset variables $W_0$ and $V_0$, while using covariance or precision structures
to encode biological relationships among entities in the factor components.

A fourth question applies to treatments: should doses be unrelated categories,
or deviations from a shared compound-level value?

## Running example: biological annotations and hierarchies

We use a small annotated screen with

- 8 *cell* lines in two *lineages*,
- 8 *compounds* in 4 *mechanism groups*, and
- 2 *doses*.

```{r entity-metadata}
cells <- sprintf("C%02d", 1:8)
drugs <- sprintf("D%02d", 1:8)

cell_data <- data.frame(
    cell = cells,
    lineage = rep(c("lineage_A", "lineage_B"), each = 4),
    pathway_activity = c(-1.4, -1.0, -0.6, -0.2, 0.2, 0.6, 1.0, 1.4),
    growth_rate = c(-0.7, 0.8, -0.2, 0.4, 0.7, -0.8, 0.2, -0.4),
    stringsAsFactors = FALSE
)
compound_data <- data.frame(
    drug = drugs,
    mechanism = rep(sprintf("MOA_%d", 1:4), each = 2),
    target_signature = c(-1.5, -1.2, -0.6, -0.3, 0.3, 0.6, 1.2, 1.5),
    physchem_pc = c(0.6, -0.7, -0.4, 0.8, -0.8, 0.4, 0.7, -0.6),
    stringsAsFactors = FALSE
)
cell_data
compound_data
```

```{r example-hierarchies, dev="svg", fig.width=7, fig.height=4.2, fig.cap="Hierarchies in the running example. Cells and compounds connect through shared roots; each compound has two nested doses."}
draw_nodes <- function(x, y, labels, fill, cex = 0.75) {
    y <- rep_len(y, length(x))
    points(x, y, pch = 21, bg = fill, col = "grey30", cex = 2.5)
    text(x, y, labels, cex = cex)
}

old_par <- par(mfrow = c(1, 2), mar = c(0.5, 0.5, 2, 0.5), xpd = NA)

plot.new()
plot.window(xlim = c(0.5, 8.5), ylim = c(0.5, 3.5))
cell_x <- seq_len(8)
lineage_x <- c(2.5, 6.5)
segments(4.5, 3, lineage_x, 2)
segments(rep(lineage_x, each = 4), 2, cell_x, 1)
draw_nodes(4.5, 3, "Cell root", "#FADDB7", 0.7)
draw_nodes(lineage_x, 2, c("Lineage A", "Lineage B"), "#D9EAF7", 0.7)
draw_nodes(cell_x, 1, cells, "white", 0.7)
title("Cells", line = 0.5)

plot.new()
plot.window(xlim = c(0.25, 8.75), ylim = c(0.25, 4.5))
drug_x <- seq_len(8)
mechanism_x <- seq(1.5, 7.5, by = 2)
dose_x <- as.vector(rbind(drug_x - 0.18, drug_x + 0.18))
segments(4.5, 4, mechanism_x, 3)
segments(rep(mechanism_x, each = 2), 3, drug_x, 2)
segments(rep(drug_x, each = 2), 2, dose_x, 1)
draw_nodes(4.5, 4, "Compound root", "#D9EAF7", 0.65)
draw_nodes(mechanism_x, 3, sprintf("MOA %d", 1:4), "#FADDB7", 0.65)
draw_nodes(drug_x, 2, drugs, "white", 0.65)
draw_nodes(dose_x, 1, rep(c("1", "2"), 8), "#E2F0D9", 0.55)
text(8.65, 1, "Dose", adj = 1, cex = 0.65)
title("Treatments", line = 0.5)

par(old_par)
```

The annotations represent prior information about the entities (cells and compounds)
involved like pathway scores, expression components, target annotations, or
other low-dimensional summaries. `growth_rate` and `physchem_pc` are included as
plausible nuisance covariates to illustrate regularized feature means.

## Model specifications

`combo_model()` separates the response surface into five configurable components:

1. the cell offset $W_0$,
2. cell factors $W$,
3. treatment offset $V_0$,
4. treatment main factors $V_1$, and
5. treatment interaction factors $V_2$.

We build six models. They are easiest to understand as three pairs:

1. `default` and `iid` are two IID baselines;
2. `features` and `nested_dose` add feature means and then dose pooling
   but otherwise leave the entities independent; and
3. `tree` and `gmrf` start from `nested_dose` and add two alternative
   dependence structures.

The models use the same rank-three Gaussian likelihood and differ only in their priors.

| Model | Feature means | Dose model | Cell-factor structure | Treatment-factor structure | Treatment shrinkage |
|:--|:--|:--|:--|:--|:--|
| `default` | none | categorical | IID | IID | horseshoe |
| `iid` | none | categorical | IID | IID | global half-Cauchy |
| `features` | offsets | categorical | IID | IID | global half-Cauchy |
| `nested_dose` | offsets | nested | IID | IID | global half-Cauchy |
| `tree` | offsets | nested | tree | tree | global half-Cauchy |
| `gmrf` | offsets | nested | GMRF | GMRF | global half-Cauchy |

### IID baseline models

#### Latent-dimension shrinkage

By assumption, cells and compounds share the same latent space and their
relative positions determine the mean response surface. The `rank` argument
controls the dimensionality of that latent space by setting an upper bound.
The default multiplicative-gamma prior promotes a lower effective rank ($d$)
by shrinking later dimensions more strongly than earlier ones:

$$
\rho_d = \prod_{h=1}^d \delta_h, \qquad \delta_h \sim \operatorname{Gamma}(a,b)
$$

Later dimensions inherit all earlier precision increments and are progressively
shrunk toward zero. This makes a moderately generous rank safer when the "true"
response surface is lower-dimensional.

```{r default-model}
default_model <- combo_model(rank = 3)
```

#### A structure-compatible IID baseline

The remaining default components priors are *horseshoe* to promote sparsity: a
global scale shrinks all entities, while local scales allow some entities to
move farther from zero. However, when dependencies are introduced (off-diagonal
entries of $Q$), the horseshoe trick does not generalize in a straightforward way
because local scales disrupt the covariance structure. `batchieR`therefore does
not provide a "generalized horseshoe" and uses only global shrinkage for structured components.

For a cleaner comparison with the structured models, we define an IID model with
just global half-Cauchy shrinkage. Later models keep this shrinkage family while
changing the mean, dose model, or dependence structure. We retain the default
horseshoe priors as a second baseline for comparison. In both baselines, means
remain zero, structures remain IID, and doses remain separate categories.

```{r compatible-iid-model}
iid_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    dose = categorical()
)
```

### Metadata-informed means and dose pooling

#### Metadata-informed additive offsets

Prepared cell covariates can define the mean of $W_0$, and compound
covariates the mean of $V_0$:

$$
W_{0k} = x_k^\top\beta_W + u_{Wk}, \qquad
V_{0i} = z_i^\top\beta_V + u_{Vi}
$$

These feature-informed means let annotations share information across entities.
They are especially useful for a cell or compound that is included and annotated
when the model is fitted but has few or no observed responses: its features can
still inform its baseline or additive effect.

Within `combo_gaussian_component()`, the two fields relevant to feature
regression have separate jobs:

- `mean` is a one-sided R formula that constructs $X$ and specifies
  $X\beta$. It must omit an intercept, for example `~ 0 + pathway_activity + growth_rate`.
- `beta_precision` specifies the prior regularization of $\beta$. It is
  required when `mean` contains predictors; here `gamma_precision()` lets the
  data learn their common prior precision.

The component-level `shrinkage` argument has a different role: it controls the
residual $u$ around $X\beta$, not the feature coefficients. This separation
allows strong feature signal and small unexplained entity variation, or the
reverse, to be learned independently.

The covariates referenced by these formulas are supplied separately from the
experiment table when the model is fitted:

- a `cell_offset` formula is evaluated using `cell_data`, matched to model
  cells by its unique, nonmissing `cell` column;
- a `treatment_offset` formula is evaluated using `compound_data`, matched by
  its unique, nonmissing `drug` column. Compound annotations are then attached
  to every modeled dose of that compound.

Both metadata tables are passed to `fit_combo()`. Variables used by a formula
must be present and finite for every relevant entity. Each formula-generated
model-matrix column is centered and divided by its sample standard deviation
before multiplying by $\beta$. Consequently, `beta_precision`, posterior
coefficients, and `prior_predict(beta_offset = ...)` all operate on this
standardized design scale, not on the raw covariate scale. For an ordinary
numeric covariate, a coefficient is the change associated with a one-standard-
deviation increase in that covariate, holding the other design columns fixed.

```{r feature-model}
feature_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        mean = ~ 0 + pathway_activity + growth_rate,
        beta_precision = gamma_precision(),
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        mean = ~ 0 + target_signature + physchem_pc,
        beta_precision = gamma_precision(),
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    dose = categorical()
)
```

Feature-informed means are currently limited to the scalar, identifiable offsets $W_0$ and $V_0$.
A coefficient in the $W_0$ mean describes how a cell feature shifts baseline
response, while a coefficient in the $V_0$ mean describes how a compound feature
shifts its additive effect across cells. The factor variables $W$, $V_1$, and $V_2$,
by contrast, enter the predictor only through products and have sign and
permutation symmetries. A coefficient for, say, the first coordinate would
change meaning under an equivalent representation that flips or reorders the factors.
Without additional identification constraints, feature means for these factors
are not well defined, so the current API keeps them at zero. Biological
information can still enter factor components through `structure`, which relates
entities without assigning a biological label to a particular latent coordinate.

#### Dose pooling within compounds

With `nested(precision = 4)`, a treatment value at dose $r$ is modeled around
a shared compound parent:

$$
V_{ir} = \bar V_i + e_{ir}, \qquad e_{ir} \mid \rho \sim \mathcal N(0, (4\rho)^{-1} I)
$$

The same construction applies to $V_0$, $V_1$, and $V_2$. It reduces uncertainty
for a sparse or missing dose by borrowing from other doses of that compound. The
value `4` is relative to the component precision $\rho$ and controls the strength
of dose-to-parent pooling. It does not use numerical dose distance, impose
monotonicity, or fit a dose-response curve.

```{r nested-dose-model}
nested_dose_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        mean = ~ 0 + pathway_activity + growth_rate,
        beta_precision = gamma_precision(),
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        mean = ~ 0 + target_signature + physchem_pc,
        beta_precision = gamma_precision(),
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    dose = nested(precision = 4)
)
```

### Tree and GMRF dependence structures

The final two models start from `nested_dose` and allow off-diagonal entries in $Q$.
In both cases, related entities can have correlated latent factors. A tree and a GMRF
are two different ways to specify those correlations.

#### Hierarchical structure with `tree()`

Biological classifications are often hierarchical: cells belong to lineages,
compounds share mechanisms, and both may sit within broader groups. A tree prior
encodes the expectation that entities with more shared ancestry should behave
more similarly. Here, lineage relationships structure the cell factors $W$,
while mechanism relationships structure the treatment factors $V_1$ and $V_2$.

For factor dimension $d$, the structured residual prior is

$$
u_{\cdot d} \mid \rho_d,Q \sim \mathcal N\left(0,(\rho_d Q)^{-1}\right)
$$

The tree builds $Q$ from parent-child edges and edge lengths. The multiplier
$\rho_d$ controls the overall precision without changing the relative tree
structure.

Within `combo_gaussian_component()`,

- `structure` supplies $Q$;
- `shrinkage` supplies $\rho_d$.

These component specifications are assigned in `combo_model()` through

- `cell_factors`: $W$
- `treatment_main_factors`: $V_1$
- `treatment_interaction_factors`: $V_2$.

We keep their means at zero. The feature-informed offsets $W_0$ and $V_0$ and
the nested-dose model remain as in the preceding section, so this model changes
only the dependence among latent factors.

`tree()` constructs the relative precision from a data frame containing `node`
and `parent` columns and, optionally, a positive `edge_length`. Each child is a
Gaussian increment from its parent. Edge precision is the inverse of edge
length, so shorter edges express stronger similarity, and leaves covary through
their shared ancestral path. Internal parents become latent Gaussian nodes.
The tree description must be valid meaning:

- every modeled cell or compound must be a terminal node
- all nonmissing parents exist
- no cycles


Here, cells share lineage parents and compounds share mechanism parents:

```{r tree-structures}
cell_tree_data <- data.frame(
    node = c(".cell_root", "lineage_A", "lineage_B", cells),
    parent = c(NA, ".cell_root", ".cell_root", rep("lineage_A", 4), rep("lineage_B", 4)),
    edge_length = c(1, 1, 1, rep(0.05, 8)),
    stringsAsFactors = FALSE
)
compound_tree_data <- data.frame(
    node = c(".drug_root", sprintf("MOA_%d", 1:4), drugs),
    parent = c(NA, rep(".drug_root", 4), rep(sprintf("MOA_%d", 1:4), each = 2)),
    edge_length = c(1, rep(1, 4), rep(0.05, 8)),
    stringsAsFactors = FALSE
)
cell_tree <- tree(node, parent, cell_tree_data, edge_length)
compound_tree <- tree(node, parent, compound_tree_data, edge_length)
```

The tree model assigns these structures to the three factor components.
It retains `multiplicative_gamma()` for `cell_factors` and uses
`global_half_cauchy()` for the two treatment-factor components.

```{r tree-model}
tree_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        mean = ~ 0 + pathway_activity + growth_rate,
        beta_precision = gamma_precision(),
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = cell_tree,
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        mean = ~ 0 + target_signature + physchem_pc,
        beta_precision = gamma_precision(),
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = compound_tree,
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = compound_tree,
        shrinkage = global_half_cauchy()
    ),
    dose = nested(precision = 4)
)
```

#### Graph structure with `gmrf()`

Not all biological knowledge forms a hierarchy. Pathway interactions, molecular
networks, and selected similarities are more naturally represented as a graph.
A Gaussian Markov random field (GMRF) instead connects entities directly.
Neighboring entities are encouraged to have similar factor values, conditional
on the rest of the graph. The figure shows the two different constructions used
in this example.

```{r tree-gmrf-connectivity, dev="svg", fig.width=7, fig.height=3.2, fig.cap="Connectivity for the cell structures. Tree leaves connect through latent lineage parents and a shared root; the example GMRF contains two disconnected within-lineage cliques."}
leaf_xy <- rbind(
    c(0.7, 1.8), c(1.7, 1.8), c(0.7, 0.6), c(1.7, 0.6),
    c(3.7, 1.8), c(4.7, 1.8), c(3.7, 0.6), c(4.7, 0.6)
)
parent_xy <- rbind(c(1.2, 1.2), c(4.2, 1.2))
root_xy <- c(2.7, 1.2)

structure_panel <- function(title) {
    plot.new()
    plot.window(xlim = c(0.2, 5.2), ylim = c(0.2, 2.2), asp = 1)
    title(title, line = 0.2)
}
draw_structure_nodes <- function(xy, labels, fill = "white", size = 2.7) {
    points(xy[, 1], xy[, 2], pch = 21, bg = fill, col = "grey25", cex = size)
    text(xy[, 1], xy[, 2], labels, cex = 0.62)
}

old_par <- par(mfrow = c(1, 2), mar = c(0.2, 0.2, 1.8, 0.2), xpd = NA)

structure_panel("Tree")
segments(
    rep(parent_xy[, 1], each = 4), rep(parent_xy[, 2], each = 4),
    leaf_xy[, 1], leaf_xy[, 2], col = "grey45"
)
segments(
    root_xy[1], root_xy[2], parent_xy[, 1], parent_xy[, 2],
    col = "grey45", lwd = 1.5
)
draw_structure_nodes(leaf_xy, cells)
draw_structure_nodes(parent_xy, c("Lineage A", "Lineage B"), "#D9EAF7", 3.2)
draw_structure_nodes(matrix(root_xy, nrow = 1), "Root", "#FADDB7", 3.0)

structure_panel("GMRF")
for (group in split(seq_len(8), rep(1:2, each = 4))) {
    edges <- combn(group, 2)
    segments(
        leaf_xy[edges[1, ], 1], leaf_xy[edges[1, ], 2],
        leaf_xy[edges[2, ], 1], leaf_xy[edges[2, ], 2],
        col = "grey45"
    )
}
draw_structure_nodes(leaf_xy, cells, "#E2F0D9")

par(old_par)
```

`batchieR` does not construct a graph from annotations. The user supplies a
sparse operator to `gmrf()`, usually built from a meaningful edge list or sparse
similarity graph. For this example, we use the combinatorial graph Laplacian

$$
L = D - A
$$

where $A$ is a non-negative symmetric adjacency matrix and $D$ is a diagonal
matrix containing $A$'s row sums. This gives the penalty

$$
x^\top Lx = \frac{1}{2}\sum_{i,j}A_{ij}(x_i-x_j)^2
$$

Larger edge weights more strongly encourage neighbors to agree. The helper
below connects cells within lineages and compounds within mechanism groups.

```{r gmrf-structures}
graph_laplacian <- function(nodes, edges) {
    adjacency <- sparseMatrix(
        i = match(edges[, 1], nodes),
        j = match(edges[, 2], nodes),
        x = 1,
        dims = rep(length(nodes), 2),
        dimnames = list(nodes, nodes)
    )
    adjacency <- adjacency + t(adjacency)
    operator <- Diagonal(
        length(nodes),
        x = rowSums(adjacency)
    ) - adjacency
    dimnames(operator) <- list(nodes, nodes)
    drop0(operator)
}

cell_edges <- do.call(rbind, lapply(split(cells, cell_data$lineage), function(group) {
    t(combn(group, 2))
}))
compound_edges <- do.call(rbind, lapply(
    split(drugs, compound_data$mechanism),
    function(group) t(combn(group, 2))
))
cell_gmrf <- gmrf(8 * graph_laplacian(cells, cell_edges), ridge = 0.5)
compound_gmrf <- gmrf(8 * graph_laplacian(drugs, compound_edges), ridge = 0.5)
```

The GMRF model replaces the tree structures with these graph precisions and
otherwise keeps the tree-model specification unchanged.

```{r gmrf-model}
gmrf_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        mean = ~ 0 + pathway_activity + growth_rate,
        beta_precision = gamma_precision(),
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = cell_gmrf,
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        mean = ~ 0 + target_signature + physchem_pc,
        beta_precision = gamma_precision(),
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = compound_gmrf,
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = compound_gmrf,
        shrinkage = global_half_cauchy()
    ),
    dose = nested(precision = 4)
)
```

#### Prior-structure selection and validation

The graph and its weights are part of the prior. A combinatorial Laplacian,
normalized Laplacian, or another precision can give different results, so
`batchieR` does not choose one automatically. The Laplacian above is only a
concrete example.

Given an operator, `gmrf(operator, ridge)` constructs

$$
Q = \text{operator} + \text{ridge} I
$$

The operator must be an exactly symmetric sparse `Matrix` with finite entries
and matching row and column names. The names must match the modeled cells or
compounds, and the regularized $Q$ must be positive definite. Sparse
neighborhoods are also cheaper to fit than dense relationships.

The finite positive ridge makes a positive-semidefinite operator into a proper
positive-definite precision. If a complete sparse positive-definite precision
is already available, `precision(Q)` accepts it without adding a ridge. In the
example, `gmrf(8 * L, ridge = 0.5)` gives $Q=8+0.5I$: the multiplier scales
the edge penalty relative to the ridge, while component `shrinkage` scales
the resulting $Q$ as a whole.

For both tree and GMRF structures, `shrinkage` must scale $Q$ as a whole.
Entity-local `local_half_cauchy()` and `horseshoe()` scales are not available:
using them would require choosing a distinct node- or edge-scaled prior.

Fixed choices such as edge lengths, graph weights, dose precision, and the GMRF ridge
should be checked with prior-predictive simulation and sensitivity analysis.
They should come from external knowledge or plausible response behavior, rather
than being tuned to the same hidden outcomes used for evaluation. For a fuller treatment, see
[Betancourt's principled Bayesian workflow](https://betanalpha.github.io/assets/case_studies/principled_bayesian_workflow.html)
and [Gabry et al. on visualization in Bayesian workflow](https://academic.oup.com/jrsssa/article/182/2/389/7070184).

## Simulated annotated combination screen

### Data-generating model

The six models above are the analysis candidates. To generate synthetic data, we use

- 3 latent dimensions
- 1 relevant and 1 nuisance feature for both cells and compounds,
- tree structure precision for both cells and compounds, and
- nested doses.

Its hyperparameters are chosen to make the behavior of the candidate models easy to see.

```{r synthetic-modelg}
synthetic_model <- combo_model(
    rank = 3,
    global_intercept = fixed_intercept(1.15),
    cell_offset = combo_gaussian_component(
        mean = ~ 0 + pathway_activity + growth_rate,
        beta_precision = fixed_scale(precision = 1000),
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = cell_tree,
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        mean = ~ 0 + target_signature + physchem_pc,
        beta_precision = fixed_scale(precision = 1000),
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = compound_tree,
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = compound_tree,
        shrinkage = global_half_cauchy()
    ),
    dose = nested(precision = 4),
    family = gaussian_response(
        precision = gamma_precision(shape = 100, rate = 1.44)
    )
)
```

The synthetic screen contains 128 single-agent experiments and 192 combinations.
These comprise four within-mechanism pairs and eight fixed cross-mechanism pairs,
each at matched low-low and high-high doses in every cell.

```{r screen-layoutg}
within_pairs <- rbind(
    c("D01", "D02"), c("D03", "D04"),
    c("D05", "D06"), c("D07", "D08")
)
cross_pairs <- rbind(
    c("D02", "D03"), c("D04", "D05"),
    c("D06", "D07"), c("D01", "D08"),
    c("D01", "D03"), c("D03", "D05"),
    c("D05", "D07"), c("D01", "D07")
)
modeled_pairs <- rbind(within_pairs, cross_pairs)

single_rows <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    dose_1 = 1:2,
    stringsAsFactors = FALSE
)
single_rows$drug_2 <- NA_character_
single_rows$dose_2 <- NA_real_

pair_index <- expand.grid(
    cell = cells,
    pair = seq_len(nrow(modeled_pairs)),
    dose = 1:2,
    stringsAsFactors = FALSE
)
pair_rows <- data.frame(
    cell = pair_index$cell,
    drug_1 = modeled_pairs[pair_index$pair, 1],
    dose_1 = pair_index$dose,
    drug_2 = modeled_pairs[pair_index$pair, 2],
    dose_2 = pair_index$dose,
    stringsAsFactors = FALSE
)

screen <- rbind(
    single_rows[c("cell", "drug_1", "dose_1", "drug_2", "dose_2")],
    pair_rows[c("cell", "drug_1", "dose_1", "drug_2", "dose_2")]
)
stopifnot(nrow(single_rows) == 128, nrow(pair_rows) == 192, nrow(screen) == 320)
```

We draw one latent conditional mean and one noisy synthetic response for each
experiment from `synthetic_model` using `prior_predict()`. The feature-coefficient
priors are centered on the intended cell and compound signals through `beta_offset`;
all other coefficients retain their zero-centered priors. Here, `0.42` and `-0.28`
are standardized-scale coefficients rather than raw-covariate slopes. Reusing
the seed for these single-draw calls keeps the latent state identical; only the
response draw includes Gaussian observation noise.

```{r generate-responsesg}
screen$true_mean <- as.numeric(prior_predict(
    synthetic_model,
    data = screen,
    cell_data = cell_data,
    compound_data = compound_data,
    draws = 1,
    seed = 20260828,
    type = "mean",
    beta_offset = list(
        cell_offset = c(pathway_activity = 0.42),
        treatment_offset = c(target_signature = -0.28)
    )
))

screen$modeled_response <- as.numeric(prior_predict(
    synthetic_model,
    data = screen,
    cell_data = cell_data,
    compound_data = compound_data,
    draws = 1,
    seed = 20260828,
    beta_offset = list(
        cell_offset = c(pathway_activity = 0.42),
        treatment_offset = c(target_signature = -0.28)
    )
))

# convert logit-scale responses to viability
screen$viability <- plogis(screen$modeled_response)
```

### Evaluation plates

We now define six plates and hide their responses.
Later we use these same plates in two ways:
first to measure prediction error, and then as candidates for experimental design.

| Plate | Four unique configurations | Main question |
|:--|:--|:--|
| `interpolation` | familiar cells, compounds, and doses | Can the model fill an ordinary gap? |
| `dose_transfer` | a held-out dose of `D05+D06` | Does pooling across doses help? |
| `unobserved_cell` | two combinations in `C04` and `C08` | Can annotations and lineage transfer to cells with no responses? |
| `unobserved_compound` | `D07+D08` in four observed cells | Can annotations and mechanism transfer to a compound with no responses? |
| `related_combination` | `D01+D03` in four cells | Can information transfer from related compounds and combinations? |
| `cross_mechanism_discovery` | `D03+D05` in four cells | How does the model treat a new cross-mechanism combination? |

Here, *unobserved* means that an entity is declared and annotated when the model
is fitted but has no observed responses. Each plate contains four unique
configurations and two wells per configuration. Prediction error is calculated
once per unique configuration. PDBAL uses all eight wells because replication
changes the information supplied by the plate.

```{r response-mask}
# Helpers for identifying compounds and unordered compound pairs.
is_combination <- !is.na(screen$drug_2)
contains_drug <- function(data, selected) {
    data$drug_1 %in% selected |
        (!is.na(data$drug_2) & data$drug_2 %in% selected)
}
pair_key <- function(drug_1, drug_2) {
    ifelse(
        is.na(drug_2),
        drug_1,
        paste(pmin(drug_1, drug_2), pmax(drug_1, drug_2), sep = ":")
    )
}

# Match a proposed experiment back to its row on the response surface.
row_matches <- function(cell, drug_1, dose_1, drug_2 = NA_character_, dose_2 = NA_real_) {
    screen$cell == cell & screen$drug_1 == drug_1 & screen$dose_1 == dose_1 &
        if (is.na(drug_2)) {
            is.na(screen$drug_2)
        } else {
            screen$drug_2 == drug_2 & screen$dose_2 == dose_2
        }
}
candidate_indices <- function(configurations) {
    indices <- integer(nrow(configurations))
    for (index in seq_len(nrow(configurations))) {
        matched <- which(row_matches(
            configurations$cell[index],
            configurations$drug_1[index],
            configurations$dose_1[index],
            configurations$drug_2[index],
            configurations$dose_2[index]
        ))
        stopifnot(length(matched) == 1)
        indices[index] <- matched
    }
    indices
}

# Build matched-dose combination configurations for candidate plates.
candidate_config <- function(cells, drug_1, drug_2, dose) {
    data.frame(
        cell = cells,
        drug_1 = drug_1,
        dose_1 = dose,
        drug_2 = drug_2,
        dose_2 = dose,
        stringsAsFactors = FALSE
    )
}

# Define each plate once, before adding replicate wells.
prediction_data <- list(
    interpolation = candidate_config(
        c("C01", "C02", "C05", "C06"),
        c("D01", "D03", "D05", "D06"),
        c("D02", "D04", "D06", "D07"),
        1
    ),
    dose_transfer = candidate_config(
        c("C01", "C02", "C05", "C06"), "D05", "D06", 2
    ),
    unobserved_cell = candidate_config(
        c("C04", "C04", "C08", "C08"),
        c("D01", "D03", "D01", "D03"),
        c("D02", "D04", "D02", "D04"),
        c(1, 2, 1, 2)
    ),
    unobserved_compound = candidate_config(
        c("C01", "C02", "C05", "C06"), "D07", "D08", c(1, 2, 1, 2)
    ),
    related_combination = candidate_config(
        c("C01", "C02", "C05", "C06"), "D01", "D03", c(1, 2, 1, 2)
    ),
    cross_mechanism_discovery = candidate_config(
        c("C01", "C02", "C05", "C06"), "D03", "D05", 1
    )
)
prediction_indices <- lapply(prediction_data, candidate_indices)

# The physical plates contain two wells for each unique configuration.
candidate_data <- lapply(prediction_data, function(data) {
    data[rep(seq_len(nrow(data)), each = 2), , drop = FALSE]
})

# Hide extra rows needed to make the cell, compound, dose, and combinations
# genuinely unobserved. The named plate rows are the rows evaluated below.
held_out_cell <- screen$cell %in% c("C04", "C08")
held_out_compound <- contains_drug(screen, "D08")
held_out_dose <- screen$dose_1 == 2 & contains_drug(screen, c("D05", "D06"))
held_out_pair <- is_combination & pair_key(screen$drug_1, screen$drug_2) %in%
    c("D01:D03", "D03:D05")

screen$role <- "training"
screen$role[held_out_cell | held_out_compound | held_out_dose | held_out_pair] <-
    "supporting_holdout"
for (plate in names(prediction_indices)) {
    screen$role[prediction_indices[[plate]]] <- plate
}
screen$response <- ifelse(
    screen$role == "training",
    screen$modeled_response,
    NA_real_
)

# Retain only the columns required by prediction and plate scoring.
prediction_columns <- c("cell", "drug_1", "dose_1", "drug_2", "dose_2")
prediction_data <- lapply(prediction_data, function(data) data[prediction_columns])
candidate_data <- lapply(candidate_data, function(data) data[prediction_columns])
stopifnot(
    all(vapply(prediction_data, nrow, integer(1)) == 4),
    all(vapply(candidate_data, nrow, integer(1)) == 8),
    all(vapply(prediction_indices, length, integer(1)) == 4)
)

# Inspect two experiments from each role before summarizing the masking design.
screen_preview <- do.call(rbind, lapply(
    split(screen, screen$role),
    head,
    2
))
rownames(screen_preview) <- NULL
screen_preview[c(prediction_columns, "viability", "role")]
table(screen$role)
```

Every row remains in the fitting data. A missing response keeps a row out of the
likelihood while retaining its entity mappings. This lets the fitted model know
about the annotated cells, compounds, and doses that appear on the plates.

## Effects of model specification

We fit all six models to the same observed rows. Sampler mechanics are covered
in the introduction. A single chain keeps this worked example simple, while a
longer warmup and sampling run makes its posterior summaries more stable. A
biological analysis should use multiple chains and explicit convergence
diagnostics.

```{r fit-models}

models <- list(
    default = default_model,
    iid = iid_model,
    features = feature_model,
    nested_dose = nested_dose_model,
    tree = tree_model,
    gmrf = gmrf_model
)
fit_data <- screen[c(prediction_columns, "response")]

set.seed(20260900)
fits <- lapply(
    models,
    fit_combo,
    data = fit_data,
    cell_data = cell_data,
    compound_data = compound_data,
    chains = 1,
    parallel_chains = 1,
    iter_warmup = 500,
    iter_sampling = 1000,
    thin = 1,
    refresh = if (interactive()) 25 else 0
)
names(fits) <- names(models)
```

We use the fitted models to inspect shrinkage and feature coefficients, predict
the hidden plates, and score those same plates for a future experiment.

### Latent-dimension shrinkage

The default cell-factor prior uses multiplicative-gamma shrinkage. The plot
shows the posterior mean global precision for each latent dimension, with 90%
credible intervals. Higher precision means stronger shrinkage toward zero, so
higher values correspond to dimensions that the model treats as less
informative.

```{r shrinkage-summary}
summarize_mapped_draws <- function(fit, selected_map) {
    selected_draws <- posterior_draws(fit, variable = selected_map$variable)
    draw_matrix <- matrix(
        selected_draws,
        ncol = length(selected_map$variable),
        dimnames = list(NULL, selected_map$variable)
    )
    data.frame(
        selected_map,
        posterior_mean = colMeans(draw_matrix),
        q05 = apply(draw_matrix, 2, quantile, 0.05),
        q95 = apply(draw_matrix, 2, quantile, 0.95),
        row.names = NULL
    )
}

default_map <- parameter_map(fits$default)
shrinkage_map <- subset(
    default_map,
    component == "cell_factors" & parameter == "global_precision"
)
shrinkage_summary <- summarize_mapped_draws(fits$default, shrinkage_map)
ggplot(shrinkage_summary, aes(x = factor(dimension), y = posterior_mean)) +
    geom_col(fill = "#2166ac") +
    geom_errorbar(aes(ymin = q05, ymax = q95), width = 0.15) +
    labs(x = "Latent dimension", y = "Global precision") +
    theme_minimal()
```

### Metadata-effect estimates

Metadata formulas affect additive offsets only. Their coefficients are on the
internally standardized feature scale. Because the synthetic responses are a
prior-predictive draw rather than a stored parameter draw, this summary is an
illustration of posterior estimation, not a parameter-recovery check.

```{r feature-coefficients}
feature_map <- subset(
    parameter_map(fits$features),
    parameter == "mean_coefficient"
)
feature_draws <- posterior_draws(
    fits$features,
    variable = feature_map$variable
)
feature_draw_matrix <- matrix(
    feature_draws,
    ncol = nrow(feature_map),
    dimnames = list(NULL, feature_map$feature)
)
bayesplot::mcmc_areas(
    feature_draw_matrix,
    prob = 0.8,
    prob_outer = 0.95,
    point_est = "median"
) +
    geom_vline(xintercept = 0, colour = "grey60", linetype = "dashed") +
    labs(x = "Feature coefficient", y = NULL)
```

The intervals remain broad and include zero for some coefficients. This example
shows how to add a feature-informed mean, not how to recover or interpret each
coefficient. Features can still help prediction even when their individual
coefficients are uncertain.

### Held-out plate prediction

For each plate, RMSE compares the posterior expected response with the hidden
latent generating mean. We calculate it over the four unique configurations,
not over the replicated wells.

```{r prediction-metrics}
metric_rows <- lapply(seq_along(fits), function(fit_index) {
    fit <- fits[[fit_index]]
    do.call(rbind, lapply(names(prediction_data), function(plate) {
        expected <- posterior_epred(fit, newdata = prediction_data[[plate]])
        stopifnot(all(is.finite(expected)))
        data.frame(
            model = names(fits)[fit_index],
            plate = plate,
            n_configurations = nrow(prediction_data[[plate]]),
            latent_rmse = sqrt(mean(
                (colMeans(expected) - screen$true_mean[prediction_indices[[plate]]])^2
            )),
            stringsAsFactors = FALSE
        )
    }))
})
prediction_metrics <- do.call(rbind, metric_rows)
rownames(prediction_metrics) <- NULL
prediction_metrics$model <- factor(
    prediction_metrics$model,
    levels = rev(names(models))
)
prediction_metrics$plate <- factor(
    prediction_metrics$plate,
    levels = names(prediction_data)
)
prediction_metrics
```

```{r prediction-metrics-plot, fig.alt="Held-out latent-mean RMSE by model and plate"}
ggplot(prediction_metrics, aes(x = latent_rmse, y = model, fill = model)) +
    geom_col(show.legend = FALSE) +
    scale_fill_brewer(palette = "Set2") +
    facet_wrap(vars(plate)) +
    labs(x = "RMSE against true latent mean", y = NULL) +
    theme_bw()
```

Read each comparison locally. The feature model can help the two unobserved-
entity plates, nested doses target `dose_transfer`, and the tree and GMRF can
help when lineage or mechanism relationships are useful. These are results from
one synthetic draw, so the plot illustrates model behavior rather than ranking
the priors in general.

### Prospective plate selection

We now score the same six plates with PDBAL. Unlike RMSE, the score does not use
their hidden outcomes. It asks which complete eight-well plate is expected to
reduce posterior uncertainty across the 320-row reference surface. Distances
are measured after converting expected logit responses back to viability; the
candidate likelihood remains Gaussian on the fitted logit scale. Lower scores
are preferred.

```{r score-candidates}
reference_grid <- screen[prediction_columns]
score_rows <- lapply(seq_along(fits), function(fit_index) {
    result <- score_plate_pdbal(
        fits[[fit_index]],
        candidate_data = candidate_data,
        reference_grid = reference_grid,
        response_transform = plogis,
        max_triplets = 1000
    )
    result$model <- names(fits)[fit_index]
    result
})
plate_scores <- do.call(rbind, score_rows)
rownames(plate_scores) <- NULL

plate_scores$normalized_score <- ave(
    plate_scores$score,
    plate_scores$model,
    FUN = function(scores) scores - min(scores)
)
plate_scores$normalized_score <- round(plate_scores$normalized_score, 3)
plate_ranks <- reshape(
    plate_scores[c("model", "plate", "normalized_score")],
    idvar = "plate",
    timevar = "model",
    direction = "wide"
)
plate_ranks <- plate_ranks[match(names(candidate_data), plate_ranks$plate), , drop = FALSE]
names(plate_ranks) <- sub("normalized_score[.]", "", names(plate_ranks))
plate_ranks
```

The normalized score is zero for the preferred plate within each model. Only
values within a model should be compared because changing the model also
changes its posterior distances and predictive distributions.

```{r selected-plates}
selected_plates <- plate_scores[plate_scores$rank == 1, c("model", "plate")]
selected_plates
```

RMSE and PDBAL answer different questions. RMSE measures how accurately the
model predicts a hidden plate. PDBAL measures how useful the model expects that
plate to be for reducing uncertainty over the wider response surface. The prior
can change both quantities and can therefore change which plate is selected.

## Session information

```{r session-information}
sessionInfo()
```
