---
title: "Incorporating biological prior information"
author: "Teo Sakel"
date: "2026-08-31"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Incorporating biological prior information}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4.5
)
library(batchieR)
library(Matrix)
library(ggplot2)
set.seed(20260827)
```

## From biological knowledge to a prior

`batchieR` models a combination response through additive and latent-factor
terms. This vignette explains how biological knowledge can enter the prior for
those terms. It assumes familiarity with the basic fitting and prediction
workflow in `vignette("introduction", package = "batchieR")`; details of fit
objects, printing, and sampler use are intentionally omitted here.

For cell $k$ and compound-dose treatments $i$ and $j$, the linear predictor is

$$
\eta_{kij} = \alpha + W_{0k} + V_{0i} + V_{0j}
    + W_k^\top\left(V_{1i} + V_{1j} + V_{2i}\odot V_{2j}\right).
$$

The five configurable variables have distinct scientific meanings:

| Variable | Role | Information that can be useful |
|:--|:--|:--|
| $W_0$ | cell baseline                   | growth, lineage, or other cell-level annotations |
| $W$   | latent cell state               | relationships among cells |
| $V_0$ | additive treatment effect       | targets, chemical features, compound and dose relationships |
| $V_1$ | cell-dependent treatment effect | compound and dose relationships |
| $V_2$ | nonadditive combination effect  | compound and dose relationships |

The current implementation limits all five configurable terms to conditionally
Gaussian components. Each component can be written schematically as

$$
\theta = X\beta + u, \qquad
u \mid \rho,Q \sim \mathcal N(0, (\rho Q)^{-1}).
$$

This separates three questions:

1. **Mean:** can measured features $X$ explain systematic differences?
2. **Magnitude:** how strongly should unexplained deviations $u$ shrink?
3. **Dependence:** should $Q$ make deviations from related entities similar?

`combo_gaussian_component()` turns these choices into a component
specification for assignment to one of the five slots in `combo_model()`. Its
arguments map directly onto the decomposition above:

| Argument | Part of the prior |
|:--|:--|
| `mean` | the systematic term $X\beta$; `~ 0` specifies a zero mean |
| `mean_shrinkage` | regularization of the coefficients $\beta$ when the mean uses features |
| `shrinkage` | the overall or entity-specific magnitude of the residual $u$ |
| `structure` | the relative residual precision $Q$; `iid()` uses $Q=I$ |

For example, a feature-informed offset component with otherwise independent residuals
can be specified as

```{r gaussian-component-template}
example_component <- combo_gaussian_component(
    mean = ~ 0 + feature_1 + feature_2,
    mean_shrinkage = gamma_precision(),
    structure = iid(),
    shrinkage = global_half_cauchy()
)
```

Like most latent-factorization methods, this model does not give individual
latent coordinates an intrinsic interpretation. The components $W$, $V_1$, and
$V_2$ enter the predictor only through products, so coordinated changes to their
representation can leave the likelihood unchanged. Consequently, coordinate-specific
nonzero means for these factors would depend on an arbitrary representation and
require additional identification constraints to be interpretable. It therefore
makes sense to define feature-informed means only for the scalar, identifiable
offset variables $W_0$ and $V_0$, while using covariance or precision structures
to encode biological relationships among entities in the factor components.

A fourth question applies to treatments: should doses be unrelated categories,
or deviations from a shared compound-level value? The sections below treat
these choices separately because they encode different information and solve
different prediction problems.

## Synthetic example

We use a small annotated screen with

- 8 *cell* lines in two *lineages*,
- 8 *compounds* in 4 *mechanism groups*, and
- 2 *doses*.

```{r entity-metadata}
cells <- sprintf("C%02d", 1:8)
drugs <- sprintf("D%02d", 1:8)

cell_data <- data.frame(
    cell = cells,
    lineage = rep(c("lineage_A", "lineage_B"), each = 4),
    pathway_activity = c(-1.4, -1.0, -0.6, -0.2, 0.2, 0.6, 1.0, 1.4),
    growth_rate = c(-0.7, 0.8, -0.2, 0.4, 0.7, -0.8, 0.2, -0.4),
    stringsAsFactors = FALSE
)
compound_data <- data.frame(
    drug = drugs,
    mechanism = rep(sprintf("MOA_%d", 1:4), each = 2),
    target_signature = c(-1.5, -1.2, -0.6, -0.3, 0.3, 0.6, 1.2, 1.5),
    physchem_pc = c(0.6, -0.7, -0.4, 0.8, -0.8, 0.4, 0.7, -0.6),
    stringsAsFactors = FALSE
)
cell_data
compound_data
```

```{r example-hierarchies, dev="svg", fig.width=7, fig.height=3.6, echo=FALSE, fig.cap="Hierarchies in the running example. Each cell belongs to a lineage; each compound belongs to a mechanism group and has two doses."}
draw_nodes <- function(x, y, labels, fill, cex = 0.75) {
    y <- rep_len(y, length(x))
    points(x, y, pch = 21, bg = fill, col = "grey30", cex = 2.5)
    text(x, y, labels, cex = cex)
}

old_par <- par(mfrow = c(1, 2), mar = c(0.5, 0.5, 2, 0.5), xpd = NA)

plot.new()
plot.window(xlim = c(0.5, 8.5), ylim = c(0.5, 2.5))
cell_x <- seq_len(8)
lineage_x <- c(2.5, 6.5)
segments(rep(lineage_x, each = 4), 2, cell_x, 1)
draw_nodes(lineage_x, 2, c("Lineage A", "Lineage B"), "#D9EAF7", 0.7)
draw_nodes(cell_x, 1, cells, "white", 0.7)
title("Cells", line = 0.5)

plot.new()
plot.window(xlim = c(0.25, 8.75), ylim = c(0.25, 3.5))
drug_x <- seq_len(8)
mechanism_x <- seq(1.5, 7.5, by = 2)
dose_x <- as.vector(rbind(drug_x - 0.18, drug_x + 0.18))
segments(rep(mechanism_x, each = 2), 3, drug_x, 2)
segments(rep(drug_x, each = 2), 2, dose_x, 1)
draw_nodes(mechanism_x, 3, sprintf("MOA %d", 1:4), "#FADDB7", 0.65)
draw_nodes(drug_x, 2, drugs, "white", 0.65)
draw_nodes(dose_x, 1, rep(c("1", "2"), 8), "#E2F0D9", 0.55)
text(8.65, 1, "Dose", adj = 1, cex = 0.65)
title("Treatments", line = 0.5)

par(old_par)
```

The annotations stand in for prepared pathway scores, expression components,
target annotations, or other low-dimensional summaries. `pathway_activity` and
`target_signature` generate most of the additive truth while `growth_rate` and
`physchem_pc` are nuisance covariates included to show that a mean formula is a
regularized model, not a command to use every annotation.

The relevant *response surface* contains all 128 single-agent experiments and
192 combinations: four within-mechanism pairs and eight fixed cross-mechanism
pairs, each at matched low-low and high-high doses in every cell.

```{r screen-layout, include=FALSE}
within_pairs <- rbind(
    c("D01", "D02"), c("D03", "D04"),
    c("D05", "D06"), c("D07", "D08")
)
cross_pairs <- rbind(
    c("D02", "D03"), c("D04", "D05"),
    c("D06", "D07"), c("D01", "D08"),
    c("D01", "D03"), c("D03", "D05"),
    c("D05", "D07"), c("D01", "D07")
)
modeled_pairs <- rbind(within_pairs, cross_pairs)

single_rows <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    dose_1 = 1:2,
    stringsAsFactors = FALSE
)
single_rows$drug_2 <- NA_character_
single_rows$dose_2 <- NA_real_

pair_index <- expand.grid(
    cell = cells,
    pair = seq_len(nrow(modeled_pairs)),
    dose = 1:2,
    stringsAsFactors = FALSE
)
pair_rows <- data.frame(
    cell = pair_index$cell,
    drug_1 = modeled_pairs[pair_index$pair, 1],
    dose_1 = pair_index$dose,
    drug_2 = modeled_pairs[pair_index$pair, 2],
    dose_2 = pair_index$dose,
    stringsAsFactors = FALSE
)

screen <- rbind(
    single_rows[c("cell", "drug_1", "dose_1", "drug_2", "dose_2")],
    pair_rows[c("cell", "drug_1", "dose_1", "drug_2", "dose_2")]
)
stopifnot(nrow(single_rows) == 128, nrow(pair_rows) == 192, nrow(screen) == 320)
```

To generate it, we draw a sample from the prior with known coefficients and
noise, then mask some rows for evaluation. The resulting synthetic screen is a
stand-in for a real experiment, and the known truth lets us evaluate how well
different prior specifications recover it.

```{r generate-responses, include=FALSE}
# Choose the signal, pooling, and noise scales of the synthetic screen.
dgp <- list(
    alpha = 1.15,
    beta_cell = c(pathway_activity = 0.42, growth_rate = 0),
    beta_compound = c(target_signature = -0.28, physchem_pc = 0),
    cell_offset_group_sd = 0.04,
    cell_offset_entity_sd = 0.025,
    compound_offset_group_sd = 0.035,
    compound_offset_entity_sd = 0.025,
    cell_factor_group_sd = c(0.85, 0.45),
    cell_factor_entity_sd = c(0.07, 0.05),
    main_factor_group_sd = c(0.55, 0.40),
    main_factor_entity_sd = c(0.04, 0.035),
    interaction_factor_group_sd = c(0.90, 0.55),
    interaction_factor_entity_sd = c(0.05, 0.04),
    dose_offset_sd = 0.045,
    dose_main_sd = c(0.025, 0.02),
    dose_interaction_sd = c(0.03, 0.025),
    response_sd = 0.12
)

# Draw related entities around a shared group value.
draw_grouped_gaussian <- function(group, group_sd, entity_sd) {
    group <- factor(group, levels = unique(group))
    n_groups <- nlevels(group)
    n_entities <- length(group)
    n_dimensions <- length(group_sd)

    group_value <- matrix(
        rnorm(n_groups * n_dimensions),
        nrow = n_groups,
        ncol = n_dimensions
    )
    group_value <- sweep(group_value, 2, group_sd, `*`)
    group_value <- sweep(group_value, 2, colMeans(group_value), `-`)

    deviation <- matrix(
        rnorm(n_entities * n_dimensions),
        nrow = n_entities,
        ncol = n_dimensions
    )
    deviation <- sweep(deviation, 2, entity_sd, `*`)
    for (level in levels(group)) {
        selected <- group == level
        deviation[selected, ] <- sweep(
            deviation[selected, , drop = FALSE],
            2,
            colMeans(deviation[selected, , drop = FALSE]),
            `-`
        )
    }
    group_value[as.integer(group), , drop = FALSE] + deviation
}

# Draw independent deviations for each latent dimension.
draw_matrix_noise <- function(n_rows, column_sd) {
    noise <- matrix(
        rnorm(n_rows * length(column_sd)),
        nrow = n_rows,
        ncol = length(column_sd)
    )
    sweep(noise, 2, column_sd, `*`)
}

# Combine feature means with lineage- and mechanism-level residuals.
cell_offset_residual <- draw_grouped_gaussian(
    cell_data$lineage,
    dgp$cell_offset_group_sd,
    dgp$cell_offset_entity_sd
)[, 1]
compound_offset_residual <- draw_grouped_gaussian(
    compound_data$mechanism,
    dgp$compound_offset_group_sd,
    dgp$compound_offset_entity_sd
)[, 1]

cell_offset_truth <- as.numeric(
    as.matrix(cell_data[names(dgp$beta_cell)]) %*% dgp$beta_cell
) + cell_offset_residual
compound_offset_truth <- as.numeric(
    as.matrix(compound_data[names(dgp$beta_compound)]) %*% dgp$beta_compound
) + compound_offset_residual
names(cell_offset_truth) <- cells
names(compound_offset_truth) <- drugs

# Draw the rank-two cell factors and compound-level factor parents.
cell_factors_truth <- draw_grouped_gaussian(
    cell_data$lineage,
    dgp$cell_factor_group_sd,
    dgp$cell_factor_entity_sd
)
main_parent <- draw_grouped_gaussian(
    compound_data$mechanism,
    dgp$main_factor_group_sd,
    dgp$main_factor_entity_sd
)
interaction_parent <- draw_grouped_gaussian(
    compound_data$mechanism,
    dgp$interaction_factor_group_sd,
    dgp$interaction_factor_entity_sd
)
rownames(cell_factors_truth) <- cells
rownames(main_parent) <- rownames(interaction_parent) <- drugs

# Draw each dose around its compound-level parent.
treatment_offset_truth <- matrix(
    compound_offset_truth,
    nrow = length(drugs),
    ncol = 2
) + matrix(
    rnorm(length(drugs) * 2, sd = dgp$dose_offset_sd),
    nrow = length(drugs),
    ncol = 2
)
treatment_main_truth <- array(
    NA_real_,
    dim = c(length(drugs), 2, 2),
    dimnames = list(drug = drugs, dose = 1:2, dimension = 1:2)
)
treatment_interaction_truth <- treatment_main_truth
for (dose in 1:2) {
    treatment_main_truth[, dose, ] <- main_parent +
        draw_matrix_noise(length(drugs), dgp$dose_main_sd)
    treatment_interaction_truth[, dose, ] <- interaction_parent +
        draw_matrix_noise(length(drugs), dgp$dose_interaction_sd)
}

data.frame(
    coefficient = c(names(dgp$beta_cell), names(dgp$beta_compound)),
    truth = c(dgp$beta_cell, dgp$beta_compound),
    row.names = NULL
)

# Look up the offset and factors for one compound-dose treatment.
treatment_truth <- function(drug, dose) {
    drug_index <- match(drug, drugs)
    list(
        offset = treatment_offset_truth[drug_index, dose],
        main = treatment_main_truth[drug_index, dose, ],
        interaction = treatment_interaction_truth[drug_index, dose, ]
    )
}

# Evaluate the package's single-agent and combination linear predictor.
linear_predictor <- function(cell, drug_1, dose_1, drug_2, dose_2) {
    first <- treatment_truth(drug_1, dose_1)
    w <- cell_factors_truth[cell, ]
    value <- dgp$alpha + cell_offset_truth[cell] + first$offset + sum(w * first$main)
    if (!is.na(drug_2)) {
        second <- treatment_truth(drug_2, dose_2)
        value <- value + second$offset + sum(w * second$main) +
            sum(w * first$interaction * second$interaction)
    }
    unname(value)
}

# Generate the noise-free surface, then add observation noise.
screen$true_mean <- mapply(
    linear_predictor,
    screen$cell,
    screen$drug_1,
    screen$dose_1,
    screen$drug_2,
    screen$dose_2
)
screen$modeled_response <- screen$true_mean +
    rnorm(nrow(screen), sd = dgp$response_sd)
screen$viability <- plogis(screen$modeled_response)
```

### Evaluation design

We retain observed `training` rows and create four diagnostic holdouts:

1. `warm_holdout` for ordinary interpolation,
2. `cold_start` for annotated cells with no observed responses,
3. `dose_transfer` for a sparse dose, and
4. `related_transfer` for an unobserved related entity or pair.

Rows used only in candidate plates are marked `candidate_only`.
Candidate membership takes precedence over the diagnostic masks.

```{r response-mask, include=FALSE}
# Helpers for identifying compounds and unordered compound pairs.
is_combination <- !is.na(screen$drug_2)
contains_drug <- function(data, selected) {
    data$drug_1 %in% selected |
        (!is.na(data$drug_2) & data$drug_2 %in% selected)
}
pair_key <- function(drug_1, drug_2) {
    ifelse(
        is.na(drug_2),
        drug_1,
        paste(pmin(drug_1, drug_2), pmax(drug_1, drug_2), sep = ":")
    )
}

# Define the prediction regime tested by each held-out row.
screen$role <- "training"
cold_start <- screen$cell %in% c("C04", "C08")
unseen_sister <- !cold_start & contains_drug(screen, "D08")
dose_transfer <- !cold_start & !unseen_sister & screen$dose_1 == 2 &
    contains_drug(screen, c("D05", "D06"))
related_transfer <- unseen_sister | (
    !cold_start & !dose_transfer & is_combination &
        screen$cell %in% c("C03", "C07") &
        pair_key(screen$drug_1, screen$drug_2) %in%
            pair_key(within_pairs[, 1], within_pairs[, 2])
)
warm_holdout <- !cold_start & !dose_transfer & !related_transfer &
    seq_len(nrow(screen)) %% 13 == 0

# Apply the masks in increasing order of precedence.
screen$role[cold_start] <- "cold_start"
screen$role[dose_transfer] <- "dose_transfer"
screen$role[related_transfer] <- "related_transfer"
screen$role[warm_holdout] <- "warm_holdout"

# Match a proposed experiment back to its row on the response surface.
row_matches <- function(cell, drug_1, dose_1, drug_2 = NA_character_, dose_2 = NA_real_) {
    screen$cell == cell & screen$drug_1 == drug_1 & screen$dose_1 == dose_1 &
        if (is.na(drug_2)) {
            is.na(screen$drug_2)
        } else {
            screen$drug_2 == drug_2 & screen$dose_2 == dose_2
        }
}
candidate_indices <- function(configurations) {
    indices <- integer(nrow(configurations))
    for (index in seq_len(nrow(configurations))) {
        matched <- which(row_matches(
            configurations$cell[index],
            configurations$drug_1[index],
            configurations$dose_1[index],
            configurations$drug_2[index],
            configurations$dose_2[index]
        ))
        stopifnot(length(matched) == 1)
        indices[index] <- matched
    }
    indices
}

# Build matched-dose combination configurations for candidate plates.
candidate_config <- function(cells, drug_1, drug_2, dose) {
    data.frame(
        cell = cells,
        drug_1 = drug_1,
        dose_1 = dose,
        drug_2 = drug_2,
        dose_2 = dose,
        stringsAsFactors = FALSE
    )
}

# Compare replication, dose transfer, related-entity transfer, and discovery.
candidate_rows <- list(
    known_replication = rbind(
        candidate_config(c("C01", "C02", "C05", "C06"), "D01", "D02", 1),
        candidate_config(c("C01", "C02", "C05", "C06"), "D03", "D04", 1)
    ),
    dose_extension = candidate_config(
        rep("C01", 8), "D01", "D02", 2
    ),
    related_entity_transfer = candidate_config(
        rep("C04", 8), "D01", "D08", 2
    ),
    cross_mechanism_discovery = candidate_config(
        rep("C04", 8), "D03", "D05", 2
    )
)
candidate_screen_indices <- lapply(candidate_rows, candidate_indices)

# Remove candidate-only rows from the likelihood without dropping their mappings.
cross_discovery_pair <- is_combination &
    pair_key(screen$drug_1, screen$drug_2) == "D03:D05"
screen$role[unique(c(
    candidate_screen_indices$dose_extension,
    which(cross_discovery_pair)
))] <- "candidate_only"
screen$response <- ifelse(
    screen$role == "training",
    screen$modeled_response,
    NA_real_
)

# Retain only the columns required by prediction and plate scoring.
prediction_columns <- c("cell", "drug_1", "dose_1", "drug_2", "dose_2")
candidate_data <- lapply(candidate_rows, function(data) data[prediction_columns])
stopifnot(
    all(vapply(candidate_data, nrow, integer(1)) == 8),
    all(vapply(candidate_screen_indices, length, integer(1)) == 8)
)

# Inspect two experiments from each role before summarizing the masking design.
screen_preview <- do.call(rbind, lapply(
    split(screen, screen$role),
    head,
    2
))
rownames(screen_preview) <- NULL
screen_preview[c(prediction_columns, "viability", "role")]
table(screen$role)
```

The simulation, masking, candidate construction, and plotting helpers are
hidden because they are not part of an applied workflow. Every row remains in
the fitting data: a missing response removes a holdout from the likelihood but
retains its entity mappings. This is essential because prediction cannot
introduce entirely new entities after fitting.

## Four ways to inform the model

`combo_model()` separates the response surface into five configurable components:

1. the cell offset $W_0$,
2. cell factors $W$,
3. treatment offset $V_0$,
4. treatment main factors $V_1$, and
5. treatment interaction factors $V_2$.

Below, we build six models of increasing biological specificity.
Each model uses the same rank-three Gaussian likelihood,
but the priors differ in their means, shrinkage, and structure.

1. the unchanged BATCHIE-style default;
2. an IID control with shrinkage compatible with structured priors;
3. cell and compound features in the additive offsets;
4. partial pooling among doses of the same compound;
5. lineage and mechanism relationships encoded by GMRFs; and
6. the same relationships encoded by trees instead.

The default and compatible IID fits are controls; the other four illustrate
features, dose pooling, and two alternative relationship structures.

```{r model-design, echo=FALSE}
model_design <- data.frame(
    model = c(
        "default", "compatible_iid", "features",
        "nested_dose", "gmrf", "tree"
    ),
    feature_means = c("none", "none", rep("offsets", 4)),
    dose_model = c(rep("categorical", 3), rep("nested", 3)),
    cell_factor_structure = c(rep("IID", 4), "GMRF", "tree"),
    treatment_factor_structure = c(rep("IID", 4), "GMRF", "tree"),
    treatment_shrinkage = c("horseshoe", rep("global half-Cauchy", 5)),
    stringsAsFactors = FALSE
)
model_design
```

### 1. Shrink unused latent dimensions

The latent rank is an upper bound, not a claim that every dimension is needed.
For factor dimension $d$, the default multiplicative-gamma prior uses

$$
\rho_d = \prod_{h=1}^d \delta_h, \qquad \delta_h \sim \operatorname{Gamma}(a,b).
$$

Later dimensions inherit all earlier precision increments and are progressively
shrunk toward zero. This makes a moderately generous rank safer when the true
response surface is lower-dimensional. The remaining default priors treat cells
and compound-dose treatments as unrelated and use entity-local horseshoes for
$V_0$, $V_1$, and $V_2$.

```{r default-model}
default_model <- combo_model(rank = 3)
```

#### A comparable IID baseline

The default horseshoe is defined for IID residuals. For one latent dimension,

$$
u_n \mid \tau, \lambda_n \sim
\mathcal N(0, \tau^2\lambda_n^2),
$$

so the local scale $\lambda_n$ lets entity $n$ escape shrinkage
independently of the others. A structured component instead starts from

$$
u \mid \tau,Q \sim \mathcal N(0, \tau^2 Q^{-1}),
$$

where off-diagonal entries of $Q$ couple related entities. The global scale
$\tau$ generalizes directly because it rescales the whole precision
matrix by $\tau^{-2}$. Local scales do not: multiplying each row of $Q$ by a
different scale generally destroys symmetry, while the symmetric construction
$D_{\lambda}^{-1}QD_{\lambda}^{-1}$ changes both an entity and all its
cross-terms. Putting scales on graph edges or tree increments is also possible,
but defines a different adaptive-smoothing prior. Trees and nested doses add
latent parent nodes, creating the further choice of whether a local scale
belongs to an observed entity, a parent, or an edge. None is a uniquely implied
"structured horseshoe." `batchieR` therefore supports entity-local horseshoes
only when $Q=I$.  The structured models use `global_half_cauchy()`, which
preserves $Q$ and lets the data learn one overall deviation scale per component dimension.

For a fair comparison, we first fit a compatible IID baseline that also uses
global half-Cauchy shrinkage. Comparing that baseline with the structured models
then isolates the effect of replacing $I$ by a biological precision $Q$; a
direct comparison with the package default would change both $Q$ and the
shrinkage family. This is why the comparison retains two baselines: the default
represents ordinary package behavior, while the compatible IID model provides
the clean $Q=I$ counterfactual for the structured fits. Means remain zero,
structures remain IID, and doses remain categorical in this baseline.  The cell
priors are written explicitly because they are retained throughout the remaining models.

```{r compatible-iid-model, echo=FALSE}
compatible_iid_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    dose = categorical()
)
```

### 2. Put measured features in the prior mean

Prepared cell covariates can define the mean of $W_0$, and compound
covariates the mean of $V_0$:

$$
W_{0k} = x_k^\top\beta_W + u_{Wk}, \qquad
V_{0i} = z_i^\top\beta_V + u_{Vi}.
$$

Offsets are the natural place for this regression because they are scalar,
identifiable contributions to the linear predictor: a coefficient in the
$W_0$ mean describes how a cell feature shifts baseline response, while a
coefficient in the $V_0$ mean describes how a compound feature shifts its
additive effect across cells. The factor variables $W$, $V_1$, and $V_2$, by
contrast, enter the predictor only through products and have sign and
permutation symmetries. A coefficient for, say, the first coordinate would
therefore change meaning under an equivalent representation that flips or
reorders the factors. Without additional identification or covariance choices,
feature means for factors are not well defined, so the current API keeps their
means at zero. Biological information can still enter factor components through
`structure`, which relates entities without assigning a biological label to a
particular latent coordinate.

Within `combo_gaussian_component()`, the two fields relevant to feature
regression have separate jobs:

- `mean` is a one-sided R formula that constructs $X$ and specifies
  $X\beta$. It must omit an intercept, for example
  `~ 0 + pathway_activity + growth_rate`.
- `mean_shrinkage` specifies the prior regularization of $\beta$. It is
  required when `mean` contains predictors; here `gamma_precision()` lets the
  data learn their common prior precision.

The component-level `shrinkage` argument has a different role: it controls the
residual $u$ around $X\beta$, not the feature coefficients. This separation
allows strong feature signal and small unexplained entity variation, or the
reverse, to be learned independently.

Biological covariates referenced by these formulas are supplied separately from
the experiment table when the model is fitted:

- a `cell_offset` formula is evaluated using `cell_data`, matched to model
  cells by its unique, nonmissing `cell` column;
- a `treatment_offset` formula is evaluated using `compound_data`, matched by
  its unique, nonmissing `drug` column. Compound annotations are then attached
  to every modeled dose of that compound.

Thus the example formulas obtain `pathway_activity` and `growth_rate` from
`cell_data`, and `target_signature` and `physchem_pc` from `compound_data`.
Both metadata tables are passed to `fit_combo()`. Variables used by a formula
must be present and finite for every relevant entity. The resulting design
columns are centered and scaled internally, so `mean_shrinkage` operates on a
common standardized scale. The principal benefit is transfer to sparsely
observed cells or compounds whose annotations are already known.

To isolate this change, `feature_model` modifies only `cell_offset` and
`treatment_offset` relative to `compatible_iid_model`: their residual
structures and shrinkage stay the same, but they gain `mean` and
`mean_shrinkage`. The `cell_factors`, `treatment_main_factors`, and
`treatment_interaction_factors` specifications remain unchanged, as does the
categorical dose model. Differences between these two fits therefore target
the contribution of the metadata-informed offset means.

```{r feature-model}
feature_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        mean = ~ 0 + pathway_activity + growth_rate,
        mean_shrinkage = gamma_precision(),
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        mean = ~ 0 + target_signature + physchem_pc,
        mean_shrinkage = gamma_precision(),
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    dose = categorical()
)
```

### 3. Partially pool doses within compounds

With `nested(precision = 4)`, a treatment value at dose $r$ is modeled around
a shared compound parent:

$$
V_{ir} = \bar V_i + e_{ir}, \qquad e_{ir} \mid \rho \sim
\mathcal N(0, (4\rho)^{-1} I).
$$

The same construction applies to $V_0$, $V_1$, and $V_2$. It reduces
uncertainty for a sparse or missing dose by borrowing from other doses of that
compound. The value `4` is relative to the component precision $\rho$ and
controls the strength of dose-to-parent pooling. It does not use numerical dose
distance, impose monotonicity, or fit a dose-response curve.

```{r nested-dose-model}
nested_dose_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        mean = ~ 0 + pathway_activity + growth_rate,
        mean_shrinkage = gamma_precision(),
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        mean = ~ 0 + target_signature + physchem_pc,
        mean_shrinkage = gamma_precision(),
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    dose = nested(precision = 4)
)
```

### 4. Relate entities through a graph or tree

Here we use prior biological information to say which cells and compounds are
expected to behave similarly. For latent factors, this information belongs in
the dependence structure, not in a feature-informed mean. The coordinates of
$W$, $V_1$, and $V_2$ can be flipped or reordered without changing the
response surface, so assigning a regression mean to a particular coordinate
would not have a stable interpretation. We therefore keep the factor means at
zero and encode relatedness only through their relative precision $Q$.

In `combo_model()`, the corresponding arguments are `cell_factors` ($W$),
`treatment_main_factors` ($V_1$), and `treatment_interaction_factors` ($V_2$).
The feature-informed offsets $W_0$ and $V_0$ remain as specified in the
preceding nested-dose model: $W_0$ stays IID, while $V_0$ keeps an IID
compound-parent structure and the existing dose hierarchy. Thus the GMRF and
tree fits isolate the effect of relating latent representations rather than
changing their mean or the additive offsets.

#### Choosing a representation

Biological relatedness can arrive in several forms. A lineage, ontology, or
target classification supplies a hierarchy; a pathway or interaction database
supplies a network; and expression or chemical descriptors supply continuous
pairwise similarities. The representation should follow that source rather
than forcing every annotation into the same matrix.

| Prior information | Representation | Package interface |
|:--|:--|:--|
| No trusted relationship | Independent entities | `iid()` |
| A known hierarchy | Gaussian tree | `tree()` constructs the precision |
| A pairwise network or sparse similarities | GMRF | construct an operator, then use `gmrf()` |
| A complete precision already exists | Direct precision | `precision(Q)` |

A tree and a GMRF express different assumptions. A tree says that entities
inherit variation along shared ancestry. A GMRF says that neighboring entities
should have similar values, conditional on the rest of the graph. Neither is
intrinsically more biological; the appropriate choice depends on how the prior
knowledge was obtained.

#### A hierarchy with `tree()`

For arbitrary cell or compound relationships, `tree()` is currently the only
package helper that constructs a structured precision from a relational
description. The user supplies a data frame with `node` and `parent` columns
and, optionally, a positive `edge_length`. Nodes must be unique, nonmissing
parents must exist, cycles are invalid, and every modeled cell or compound must
be a terminal node. Internal parents become latent Gaussian nodes.

Each child is a Gaussian increment from its parent. Edge precision is the
inverse of edge length, so shorter edges express stronger similarity; leaves
covary through their shared ancestral path. The resulting joint precision is
sparse, with one connection per parent-child edge. In the example below, cells
share lineage parents and compounds share mechanism parents.

#### Pairwise relationships with a GMRF

A GMRF is useful when prior information is naturally an edge list or pairwise
similarity graph. `batchieR` does not currently construct this operator from
annotations: the user builds a sparse operator and passes it to `gmrf()`. A
common choice is the weighted graph Laplacian

$$
L = \operatorname{diag}(A\mathbf 1)-A,
$$

for a symmetric adjacency matrix $A$ with nonnegative weights and zero diagonal.
It penalizes disagreement across strongly weighted edges. The example helper
below converts an edge list to an unweighted Laplacian. It connects cells within
lineages and compounds within mechanism pairs.

The code defines the tree first because it uses the package-provided helper; the
GMRF construction that follows is vignette-local code that users would replace
with an operator derived from their own biological graph.

```{r structures}
cell_tree_data <- data.frame(
    node = c(".cell_root", "lineage_A", "lineage_B", cells),
    parent = c(NA, ".cell_root", ".cell_root", rep("lineage_A", 4), rep("lineage_B", 4)),
    edge_length = c(1, 1, 1, rep(0.05, 8)),
    stringsAsFactors = FALSE
)
compound_tree_data <- data.frame(
    node = c(".drug_root", sprintf("MOA_%d", 1:4), drugs),
    parent = c(NA, rep(".drug_root", 4), rep(sprintf("MOA_%d", 1:4), each = 2)),
    edge_length = c(1, rep(1, 4), rep(0.05, 8)),
    stringsAsFactors = FALSE
)
cell_tree <- tree(node, parent, cell_tree_data, edge_length)
compound_tree <- tree(node, parent, compound_tree_data, edge_length)

graph_laplacian <- function(nodes, edges) {
    adjacency <- sparseMatrix(
        i = match(edges[, 1], nodes),
        j = match(edges[, 2], nodes),
        x = 1,
        dims = rep(length(nodes), 2),
        dimnames = list(nodes, nodes)
    )
    adjacency <- adjacency + t(adjacency)
    operator <- Diagonal(
        length(nodes),
        x = rowSums(adjacency)
    ) - adjacency
    dimnames(operator) <- list(nodes, nodes)
    drop0(operator)
}

cell_edges <- do.call(rbind, lapply(split(cells, cell_data$lineage), function(group) {
    t(combn(group, 2))
}))
compound_edges <- within_pairs
cell_gmrf <- gmrf(8 * graph_laplacian(cells, cell_edges), ridge = 0.5)
compound_gmrf <- gmrf(8 * graph_laplacian(drugs, compound_edges), ridge = 0.5)
```

#### Applying a structure to the factor components

For factor dimension $d$, `structure` supplies the relative precision $Q$
and `shrinkage` supplies its precision multiplier $\rho_d$:

$$
u_{\cdot d} \mid \rho_d,Q \sim \mathcal N\left(0,(\rho_d Q)^{-1}\right).
$$

A non-IID factor structure can use `fixed_scale()`, `gamma_precision()`,
`global_half_cauchy()`, or `multiplicative_gamma()`. Entity-local
`local_half_cauchy()` and `horseshoe()` scales are not compatible because they
do not multiply $Q$ as a whole; they would define a different node- or
edge-scaled prior. Here `cell_factors` retains `multiplicative_gamma()`, while
`treatment_main_factors` and `treatment_interaction_factors` use `global_half_cauchy()`.

The full specifications below retain the feature-informed offsets and nested
doses from the preceding model, changing only the three factor structures.

#### GMRF model

```{r gmrf-model}
gmrf_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        mean = ~ 0 + pathway_activity + growth_rate,
        mean_shrinkage = gamma_precision(),
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = cell_gmrf,
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        mean = ~ 0 + target_signature + physchem_pc,
        mean_shrinkage = gamma_precision(),
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = compound_gmrf,
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = compound_gmrf,
        shrinkage = global_half_cauchy()
    ),
    dose = nested(precision = 4)
)
```

#### Tree model

```{r tree-model}
tree_model <- combo_model(
    rank = 3,
    cell_offset = combo_gaussian_component(
        mean = ~ 0 + pathway_activity + growth_rate,
        mean_shrinkage = gamma_precision(),
        structure = iid(),
        shrinkage = gamma_precision()
    ),
    cell_factors = combo_gaussian_component(
        structure = cell_tree,
        shrinkage = multiplicative_gamma()
    ),
    treatment_offset = combo_gaussian_component(
        mean = ~ 0 + target_signature + physchem_pc,
        mean_shrinkage = gamma_precision(),
        structure = iid(),
        shrinkage = global_half_cauchy()
    ),
    treatment_main_factors = combo_gaussian_component(
        structure = compound_tree,
        shrinkage = global_half_cauchy()
    ),
    treatment_interaction_factors = combo_gaussian_component(
        structure = compound_tree,
        shrinkage = global_half_cauchy()
    ),
    dose = nested(precision = 4)
)
```

#### Technical requirements and scaling

For `gmrf(operator, ridge)`, the package constructs

$$
Q = \text{operator} + \text{ridge} I.
$$

The operator must be a nonempty square `Matrix` sparse matrix with finite entries.
Its row and column names must be identical, unique, nonmissing, and collectively
match the modeled cells or compounds exactly. It must be exactly symmetric, and
after adding the finite positive ridge, $Q$ must be positive definite. The
operator itself may be singular: a graph Laplacian is positive semidefinite, and
the ridge makes its Gaussian prior proper. If a complete sparse
positive-definite precision already exists, `precision(Q)` accepts it directly
without adding a ridge.

The Gibbs updates traverse the nonzero neighbors of $Q$, so memory and update
cost depend on its actual nonzero pattern, not merely on use of a sparse-matrix
class. A dense similarity matrix therefore should not be passed through as a
dense graph: first retain a sparse set of biologically meaningful edges, then
construct the graph Laplacian.

For continuous annotations, packages such as
[`kernlab`](https://search.r-project.org/CRAN/refmans/kernlab/html/kernelMatrix.html)
can construct a kernel Gram matrix $K$. Such matrices are typically dense
similarities rather than precision operators, and $(K+\epsilon I)^{-1}$ is
usually dense.

Thresholding similarities to construct a graph is different. Threshold $K$
into a symmetric, nonnegative adjacency $A$ with zero diagonal, then recompute
$L=\operatorname{diag}(A\mathbf 1)-A$. Because

$$
x^\top Lx = \frac{1}{2}\sum_{i,j}A_{ij}(x_i-x_j)^2 \geq 0,
$$

this $L$ is positive semidefinite and `gmrf(L, ridge > 0)` produces a proper
precision. Common heuristics include similarity thresholds, radius graphs,
k-nearest-neighbor (k-NN) graphs, and shared-nearest-neighbor (SNN) graphs. The
resulting graph should be symmetrized before constructing its Laplacian. These
choices define which entities directly share information and should be treated
as part of the prior specification.

In the example, `gmrf(8 * L, ridge = 0.5)` gives $Q=8L+0.5I$: the multiplier
controls the relative edge penalty, the ridge supplies a weaker common anchor,
and component `shrinkage` learns the overall magnitude of the factor
deviations.

#### Comparing the induced structures

The heatmaps below show the prior correlation between entity effects after
removing the component-specific scale. The cell kernel applies to every column
of $W$; the compound kernel applies to every column of $V_1$ and $V_2$.
The offsets do not use either kernel. For nested treatment factors, the
compound heatmap is the parent-level kernel before dose nodes are attached.

```{r structure-kernels, echo=FALSE, fig.height=6, fig.alt="Prior correlation heatmaps for cell and compound effects under GMRF and tree structures"}
gmrf_correlation <- function(spec) {
    precision <- spec$operator +
        Diagonal(nrow(spec$operator), x = spec$ridge)
    cov2cor(solve(as.matrix(precision)))
}

tree_correlation <- function(spec, leaves) {
    paths <- lapply(leaves, function(leaf) {
        path <- character()
        current <- leaf
        while (!is.na(current)) {
            path <- c(path, current)
            current <- spec$parent[match(current, spec$node)]
        }
        path
    })

    covariance <- outer(
        seq_along(leaves),
        seq_along(leaves),
        Vectorize(function(i, j) {
            shared_path <- intersect(paths[[i]], paths[[j]])
            sum(spec$edge_length[match(shared_path, spec$node)])
        })
    )
    dimnames(covariance) <- list(leaves, leaves)
    cov2cor(covariance)
}

kernel_table <- function(kernel, prior, entities) {
    result <- as.data.frame(as.table(kernel), stringsAsFactors = FALSE)
    names(result) <- c("row", "column", "correlation")
    result$prior <- prior
    result$entities <- entities
    result
}

kernel_data <- rbind(
    kernel_table(gmrf_correlation(cell_gmrf), "GMRF", "Cells"),
    kernel_table(gmrf_correlation(compound_gmrf), "GMRF", "Compounds"),
    kernel_table(tree_correlation(cell_tree, cells), "Tree", "Cells"),
    kernel_table(tree_correlation(compound_tree, drugs), "Tree", "Compounds")
)
kernel_data$row <- factor(
    kernel_data$row,
    levels = c(rev(drugs), rev(cells))
)
kernel_data$column <- factor(
    kernel_data$column,
    levels = c(cells, drugs)
)

ggplot(kernel_data, aes(column, row, fill = correlation)) +
    geom_tile(colour = "white") +
    facet_wrap(vars(entities, prior), scales = "free") +
    scale_fill_gradient(
        low = "white",
        high = "#2166ac",
        limits = c(0, 1)
    ) +
    labs(x = NULL, y = NULL, fill = "Prior\ncorrelation") +
    theme_minimal() +
    theme(
        panel.grid = element_blank(),
        aspect.ratio = 1
    )
```

The GMRF has block-diagonal correlation because its graph connects entities
only within a lineage or mechanism; the ridge makes each block nonsingular.
The tree also correlates entities from different groups through their shared
root, while short terminal edges make siblings especially similar. These are
different covariance geometries and should be treated as alternative
assumptions rather than combined automatically.

## What does the prior buy us?

We fit all six models to the same observed rows. Sampler mechanics are covered
in the introduction; the short run below exists only to keep the vignette
executable and is not sufficient for a biological analysis.

```{r fit-models, include=FALSE}

models <- list(
    default = default_model,
    compatible_iid = compatible_iid_model,
    features = feature_model,
    nested_dose = nested_dose_model,
    gmrf = gmrf_model,
    tree = tree_model
)
fit_data <- screen[c(prediction_columns, "response")]

set.seed(20260900)
fits <- lapply(
    models,
    fit_combo,
    data = fit_data,
    cell_data = cell_data,
    compound_data = compound_data,
    chains = 2,
    parallel_chains = 1,
    iter_warmup = 100,
    iter_sampling = 150,
    thin = 1,
    refresh = if (interactive()) 25 else 0
)
names(fits) <- names(models)
```


The staged design makes each comparison local: first inspect whether the
over-specified rank is shrunk, then whether metadata coefficients recover the
known signal, then whether each structural feature helps in its intended
holdout regime, and finally whether those posterior changes alter plate choice.

### Latent-dimension shrinkage

The default cell-factor prior uses multiplicative-gamma shrinkage. Posterior
global precision by dimension shows how the model can regularize a rank-three
fit when the data-generating response surface has only two dimensions.

```{r shrinkage-summary, echo=FALSE}
summarize_mapped_draws <- function(fit, selected_map) {
    selected_draws <- posterior_draws(fit, variable = selected_map$variable)
    draw_matrix <- matrix(
        selected_draws,
        ncol = length(selected_map$variable),
        dimnames = list(NULL, selected_map$variable)
    )
    data.frame(
        selected_map,
        posterior_mean = colMeans(draw_matrix),
        q05 = apply(draw_matrix, 2, quantile, 0.05),
        q95 = apply(draw_matrix, 2, quantile, 0.95),
        row.names = NULL
    )
}

default_map <- parameter_map(fits$default)
shrinkage_map <- subset(
    default_map,
    component == "cell_factors" & parameter == "global_precision"
)
shrinkage_summary <- summarize_mapped_draws(fits$default, shrinkage_map)
shrinkage_summary[c("component", "parameter", "dimension", "posterior_mean", "q05", "q95")]
```

### Feature coefficient recovery

Metadata formulas affect additive offsets only. Their coefficients are on the
internally standardized feature scale. In this controlled example, the
posterior should identify `pathway_activity` and `target_signature` more
clearly than the two nuisance features.

```{r feature-coefficients, echo=FALSE}
feature_map <- subset(
    parameter_map(fits$features),
    parameter == "mean_coefficient"
)
feature_summary <- summarize_mapped_draws(fits$features, feature_map)
feature_summary[c(
    "component", "parameter", "feature", "posterior_mean", "q05", "q95"
)]
```

### Prediction in the regime each feature targets

RMSE is measured against the noise-free synthetic mean. Coverage uses the
noisy held-out response and replicated posterior observations. Mean posterior
SD measures epistemic uncertainty in the expected response.

```{r prediction-metrics, echo=FALSE}
evaluation_roles <- c(
    "warm_holdout", "cold_start", "dose_transfer", "related_transfer"
)
evaluation_indices <- which(screen$role %in% evaluation_roles)
evaluation_data <- screen[evaluation_indices, prediction_columns]

metric_rows <- lapply(seq_along(fits), function(fit_index) {
    fit <- fits[[fit_index]]
    expected <- posterior_epred(fit, newdata = evaluation_data)
    replicated <- posterior_predict(fit, newdata = evaluation_data)
    stopifnot(all(is.finite(expected)), all(is.finite(replicated)))

    do.call(rbind, lapply(evaluation_roles, function(role) {
        selected <- screen$role[evaluation_indices] == role
        intervals <- apply(
            replicated[, selected, drop = FALSE],
            2,
            quantile,
            c(0.05, 0.95)
        )
        data.frame(
            model = names(fits)[fit_index],
            regime = role,
            n = sum(selected),
            latent_rmse = sqrt(mean(
                (colMeans(expected[, selected, drop = FALSE]) -
                     screen$true_mean[evaluation_indices][selected])^2
            )),
            coverage_90 = mean(
                screen$modeled_response[evaluation_indices][selected] >= intervals[1, ] &
                    screen$modeled_response[evaluation_indices][selected] <= intervals[2, ]
            ),
            mean_posterior_sd = mean(apply(
                expected[, selected, drop = FALSE],
                2,
                sd
            )),
            stringsAsFactors = FALSE
        )
    }))
})
prediction_metrics <- do.call(rbind, metric_rows)
rownames(prediction_metrics) <- NULL
prediction_metrics

target_rmse <- function(model, regime) {
    prediction_metrics$latent_rmse[
        prediction_metrics$model == model & prediction_metrics$regime == regime
    ]
}
```

```{r prediction-metrics-plot, echo=FALSE, fig.alt="Held-out latent-mean RMSE by model and evaluation regime"}
ggplot(prediction_metrics, aes(x = model, y = latent_rmse, fill = model)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(vars(regime), scales = "free_y") +
    coord_flip() +
    labs(x = NULL, y = "RMSE against true logit mean")
```

The comparisons should be read locally rather than as a tournament. The
feature ablation targets cold starts, the nested-dose ablation targets missing
doses, and the two relationship models target transfer among related entities.
An informed prior need not improve ordinary warm holdouts.

### Plate selection

All models score the same four eight-experiment plates against the complete
320-row response surface. Distances are measured after converting expected
logit responses back to viability. Candidate likelihoods remain Gaussian on
the fitted logit scale. The candidates represent observed within-mechanism
replication, eight replicate wells for one held-out high-dose combination, a
related-entity transfer plate for `D01`/`D08`, and a wholly untested
`D03`/`D05` cross-mechanism pair. The last two plates use eight replicate wells
at high dose in cold-start cell `C04`, holding cell, dose, and replication fixed
while changing the biological question. Both compounds on the discovery plate
are observed in other contexts.

```{r score-candidates}
reference_grid <- screen[prediction_columns]
score_rows <- lapply(seq_along(fits), function(fit_index) {
    result <- score_plate_pdbal(
        fits[[fit_index]],
        candidate_data = candidate_data,
        reference_grid = reference_grid,
        response_transform = plogis,
        max_triplets = 1000
    )
    result$model <- names(fits)[fit_index]
    result
})
plate_scores <- do.call(rbind, score_rows)
rownames(plate_scores) <- NULL

plate_ranks <- reshape(
    plate_scores[c("model", "plate", "rank")],
    idvar = "plate",
    timevar = "model",
    direction = "wide"
)
names(plate_ranks) <- sub("rank[.]", "", names(plate_ranks))
plate_ranks
```

Only ranks within a fitted model are comparable. The unnormalized log PDBAL
score includes posterior distances, likelihood overlap, and predictive
entropy, all of which change when the model changes. Comparing raw score
magnitudes across columns would therefore be misleading.

```{r selected-plates, echo=FALSE}
selected_plates <- plate_scores[plate_scores$rank == 1, c("model", "plate")]
selected_plates
```

The default model is attracted to a plate dominated by prior-driven gaps. Once
known lineage, mechanism, and dose relationships propagate through the
posterior, both structured alternatives prioritize the cross-mechanism plate:
it asks about genuinely new interactions rather than re-learning relationships
already encoded by the annotations. This is the practical value of informing
the model before scoring.

## Boundaries and practical guidance

- Mean formulas currently inform `cell_offset` and `treatment_offset`; latent
  factor means remain zero.
- `nested()` pools compound-dose treatments around a compound parent. It does
  not use numerical dose distance, impose monotonicity, or fit a dose-response
  curve.
- `tree()` and `gmrf()` are alternative encodings. Choose one whose geometry
  matches the scientific relationship; adding both is not automatically more
  informative.
- Supplied relationships are assumptions. Check transfer performance on
  biologically meaningful holdouts, and compare against an IID model before
  using them for acquisition.
- Prediction and scoring rows may rearrange fitted entities but cannot
  introduce cells or compound-dose treatments absent from the fit. Include
  future entities as missing-response rows when the screen is compiled.
- The Gaussian identity-link model sees the logit-viability response supplied
  here. `plogis()` changes the interpretation scale for distances; it does not
  turn the likelihood into a beta or binomial model.

## Session information

```{r session-information}
sessionInfo()
```
