---
title: "Introduction to batchieR"
author: "Teo Sakel"
date: "2026-08-04"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Introduction to batchieR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4.5
)
```

## Introduction

`batchieR` fits Bayesian latent-factor models to continuous responses from combination screens.
It separates variation associated with cellular context, treatment main effects,
and treatment interactions, and uses the fitted posterior to compare possible
follow-up plates. This is useful when cells and compound-dose treatments recur
across a screen and information can be shared across experiments.

This vignette shows how to prepare the required experiment table, understand
the default model, run a small fit, assess it with posterior predictive checks,
predict withheld responses, and rank candidate plates with
posterior-distance-based active learning (PDBAL).

The package currently supports **only a Gaussian response with conjugate priors**.
Binary, binomial, beta, count, and other non-Gaussian response families are not implemented.

## Installation

Install the current development version from GitHub:

```{r install, eval=FALSE}
remotes::install_github("TeoSakel/batchieR")
```

```{r load-package}
library(batchieR)
```

## Model

`batchieR` implements the BATCHIE model of Tosh et al. (2025).
For an observed response involving cell $k$ and treatments $i$ and $j$,
the current observation model is

$$
y_{kij} \mid \eta_{kij}, \tau \sim \mathcal{N}(\eta_{kij}, \tau^{-1}),
$$

Since the response variable $y_{kij}$ is modeled as Gaussian, it is assumed to be continuous and unbounded.
However, typically in drug screen experiments, response represents cell viability,
which is naturally bounded between 0 and 1. Thus, the values have to be transformed
to the real line before fitting the model.  Following the original BATCHIE paper,
the suggested transformation is the logit, which is implemented in R as `qlogis()`.
This is not a binomial model but rather a Gaussian model on the logit scale.
Model predictions can then be transformed back to the viability scale with `plogis()`.

$\tau$ is the observation precision and the mean $\eta_{kij}$ is the linear predictor

$$
\eta_{kij} = \alpha + x_{kij}^{\mathsf T}\beta
    + W^{(0)}_k + V^{(0)}_i + V^{(0)}_j
    + \left(W^{(1)}_k\right)^\top\left(V^{(1)}_i + V^{(1)}_j + V^{(2)}_i\odot V^{(2)}_j\right).
$$

The systematic component $\alpha + x_{kij}^{\mathsf T}\beta$ is expressed in
the observed covariate space supplied by the user: $\alpha$ sets the overall
level and $x_{kij}^{\mathsf T}\beta$ accounts for optional observation, cell,
and compound metadata.

The remaining terms capture cell and treatment effects and their interactions.
In particular, the rows of $W^{(1)}$, $V^{(1)}$, and $V^{(2)}$ represent cells
and treatments in a shared $d$-dimensional latent space learned from the data.
Superscripts label component types, subscripts index entities, and $\odot$
denotes element-wise multiplication.

`combo_model()` encodes this model using the following arguments:

| Model argument | Symbol | Dimension | Role |
|:--|:--|:--|:--|
| `mean` | $\alpha + x_{kij}^{\mathsf T}\beta$ | scalar per observation | Overall level and optional metadata regression |
| `cell_offset` | $W^{(0)}$ | $K$ | Baseline differences among cells |
| `cell_factors` | $W^{(1)}$ | $K \times d$ | Latent characteristics of cells |
| `treatment_offset` | $V^{(0)}$ | $T$ | Treatment effects shared across cells |
| `treatment_main_factors` | $V^{(1)}$ | $T \times d$ | Cell-dependent treatment main effects |
| `treatment_interaction_factors` | $V^{(2)}$ | $T \times d$ | Cell-dependent nonadditive combination effects |

Here $K$ is the number of cells, $T$ is the number of treatments (compound-dose pairs),
and $d$ is the common latent dimension shared by the cell and treatment representations.
The `rank` argument of `combo_model()` controls $d$. For monotherapies (single drug treatments),
the second-treatment terms and the treatment-interaction term are omitted.

### Systematic mean

The systematic mean plays a role similar to *fixed effects* in mixed effects models.
`batchieR` offers 3 ways to specify the base mean value of the response:

- `empirical_mean()`: fixes $\alpha$ at the mean of the mean of the observed responses
  and $\beta$ at zero. This is the default option implemented in the original BATCHIE.
- `fixed_mean(value)`: fixes $\alpha$ at the given `value` and $\beta$ at zero.
- `formula_mean()`: uses the supplied covariates $x_{kij}$ to learn $\alpha$ and $\beta$ from data.
  see `vignette("biologically_informed_models", package = "batchieR")` for details.

### Latent components

The remaining five components play a role similar to *random effects*, capturing
cell and treatment variation beyond the systematic mean. Thus, their Gaussian priors
are centered at zero to express residual variation around the systematic component.
The offsets resemble random intercepts, while the latent factors describe
interactions through products of learned representations.

All five components are specified using `combo_gaussian_component()`, which
separates their dependence structure from their magnitude through two arguments:

- `structure` describes relationships among cells or treatments and their relative prior precision.
  By default (`iid`) entities are treated as independent.
- `shrinkage` controls the magnitude of deviations around zero,
  determining how strongly component values are pulled toward zero.

Setting a component argument to `NULL` removes its term from the predictor.
Treatment factor components require `cell_factors` because they enter the
predictor through their product with $W^{(1)}_k$. For more detail on using known
biological relationships to specify `structure`, see the
`vignette("biologically_informed_models", package = "batchieR")`.

### Default specification

By default, `combo_model()` constructs a rank-12 model as specifid in the original BATCHIE implementation.

```{r model-specification}
combo_model()
```

The components use independent (`iid()`) structures and the following priors:

- `cell_offset`: gamma precision shrinkage.
- `cell_factors`: multiplicative-gamma shrinkage.
- `treatment_offset`, `treatment_main_factors`, and `treatment_interaction_factors` use horseshoe shrinkage.
- The Gaussian observation precision has a gamma prior.

The default dose specification, `categorical()`, treats each exact
compound-dose pair as a separate treatment, with no dose ordering,
interpolation, or automatic sharing between doses.

The published default is useful for reproducing BATCHIE analyses, but its
entity-local horseshoe scales add substantial posterior geometry. A practical
workflow should begin with the simpler global half-Cauchy prior and add local
scales only when they improve predictive performance.

For this tutorial we use rank 2, fix the systematic mean at 1.5^[in logit scale, `r round(plogis(1.5), 2)` viability],
and assign one global half-Cauchy scale per treatment-factor dimension:

```{r example-model}
treatment_component <- combo_gaussian_component(
    shrinkage = global_half_cauchy(scale = 1)
)
model <- combo_model(
    rank = 2,
    mean = fixed_mean(1.5),
    treatment_offset = treatment_component,
    treatment_main_factors = treatment_component,
    treatment_interaction_factors = treatment_component
)
model
```

Use `update()` to generate a new model by modifying selected components of an existing
model specification. Components are replaced in full, and `NULL` disables a component.

```{r update-model-specification}
revised_model <- update(model, rank = 4, treatment_interaction_factors = NULL)
revised_model
```

## Generating synthetic data

To illustrate the workflow, we will generate a synethetic experiment with three
cells (`A`, `B`, and `C`) and four drugs (`W`, `X`, `Y`, and `Z`), each used at
a single dose. We organize the synthetic experiments into seven plates:

- **Monotherapy (plate 1):** each drug is tested alone in each cell, giving 12 experiments.
- **Combination training (plates 2–4):** each plate tests `X+Y` and `Z+W` in cells A and B.
- **Holdout (plate 5):** four independent replicate measurements of the same
  cell–pair combinations as the training plates. Their responses are withheld
  from fitting and reserved for evaluating predictions of repeat experiments.
- **Candidate (plates 6–7):** experiments available to run next, ranked by
  their expected informativeness. Their simulated responses are withheld from
  fitting to represent measurements that have not yet been collected.

The two candidate plates explore different gaps in the training data. Each
contains two distinct cell–pair combinations measured twice, for four rows:

- **Known pairs in a new cell context (plate 6):** `X+Y` and `Z+W`, each
  measured twice in cell C. Cell C has monotherapy training data, but no
  observed combination responses.
- **New pairs in known cell contexts (plate 7):** `X+Z` in cell A and `Y+W`
  in cell B, each measured twice. Both cells have combination-training data,
  but neither drug pair has been observed in training.

We also include the holdout plates' design in the follow-up candidates
to act as an "uninformative" baseline.

```{r synthetic-data}
cells <- c("A", "B", "C")
drugs <- c("X", "Y", "Z", "W")

single_drug <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    drug_2 = NA_character_,
    plate = "1",
    stringsAsFactors = FALSE
)

known_pairs <- data.frame(
    cell = c("A", "A", "B", "B"),
    drug_1 = c("X", "Z", "X", "Z"),
    drug_2 = c("Y", "W", "Y", "W"),
    stringsAsFactors = FALSE
)

# training_pairs = known_pairs in triplicates
training_pairs <- do.call(rbind, replicate(3, known_pairs, simplify = FALSE))
training_pairs$plate <- as.character(rep(2:4, each = nrow(known_pairs)))
holdout_pairs <- known_pairs
holdout_pairs$plate <- "5"

candidate_new_cells <- data.frame(
    cell = rep("C", 4),
    drug_1 = c("X", "Z", "X", "Z"),
    drug_2 = c("Y", "W", "Y", "W"),
    plate = "6",
    stringsAsFactors = FALSE
)
candidate_new_pairs <- data.frame(
    cell = c("A", "A", "B", "B"),
    drug_1 = c("X", "X", "Y", "Y"),
    drug_2 = c("Z", "Z", "W", "W"),
    plate = "7",
    stringsAsFactors = FALSE
)
screen <- rbind(
    single_drug,
    training_pairs,
    holdout_pairs,
    candidate_new_cells,
    candidate_new_pairs
)
plate_roles <- c(
    "1" = "training", "2" = "training", "3" = "training",
    "4" = "training", "5" = "holdout", "6" = "new_cell",
    "7" = "new_drug_pairs"
)
screen$role <- unname(plate_roles[screen$plate])
screen$dose_1 <- 1
screen$dose_2 <- ifelse(is.na(screen$drug_2), NA_real_, 1)
```

Once we specify the experimental design, we can simulate responses from a model
using `prior_predict`, which draws one or more set of model parameters from
their priors and generates responses for the supplied design. Since the response
is sampled from a Gaussian, it is not in viability or any other typical scale
used in biological experiments.

```{r simulate-responses}
simulated_response <- prior_predict(
    model,
    screen,
    draws = 1,
    type = "response",
    seed = 20260725
)
simulated_response <- as.numeric(simulated_response[1, ])  # flatten
screen$response <- ifelse(screen$role == "training", simulated_response, NA_real_)

training_cols <- c("cell", "drug_1", "dose_1", "drug_2", "dose_2", "response")
experiments <- screen[training_cols]
screen[c("plate", "role", training_cols)]
```

The training data (`experiments`) must contain `cell`, `drug_1`, `dose_1`,
`drug_2`, `dose_2`, and numeric `response` columns. In this example, `plate` and
`role` are organizational labels not used in the model and are thus omitted.

A missing (`NA`) response keeps a holdout or candidate row available for prediction
without adding it to the likelihood. For a single-treatment experiment, the drug and
dose in the absent arm must both be `NA`; partially missing treatment arms are invalid.

## Fit the model

To sample from the posterior distribution, we use `fit_combo()` with the `model`
and `experiments` objects constructed above. A Gibbs sampler is currently the
only sampler implemented.

`fit_combo()` uses familiar MCMC sampling controls, with argument names similar
to `CmdStanR`: `chains` sets the number of chains, while `iter_warmup` and
`iter_sampling` control their length. Here we run two short chains and set a
seed for reproducibility. The goal of this deliberately inexpensive fit is to
demonstrate the API; not to accurately represent the posterior.

```{r fit-model}
fit_chains <- 2
fit_parallel_chains <- min(fit_chains, future::availableCores())

fit <- fit_combo(
    model,
    experiments,
    chains = fit_chains,
    parallel_chains = fit_parallel_chains,
    iter_warmup = 100,
    iter_sampling = 200,
    thin = 1,
    seed = 20260725,
    refresh = if (interactive()) 10 else 0
)
```

The result is a `combo_fit` object containing the posterior draws and the
information needed for prediction and model checking. Printing it gives an
overview of the data and sampling run, followed by compact convergence
diagnostics across the public posterior variables. The diagnostics report the

- maximum R-hat,
- minimum bulk and tail ESS, and
- counts of unavailable values across public posterior variables

```{r print-fit}
fit
```

`summary()` provides estimates, credible intervals, and diagnostics for
individual public variables or groups selected by name prefix. Exact variable
names select only that variable; otherwise `select` expands to every available
name starting with the supplied text. Overlapping selections are included once,
and unmatched entries are errors.

```{r summary-fit}
summary(fit, select = c("sigma", "observation_precision", "main_effect", "interaction_effect"))
```

The factorization is not uniquely identified. Sign changes, permutations,
rotations, and compensating rescalings can change individual entries of
$W^{(1)}$, $V^{(1)}$, and $V^{(2)}$ without changing the scientifically
meaningful response surface. Consequently, routine convergence assessment
should emphasize reconstructed main effects, interaction effects, and
predictions rather than requiring every raw factor coordinate to agree.
The component terms `main_effect[combo_id]` and `interaction_effect[combo_id]`
are calculated within each posterior draw as

$$
\begin{aligned}
    \mathrm{main\_effect}_{kij}
        &= \sum_{h=1}^{d} W^{(1)}_{kh}\left(V^{(1)}_{ih} + V^{(1)}_{jh}\right), \\
    \mathrm{interaction\_effect}_{kij}
        &= \sum_{h=1}^{d} W^{(1)}_{kh} V^{(2)}_{ih} V^{(2)}_{jh}.
\end{aligned}
$$

Here `combo_id` is the integer index corresponding to the cell-treatment
combination $kij$, with $(k,i,j)$ and $(k,j,i)$ sharing the same ID.

```{r combo-map-fit}
combo_map(fit)
```

Besides, `combo_map` which reports unique cell and drug-dose pair combinations,
`cell_map()` and `treatment_map()` can be used to identify unique cells
and treatments (drug-dose) to variable indices. `row_map()` connects original
input rows to these combinations including replicates

A `combo_fit` object is also compatible with the standard `broom` verbs:
`tidy()`, `glance()`, and `augment()`. The API is similar to `rstan::tidy()` and
`broom.mixed::tidy()`, with a `select` argument accepting variable names or prefixes.
For example, we can summarize the cell offsets and their shrinkage precisions
(use `"cell_offset["` to select only offset values):

```{r tidy-fit}
tidy(fit, select = "cell_offset")
```

Individual factors and their shrinkage precisions are excluded from the public defaults.
They remain available through `include = "all"` for specialized investigations,
but poor diagnostics for those coordinates alone do not establish a problem with
identifiable effects or predictions. The [model-checking vignette](model_checking.html)
develops the corresponding multi-chain diagnostic workflow.

## Posterior predictive checks

Posterior predictive checks compare observed logit responses with responses
replicated from the fitted Gaussian model. They use only rows with observed
responses and stay on the scale supplied to `fit_combo()`.

```{r pp-check-density, fig.alt="Density overlay of observed and posterior-predictive logit responses"}
bayesplot::pp_check(fit, type = "dens_overlay", ndraws = 20)
```

```{r pp-check-ecdf, fig.alt="ECDF overlay of observed and posterior-predictive logit responses"}
bayesplot::pp_check(fit, type = "ecdf_overlay", ndraws = 20)
```

Systematic shifts between observed and replicated curves suggest a location mismatch.
Different shapes or tails suggest that the latent structure or the Gaussian
observation model misses important aspects of the data.

For more details, including LOO-PIT calibration, grouped checks, predictive
errors, test statistics, and guidance on interpreting diagnostic patterns,
see `vignette("model_checking", package = "batchieR")`.

### Predicting withheld responses

The holdout plate lets us compare predictions with measurements that were
excluded from fitting. `posterior_epred()` returns draws of the expected response,
while `posterior_predict()` also includes Gaussian observation noise for a
future measurement. Both return a matrix with posterior draws in rows and
experiments in columns, on the fitted scale.

Here we use `posterior_predict()` to obtain a median and a 95% predictive
interval for each holdout measurement. We compare these with the simulated
responses saved before masking the holdout rows:

```{r holdout-predictions}
prediction_columns <- setdiff(training_cols, "response")
holdout_rows <- screen$role == "holdout"
holdout_data <- screen[holdout_rows, prediction_columns]

holdout_predictions <- fit |>
    posterior_predict(newdata = holdout_data) |>
    apply(2, quantile, probs = c(0.5, 0.025, 0.975)) |>
    t()
holdout_predictions <- cbind(simulated_response[holdout_rows], holdout_predictions)
colnames(holdout_predictions) <- c("observed", "predicted", "lower", "upper")
holdout_predictions <- cbind(holdout_data, holdout_predictions)
holdout_predictions
```

These intervals describe repeat measurements of training conditions. To
summarize uncertainty in their expected responses instead, use `posterior_epred()`
in the same calculation. To report future measurements on the viability scale,
apply `plogis()` (assuming that the data are in logit-scale) to the predictive
draws before summarizing.

## Rank candidate plates with PDBAL

Given a set of candidate plates, which should we run next? Here a *plate* is
a set of experiments performed together, each specifying a cell and one or
two treatments at particular doses. Selecting a plate means selecting the
whole batch of experiments.

We want to choose the plate expected to be most informative about the response
surface. Following BATCHIE, we use PDBAL to rank candidates by their expected
ability to distinguish competing predictions from the fitted posterior. A
`reference_grid` specifies the experiments over which this disagreement
matters: the candidates define what we could measure next, while the grid
defines where we want predictions to improve. For this illustration, we include
every cell with all six unordered drug pairs, for 18 combination experiments in total.

```{r reference-grid}
reference_grid <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    dose_1 = 1,
    drug_2 = drugs,
    dose_2 = 1,
    stringsAsFactors = FALSE
)
reference_grid <- subset(reference_grid, drug_1 < drug_2)
reference_grid
```

We evaluate the following candidate plates:

- **Plate 5 (holdout):** repeats the four training combinations once each.
  Acts as a baseline of comparison.
- **Plate 6 (new cell context for combinations):** tests the known pairs
  (`X+Y` and `Z+W`) in cell C, which already has monotherapy training data,
  with two measurements of each combination.
- **Plate 7 (New drug-pairs):** tests the new pair `X+Z` in cell A and the new pair
  `Y+W` in cell B, with two measurements of each cell–pair combination.

Only the experimental designs are passed to scoring; the withheld responses are not used.

```{r candidate-plates, results='asis'}
candidate_plates <- subset(screen, role != "training")
candidate_plates <- split(candidate_plates[, prediction_columns], candidate_plates$role)
```

The candidate plates and reference grid can contain new combinations of fitted
entities, but cannot introduce cells or compound-dose treatments absent from the fit.

We pass the candidates and reference grid to `score_plate_pdbal()`, using
`plogis` to compare predictions on the viability scale:

```{r score-plates}
plate_scores <- score_plate_pdbal(
    fit,
    candidate_data = candidate_plates,
    reference_grid = reference_grid,
    response_transform = plogis,
    max_triplets = 1000
)
plate_scores
```

Lower scores are preferred: the plate with `rank = 1` is the recommended next
batch under this criterion. The output identifies each candidate by `plate`,
reports its size in `n_experiments`, and gives its log PDBAL `score` and `rank`.
The scores compare how informative the plates are expected to be about the
reference surface; they do not rank plates by predicted treatment efficacy.

We use `plogis` because differences in viability are assumed to be of scientific
interest here. This is a choice about what the ranking should prioritize:
distances on the viability scale weight differences differently from distances
on the logit scale, so the transformation can affect which plate is preferred.
Use `response_transform = identity` to compare on the fitted scale, or supply
another transformation appropriate to the scientific question. The function must
preserve the prediction matrix dimensions and return finite values; `distance`
also accepts a custom distance function.

## Limitations and next steps

This introduction uses a simple model and a small dataset in order to demostrate
the API of `bathcieR`. For substantive analyses, the users should think more
about how much structure they would like to incorporate to the model and how
intensly they need to explore the posterior.

To construct more complex models, see `vignette("biologically_informed_models", package = "batchieR")`
for a more rigorous treatment of prior information through the global metadata
regression, shrinkage, dose pooling, and GMRF or tree structures. It also shows
how those assumptions propagate to prediction and plate ranking.

For a closer assessment of model fit, see `vignette("model_checking", package = "batchieR")`.
It covers multi-chain agreement in prediction space, out-of-sample calibration,
grouped predictive checks, predictive errors, and checks of dispersion and tails.

The package also provides other posterior interfaces, but non-Gaussian response
families are not currently available. See `?combo_model`, `?combo_gaussian_component`,
`?fit_combo`, `?prior_predict`, `?posterior_predictions`, and `?score_plate_pdbal` for API details.

## References

Tosh, C. et al. (2025). A Bayesian active learning platform for scalable
combination drug screens. *Nature Communications*, 16, 156.
doi: [10.1038/s41467-024-55287-7](https://doi.org/10.1038/s41467-024-55287-7).

## Session information

```{r session-information}
sessionInfo()
```
