---
title: "Introduction to batchieR"
author: "Teo Sakel"
date: "2026-08-04"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Introduction to batchieR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4.5
)
```

## Introduction

`batchieR` fits Bayesian latent-factor models to continuous responses from
combination screens. It separates variation associated with cellular context,
treatment main effects, and treatment interactions, and uses the fitted
posterior to compare possible follow-up plates. This is useful when cells and
compound-dose treatments recur across a screen and information can be shared
across experiments.

This vignette shows how to prepare the required experiment table, understand
the default model, run a small fit, assess it with posterior predictive checks,
predict withheld responses, and rank candidate plates with
posterior-distance-based active learning (PDBAL).

The package currently supports **only a Gaussian response with conjugate priors**.
Binary, binomial, beta, count, and other non-Gaussian response families are not implemented.

## Installation

Install the current development version from GitHub:

```{r install, eval=FALSE}
remotes::install_github("TeoSakel/batchieR")
```

```{r load-package}
library(batchieR)
```

## Model

Screen responses often represent cell viability, bounded between zero and one.
For the Gaussian model, transform these values to an unbounded scale before fitting.
Following the original BATCHIE paper, the suggested transformation is the logit,
which is implemented in R as `qlogis()`. This is not a binomial model but
rather a Gaussian model on the logit scale. Model predictions can then be
transformed back to the viability scale with `plogis()`.

For an observed response involving cell $k$ and treatments $i$ and $j$,
the current observation model is

$$
y_{kij} \mid \eta_{kij}, \tau \sim \mathcal{N}(\eta_{kij}, \tau^{-1}),
$$

where $\tau$ is the observation precision and the mean $\eta_{kij}$ is the linear predictor

$$
\eta_{kij} = \alpha + x_{kij}^{\mathsf T}\beta
    + W^{(0)}_k + V^{(0)}_i + V^{(0)}_j
    + \left(W^{(1)}_k\right)^\top\left(V^{(1)}_i + V^{(1)}_j + V^{(2)}_i\odot V^{(2)}_j\right).
$$

The systematic component $\alpha + x_{kij}^{\mathsf T}\beta$ is expressed in
the observed covariate space supplied by the user: $\alpha$ sets the overall
level and $x_{kij}^{\mathsf T}\beta$ accounts for optional observation, cell,
and compound metadata. The remaining terms capture cell and treatment effects
and their interactions. In particular, the rows of $W^{(1)}$, $V^{(1)}$, and
$V^{(2)}$ represent cells and treatments in a shared $d$-dimensional latent
space learned from the data. Superscripts label component types, subscripts
index entities, and $\odot$ denotes element-wise multiplication.

`combo_model()` encodes this model using the following arguments:

| Model argument | Symbol | Dimension | Role |
|:--|:--|:--|:--|
| `mean` | $\alpha + x_{kij}^{\mathsf T}\beta$ | scalar per observation | Overall level and optional metadata regression |
| `cell_offset` | $W^{(0)}$ | $K$ | Baseline differences among cells |
| `cell_factors` | $W^{(1)}$ | $K \times d$ | Latent characteristics of cells |
| `treatment_offset` | $V^{(0)}$ | $T$ | Treatment effects shared across cells |
| `treatment_main_factors` | $V^{(1)}$ | $T \times d$ | Cell-dependent treatment main effects |
| `treatment_interaction_factors` | $V^{(2)}$ | $T \times d$ | Cell-dependent nonadditive combination effects |

Here $K$ is the number of cells, $T$ is the number of treatments (compound-dose pairs),
and $d$ is the common latent dimension shared by the cell and treatment representations.
The `rank` argument of `combo_model()` controls $d$. For monotherapies (single drug treatments),
the second-treatment terms and the treatment-interaction term are omitted.

### Systematic mean

The default `mean = empirical_mean()` follows BATCHIE: it fixes $\alpha$ at the
mean of the non-missing fitting responses and $\beta$ at zero. The intercept is
computed once and is not updated by the sampler. `fixed_mean(value)` instead
fixes it at a supplied value, as in this example. For covariate effects through
`formula_mean()`, see the `vignette("biologically_informed_models", package = "batchieR")`.

### Latent components

The remaining five components play a role similar to *random effects*, capturing
cell and treatment variation beyond the systematic mean. Thus, their Gaussian
priors are centered at zero to express residual variation around the systematic
component. The offsets resemble random intercepts, while the latent factors
describe interactions through products of learned representations.

All five components are specified using `combo_gaussian_component()`, which
separates their dependence structure from their magnitude through two arguments:

- `structure` describes relationships among cells or treatments and their
  relative prior precision. The default, `iid()`, treats entities as independent.
- `shrinkage` controls the magnitude of deviations around zero, determining how
  strongly component values are pulled toward zero.

Setting a component argument to `NULL` removes its term from the predictor.
Treatment factor components require `cell_factors` because they enter the
predictor through their product with $W^{(1)}_k$. For more detail on using known
biological relationships to specify `structure`, see the
`vignette("biologically_informed_models", package = "batchieR")`.

### Default specification

Calling `combo_model()` constructs the default rank-12 model. Printing it
shows the response family, rank, mean specification, dose model, and the
structure and shrinkage assigned to each component.

```{r model-specification}
combo_model()
```

The default priors follow the BATCHIE model. The components use independent
(`iid()`) structures and the following priors:

- `cell_offset` uses gamma precision shrinkage.
- `cell_factors` uses multiplicative-gamma shrinkage.
- `treatment_offset`, `treatment_main_factors`, and
  `treatment_interaction_factors` use horseshoe shrinkage.
- The Gaussian observation precision has a gamma prior.

The default dose specification, `categorical()`, treats each exact
compound-dose pair as a separate treatment, with no dose ordering,
interpolation, or automatic sharing between doses.

For simplicity in this vignette, we use rank 2 and fix the systematic mean at
1.5 for both simulation and fitting, keeping the remaining defaults:

```{r example-model}
model <- combo_model(rank = 2, mean = fixed_mean(1.5))
model
```

Use `update()` to change selected settings while retaining the rest of a model
specification. Components are replaced in full, and `NULL` disables a component.
The original specification is unchanged; fit the result with `fit_combo()` when
ready.

```{r update-model-specification}
revised_model <- update(model, rank = 4, treatment_interaction_factors = NULL)
revised_model
```

## Generating synthetic data

We illustrate the workflow with three cells (`A`, `B`, and `C`) and four
drugs (`W`, `X`, `Y`, and `Z`), each used at a single dose. We organize the
synthetic experiments into seven plates:

- **Monotherapy (plate 1):** each drug is tested alone in each cell, giving 12 experiments.
- **Combination training (plates 2–4):** each plate tests `X+Y` and `Z+W` in
  cells A and B, giving four experiments per plate.
- **Holdout (plate 5):** four independent replicate measurements of the same
  cell–pair combinations as the training plates. Their responses are withheld
  from fitting and reserved for evaluating predictions of repeat experiments.
- **Candidate (plates 6–7):** experiments available to run next, ranked by
  their expected informativeness. Their simulated responses are withheld from
  fitting to represent measurements that have not yet been collected.

The two candidate plates explore different gaps in the training data. Each
contains two distinct cell–pair combinations measured twice, for four rows:

- **Known pairs in a new cell context (plate 6):** `X+Y` and `Z+W`, each
  measured twice in cell C. Cell C has monotherapy training data, but no
  observed combination responses.
- **New pairs in known cell contexts (plate 7):** `X+Z` in cell A and `Y+W`
  in cell B, each measured twice. Both cells have combination-training data,
  but neither drug pair has been observed in training.

We also include the holdout plates' design in the follow-up candidates
to act as an "uninformative" baseline.

Once we specify the experimental design, then can simulate responses from a
model using `prior_predict()`.

```{r synthetic-data}
cells <- c("A", "B", "C")
drugs <- c("X", "Y", "Z", "W")

single_drug <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    drug_2 = NA_character_,
    plate = "1",
    stringsAsFactors = FALSE
)

known_pairs <- data.frame(
    cell = c("A", "A", "B", "B"),
    drug_1 = c("X", "Z", "X", "Z"),
    drug_2 = c("Y", "W", "Y", "W"),
    stringsAsFactors = FALSE
)

# training_pairs = known_pairs in triplicates
training_pairs <- do.call(rbind, replicate(3, known_pairs, simplify = FALSE))
training_pairs$plate <- as.character(rep(2:4, each = nrow(known_pairs)))
holdout_pairs <- known_pairs
holdout_pairs$plate <- "5"

candidate_new_cells <- data.frame(
    cell = rep("C", 4),
    drug_1 = c("X", "Z", "X", "Z"),
    drug_2 = c("Y", "W", "Y", "W"),
    plate = "6",
    stringsAsFactors = FALSE
)
candidate_new_pairs <- data.frame(
    cell = c("A", "A", "B", "B"),
    drug_1 = c("X", "X", "Y", "Y"),
    drug_2 = c("Z", "Z", "W", "W"),
    plate = "7",
    stringsAsFactors = FALSE
)
screen <- rbind(
    single_drug,
    training_pairs,
    holdout_pairs,
    candidate_new_cells,
    candidate_new_pairs
)
plate_roles <- c(
    "1" = "training", "2" = "training", "3" = "training",
    "4" = "training", "5" = "holdout", "6" = "new_cell",
    "7" = "new_drug_pairs"
)
screen$role <- unname(plate_roles[screen$plate])
screen$dose_1 <- 1
screen$dose_2 <- ifelse(is.na(screen$drug_2), NA_real_, 1)
```

`prior_predict()` draws one set of model parameters from their priors and
generates responses for the supplied design. We fix the systematic mean at
1.5^[in logit scale, `r round(plogis(1.5), 2)` viability] and generate a single
draw including observation noise.

```{r simulate-responses}
simulated_response <- prior_predict(
    model,
    screen,
    draws = 1,
    type = "response",
    seed = 20260725
)
modeled_response <- as.numeric(simulated_response[1, ])
screen$viability <- plogis(modeled_response)
screen$response <- ifelse(screen$role == "training", modeled_response, NA_real_)

training_cols <- c("cell", "drug_1", "dose_1", "drug_2", "dose_2", "response")
experiments <- screen[training_cols]
screen[c("plate", "role", training_cols, "viability")]
```

The training data must contain `cell`, `drug_1`, `dose_1`, `drug_2`, `dose_2`,
and numeric `response` columns. In this example, `plate` and `role` are
organizational labels not used in the model and are thus omitted from `experiments`.
A missing (`NA`) response keeps a holdout or candidate row available for prediction
without adding it to the likelihood. For holdout rows, we keep `viability` to
evaulate the predictions later. For a single-treatment experiment, the drug and
dose in the absent arm must both be `NA`; partially missing treatment arms are invalid.

## Fit the model

To sample from the posterior distribution, we use `fit_combo()` with the `model`
and `experiments` objects constructed above. A Gibbs sampler is currently the
only sampler implemented. `fit_combo()` uses familiar MCMC sampling controls,
with argument names similar to `CmdStanR`: `chains` sets the number of chains,
while `iter_warmup` and `iter_sampling` control their length.  Here we run two
short chains and set a seed for reproducibility.

```{r fit-model}
fit_chains <- 2
fit_parallel_chains <- min(fit_chains, future::availableCores())

fit <- fit_combo(
    model,
    experiments,
    chains = fit_chains,
    parallel_chains = fit_parallel_chains,
    iter_warmup = 100,
    iter_sampling = 200,
    thin = 1,
    seed = 20260725,
    refresh = if (interactive()) 10 else 0
)
```

The result is a `combo_fit` object containing the posterior draws and the
information needed for prediction and model checking. Printing it gives an
overview of the data and sampling run, followed by compact convergence
diagnostics across the public posterior variables:

```{r print-fit}
fit
```

The diagnostics report the maximum R-hat, minimum bulk and tail ESS, and
counts of unavailable values across public posterior variables. `summary()`
provides parameter estimates, credible intervals, and diagnostics for individual
variables. Here we select the observation scale and precision:

```{r summarize-fit}
summary(fit, variable = c("sigma", "observation_precision"))
```

A `combo_fit` object is also compatible with the standard `broom` verbs:
`tidy()`, `glance()`, and `augment()`. The API is similar to `rstan::tidy()` and
`broom.mixed::tidy()`, with the addition of a `components` argument to select
which model component to summarize. For example, we can extract the posterior
draws of the cell offsets:

```{r tidy-fit}
tidy(fit, components = "cell_offset")
```

## Posterior predictive checks

Posterior predictive checks compare observed logit responses with responses
replicated from the fitted Gaussian model. They use only rows with observed
responses and stay on the scale supplied to `fit_combo()`.

```{r pp-check-density, fig.alt="Density overlay of observed and posterior-predictive logit responses"}
bayesplot::pp_check(fit, type = "dens_overlay", ndraws = 20)
```

```{r pp-check-ecdf, fig.alt="ECDF overlay of observed and posterior-predictive logit responses"}
bayesplot::pp_check(fit, type = "ecdf_overlay", ndraws = 20)
```

Systematic shifts between observed and replicated curves suggest a location
mismatch. Different shapes or tails suggest that the latent structure or the
Gaussian observation model misses important aspects of the data.

For more details, including LOO-PIT calibration, grouped checks, predictive
errors, test statistics, and guidance on interpreting diagnostic patterns, see
`vignette("model_checking", package = "batchieR")`.

### Predicting withheld responses

The holdout plate lets us compare predictions with measurements that were
excluded from fitting. `posterior_epred()` returns draws of the expected response,
while `posterior_predict()` also includes Gaussian observation noise for a
future measurement. Both return a matrix with posterior draws in rows and
experiments in columns, on the fitted logit scale.

Here we use `posterior_predict()` to obtain a median and a 95% predictive
interval for each holdout measurement. We compare these with the simulated
responses saved before masking the holdout rows:

```{r holdout-predictions}
prediction_columns <- setdiff(training_cols, "response")
holdout_rows <- screen$role == "holdout"
holdout_data <- screen[holdout_rows, prediction_columns]

holdout_predictions <- posterior_predict(fit, newdata = holdout_data) |>
    apply(2, quantile, probs = c(0.5, 0.025, 0.975)) |>
    t()
holdout_predictions <- cbind(modeled_response[holdout_rows], holdout_predictions)
colnames(holdout_predictions) <- c("observed", "predicted", "lower", "upper")
holdout_predictions <- cbind(holdout_data, holdout_predictions)
holdout_predictions
```

These intervals describe repeat measurements of training conditions. To
summarize uncertainty in their expected responses instead, use `posterior_epred()`
in the same calculation. To report future measurements on the viability scale,
apply `plogis()` to the predictive draws before summarizing.

## Rank candidate plates with PDBAL

Given a set of candidate plates, which should we run next? Here a *plate* is
a set of experiments performed together, each specifying a cell and one or
two treatments at particular doses. Selecting a plate means selecting the
whole batch of experiments.

We want to choose the plate expected to be most informative about the response
surface. Following BATCHIE, we use PDBAL to rank candidates by their expected
ability to distinguish competing predictions from the fitted posterior. A
`reference_grid` specifies the experiments over which this disagreement
matters: the candidates define what we could measure next, while the grid
defines where we want predictions to improve.

We evaluate the following candidate plates:

- **Plate 5 (holdout):** repeats the four training combinations once each.
  Acts as a baseline of comparison.
- **Plate 6 (new cell context for combinations):** tests the known pairs
  (`X+Y` and `Z+W`) in cell C, which already has monotherapy training data,
  with two measurements of each combination.
- **Plate 7 (New drug-pairs):** tests the new pair `X+Z` in cell A and the new pair
  `Y+W` in cell B, with two measurements of each cell–pair combination.

Only the experimental designs are passed to scoring; the withheld responses are not used.

```{r candidate-plates, results='asis'}
candidate_plates <- subset(screen, role != "training")
candidate_plates <- split(candidate_plates[, prediction_columns], candidate_plates$role)
```

For the reference grid, we include every cell with all six unordered drug pairs,
for 18 combination experiments in total.

```{r reference-grid}
reference_grid <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    dose_1 = 1,
    drug_2 = drugs,
    dose_2 = 1,
    stringsAsFactors = FALSE
)
reference_grid <- subset(reference_grid, drug_1 < drug_2)
reference_grid
```

The candidate plates and reference grid can contain new combinations of fitted
entities, but cannot introduce cells or compound-dose treatments absent from the fit.
Under `categorical()`, a new dose of a known drug is also a new treatment entity.

We pass the candidates and reference grid to `score_plate_pdbal()`, using
`plogis` to compare predictions on the viability scale:

```{r score-plates}
plate_scores <- score_plate_pdbal(
    fit,
    candidate_data = candidate_plates,
    reference_grid = reference_grid,
    response_transform = plogis,
    max_triplets = 1000
)
plate_scores
```

Lower scores are preferred: the plate with `rank = 1` is the recommended next
batch under this criterion. The output identifies each candidate by `plate`,
reports its size in `n_experiments`, and gives its log PDBAL `score` and `rank`.
The scores compare how informative the plates are expected to be about the
reference surface; they do not rank plates by predicted treatment efficacy.  The
preferred choice depends on the fitted posterior and the reference grid, so a
discovery plate need not always outrank a plate of repeated experiments.

We use `plogis` because differences in viability are assumed to be of scientific
interest here. This is a choice about what the ranking should prioritize:
distances on the viability scale weight differences differently from distances
on the logit scale, so the transformation can affect which plate is preferred.
Use `response_transform = identity` to compare on the fitted scale, or supply
another transformation appropriate to the scientific question. The function must
preserve the prediction matrix dimensions and return finite values; `distance`
also accepts a custom distance function.

This flexibility is a general benefit of posterior simulation. Once posterior
draws are available, they can be used to approximate posterior integrals of
functions of the model parameters, whenever those integrals exist. Analysis is
therefore not limited to posterior means: probabilities of events, expected
utilities, and other scientifically meaningful summaries can be computed from
the same draws. Transformations are applied to individual draws before
summarizing, preserving the uncertainty relevant to the chosen scale.

For this PDBAL implementation, `response_transform` acts on the expected
response for each posterior draw at the reference-grid experiments, before
computing distances. Candidate likelihoods and their overlap remain Gaussian on
the fitted logit scale.

## Limitations and next steps

This introduction uses one dose, IID structures, one prior-predictive screen,
and a short sampler run. See `vignette("biologically_informed_models", package = "batchieR")`
for a more rigorous treatment of prior information through the global metadata
regression, shrinkage, dose pooling, and GMRF or tree structures. It also shows
how those assumptions propagate to prediction and plate ranking.

For a closer assessment of model fit, see `vignette("model_checking", package = "batchieR")`.
It covers out-of-sample calibration, grouped predictive checks, predictive errors,
and checks of dispersion and tails, with guidance on interpreting the results.

The package also provides other posterior interfaces, but non-Gaussian response
families are not currently available. See `?combo_model`, `?combo_gaussian_component`,
`?fit_combo`, `?prior_predict`, `?posterior_predictions`, and `?score_plate_pdbal` for API details.

## References

Tosh, C. et al. (2025). A Bayesian active learning platform for scalable
combination drug screens. *Nature Communications*, 16, 156.
doi: [10.1038/s41467-024-55287-7](https://doi.org/10.1038/s41467-024-55287-7).

## Session information

```{r session-information}
sessionInfo()
```
