---
title: "Introduction to batchieR"
author: "Teo Sakel"
date: "2026-08-04"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Introduction to batchieR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4.5
)
```

## Introduction

`batchieR` fits Bayesian latent-factor models to continuous responses from
combination screens. It separates variation associated with cellular context,
treatment main effects, and treatment interactions, and uses the fitted
posterior to compare possible follow-up plates. This is useful when cells and
compound-dose treatments recur across a screen and information can be shared
across experiments.

This vignette shows how to prepare the required experiment table, understand
the default model, run a small fit, assess it with posterior predictive checks,
predict on a reference response surface, and rank candidate plates with
posterior-distance-based active learning (PDBAL).

The package currently supports **only a Gaussian response with an identity link**.
Binary, binomial, beta, count, and other non-Gaussian response families are not implemented.

## Installation

Install the current development version from GitHub:

```{r install, eval=FALSE}
remotes::install_github("TeoSakel/bathieR")
```

```{r load-package}
library(batchieR)
```

## Model

For an observed response involving cell $k$ and treatments $i$ and $j$,
the current observation model is

$$
y_{kij} \mid \eta_{kij}, \tau \sim \mathcal{N}(\eta_{kij}, \tau^{-1}),
$$

where `gaussian_response()` specifies the observation precision $\tau$.
The identity-link linear predictor is

$$
\eta_{kij} = \alpha + W_{0k} + V_{0i} + V_{0j}
    + W_k^\top\left(V_{1i} + V_{1j} + V_{2i}\odot V_{2j}\right).
$$

| Model argument | Symbol | Dimension | Role |
|:--|:--|:--|:--|
| `global_intercept` | $\alpha$ | scalar | Overall response level |
| `cell_offset` | $W_0$ | $K$ | Baseline differences among cells |
| `cell_factors` | $W$ | $K \times d$ | Latent characteristics of cells |
| `treatment_offset` | $V_0$ | $T$ | Treatment effects shared across cells |
| `treatment_main_factors` | $V_1$ | $T \times d$ | Cell-dependent treatment main effects |
| `treatment_interaction_factors` | $V_2$ | $T \times d$ | Cell-dependent nonadditive combination effects |

Here $K$ is the number of cells, $T$ is the number of compound-dose treatment
entities, and $d$ is the latent dimension selected by `rank`. The matrices
$W$, $V_1$, and $V_2$ share this latent dimension.
A single-treatment row has no second-treatment or interaction contribution.
Exchanging the first and second treatment does not change the predictor.

Raw latent coordinates have sign and permutation symmetries. Consequently,
predictions and other identifiable response-scale summaries are generally more
useful to interpret than individual factor coordinates.

### Gaussian responses and transformations

`gaussian_response(link = "identity")` is currently the only supported family and link.
It models the numeric `response` column exactly as supplied. Applying `qlogis()`
to viability does not turn this into a binomial or beta model: it is still a
Gaussian model, now for continuous logit-viability values. Applying `plogis()`
to posterior predictions is a transformation for interpretation; it does not
create a beta or binomial posterior predictive distribution.

## A small synthetic screen

The example is the combination screen used by the package's adaptation
demonstrations. It contains a monotherapy plate, three combination-training
plates, a held-out combination plate, and two candidate combination plates.
The responses come from fixed, visible rank-two parameters. They are
illustrative ground truth, not draws from the priors in `combo_model()`.

```{r synthetic-data}
cells <- c("A", "B", "C")
drugs <- c("X", "Y", "Z", "W")

truth <- list(
    alpha = 1.5,
    cell_offset = c(A = 0.15, B = -0.10, C = 0),
    cell_factors = matrix(
        c(1, 0.2, 0.2, 1, 0.6, 0.6),
        nrow = 3L,
        byrow = TRUE,
        dimnames = list(cells, NULL)
    ),
    treatment_offset = c(X = -0.45, Y = -0.55, Z = -0.40, W = -0.50),
    treatment_main_factors = matrix(
        c(-0.35, -0.05, -0.30, -0.05, -0.05, -0.35, -0.05, -0.30),
        nrow = 4L,
        byrow = TRUE,
        dimnames = list(drugs, NULL)
    ),
    treatment_interaction_factors = matrix(
        c(0.65, 0.05, -0.70, 0.05, 0.05, 0.65, 0.05, -0.70),
        nrow = 4L,
        byrow = TRUE,
        dimnames = list(drugs, NULL)
    )
)

single_rows <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    stringsAsFactors = FALSE
)
single_rows$drug_2 <- NA_character_
single_rows$plate <- "1"

known_pairs <- data.frame(
    cell = c("A", "A", "B", "B"),
    drug_1 = c("X", "Z", "X", "Z"),
    drug_2 = c("Y", "W", "Y", "W"),
    stringsAsFactors = FALSE
)
training_pairs <- do.call(rbind, lapply(2:4, function(plate_id) {
    rows <- known_pairs
    rows$plate <- as.character(plate_id)
    rows
}))
holdout_pairs <- known_pairs
holdout_pairs$plate <- "5"
candidate_confirmatory <- data.frame(
    cell = c("C", "C", "A", "B"),
    drug_1 = c("X", "Z", "X", "Z"),
    drug_2 = c("Y", "W", "Y", "W"),
    plate = "6",
    stringsAsFactors = FALSE
)
candidate_discovery <- data.frame(
    cell = c("A", "A", "B", "B"),
    drug_1 = c("X", "Y", "X", "Y"),
    drug_2 = c("Z", "W", "W", "Z"),
    plate = "7",
    stringsAsFactors = FALSE
)
screen <- rbind(
    single_rows[c("cell", "drug_1", "drug_2", "plate")],
    training_pairs,
    holdout_pairs,
    candidate_confirmatory,
    candidate_discovery
)
plate_roles <- c(
    "1" = "training", "2" = "training", "3" = "training",
    "4" = "training", "5" = "holdout", "6" = "candidate",
    "7" = "candidate"
)
screen$role <- unname(plate_roles[screen$plate])
screen$dose_1 <- 1
screen$dose_2 <- ifelse(is.na(screen$drug_2), NA_real_, 1)

linear_predictor <- function(cell, drug_1, drug_2) {
    w <- truth$cell_factors[cell, ]
    eta <- truth$alpha + truth$cell_offset[cell] +
        truth$treatment_offset[drug_1] +
        sum(w * truth$treatment_main_factors[drug_1, ])
    if (!is.na(drug_2)) {
        eta <- eta + truth$treatment_offset[drug_2] +
            sum(w * truth$treatment_main_factors[drug_2, ]) +
            sum(
                w *
                    truth$treatment_interaction_factors[drug_1, ] *
                    truth$treatment_interaction_factors[drug_2, ]
            )
    }
    unname(eta)
}

eta <- mapply(
    linear_predictor,
    screen$cell,
    screen$drug_1,
    screen$drug_2
)
set.seed(20260725L)
screen$viability <- plogis(eta + stats::rnorm(nrow(screen), sd = 0.05))
modeled_response <- qlogis(pmin(pmax(screen$viability, 0.01), 0.99))
screen$response <- ifelse(screen$role == "training", modeled_response, NA_real_)

experiments <- screen[c(
    "cell", "drug_1", "dose_1", "drug_2", "dose_2", "response"
)]
screen[c(
    "plate", "role", "cell", "drug_1", "dose_1", "drug_2",
    "dose_2", "response"
)]
```

The fitting data must contain `cell`, `drug_1`, `dose_1`, `drug_2`, `dose_2`,
and numeric `response` columns. Plate and role are shown for organization but
are not required by `fit_combo()`. A missing response keeps a holdout or
candidate row available for prediction without adding it to the likelihood.
For a single-treatment experiment, the drug and dose in the absent arm must
both be `NA`; partially missing treatment arms are invalid. Clipping avoids the
infinite logits produced by viability values of exactly zero or one.

There are three distinct scales in this example:

1. `viability` is bounded between zero and one.
2. `response` is continuous logit viability and is what the Gaussian model sees.
3. `plogis()` converts expected responses back to viability for interpretation
   or for measuring PDBAL distances on that scale.

## Specify and fit the default model

```{r model-specification}
model <- combo_model(rank = 2L)
model
```

`combo_model()` assigns a `combo_gaussian_component()` to every model term.
Each component separates four ideas:

1. `mean` describes a feature-informed mean,
2. `mean_shrinkage` regularizes its coefficients,
3. `structure` gives the relative precision geometry among entities, and
4. `shrinkage` controls deviations around the mean.

The default components have zero means, so no metadata tables are needed here.

The default `iid()` structure uses independent unit relative precision.
`categorical()` treats each exact compound-dose pair as a separate treatment;
it provides no dose ordering, interpolation, or automatic sharing between
doses. The defaults implement the BATCHIE model as described in the
[paper](https://doi.org/10.1038/s41467-024-55287-7) but with lower latent
dimensionality (2 instead of 100):

- the `cell_offset` uses gamma precision shrinkage,
- `cell_factors` use multiplicative-gamma shrinkage
- `treatment_offset`, `main_factors`, and `interaction_factors` use horseshoe shrinkage.
- the `global_intercept` is the empirical mean of the observed responses,
- the Gaussian *observation precision* has a gamma prior

Independent chains can run concurrently. Use `future::availableCores()` to
respect limits imposed by the operating system, containers, job schedulers,
and R package checks, then cap the worker count at the number of chains. This
falls back to sequential execution when only one core is available.

```{r parallel-resources}
fit_chains <- 2L
fit_parallel_chains <- min(
    fit_chains,
    as.integer(future::availableCores())
)

c(
    chains = fit_chains,
    available_cores = future::availableCores(),
    parallel_chains = fit_parallel_chains
)
```

```{r fit-model}
fit <- fit_combo(
    model,
    experiments,
    chains = fit_chains,
    parallel_chains = fit_parallel_chains,
    iter_warmup = 25L,
    iter_sampling = 25L,
    thin = 1L,
    seed = 20260804L,
    refresh = if (interactive()) 10L else 0L
)
fit
summary(fit)
```

Printing the `combo_fit` reports sampling dimensions and numbers of
observations, cells, and compound-dose treatments. `summary()` reports
posterior quantiles for sampler RMSE and observation precision. These values
are useful basic summaries, but they are not convergence diagnostics.

The deliberately short run keeps this vignette portable and executable. A
substantive analysis should use longer chains and assess convergence and
effective sample sizes before drawing scientific conclusions.

In an interactive session, `refresh = 10L` updates one progress bar per chain
every ten iterations and at warmup and sampling boundaries. Progress is
disabled above only when the vignette is knitted, where dynamic terminal output
would obscure the rendered document. For a larger fit, the same safe pattern is
`parallel_chains = min(chains, future::availableCores())`. A fixed `seed`
produces the same chain draws regardless of the effective worker count.

## Posterior predictive checks

Posterior predictive checks compare observed logit responses with responses
replicated from the fitted Gaussian model. They use only rows with observed
responses and stay on the scale supplied to `fit_combo()`.

```{r pp-check-density, fig.alt="Density overlay of observed and posterior-predictive logit responses"}
set.seed(20260804L)
bayesplot::pp_check(fit, type = "dens_overlay", ndraws = 20L)
```

```{r pp-check-ecdf, fig.alt="ECDF overlay of observed and posterior-predictive logit responses"}
set.seed(20260804L)
bayesplot::pp_check(fit, type = "ecdf_overlay", ndraws = 20L)
```

Systematic shifts between observed and replicated curves suggest a location
mismatch. Different shapes or tails suggest that the latent structure or the
Gaussian observation model misses important aspects of the data.

These pooled overlays are only a first pass. See
`vignette("model-checking", package = "batchieR")` for LOO-PIT calibration,
grouped checks, predictive errors, test statistics, and guidance on interpreting
diagnostic patterns.

## Rank candidate plates with PDBAL

PDBAL compares posterior hypotheses over a user-defined `reference_grid`. Here
the grid contains every cell with all four single treatments and all six
unordered treatment pairs, for 30 experiments in total.

```{r reference-grid}
single_configurations <- data.frame(
    drug_1 = drugs,
    drug_2 = NA_character_,
    stringsAsFactors = FALSE
)
drug_pairs <- t(utils::combn(drugs, 2L))
pair_configurations <- data.frame(
    drug_1 = drug_pairs[, 1L],
    drug_2 = drug_pairs[, 2L],
    stringsAsFactors = FALSE
)
treatment_configurations <- rbind(
    single_configurations,
    pair_configurations
)

reference_grid <- merge(
    data.frame(cell = cells),
    treatment_configurations,
    by = NULL
)
reference_grid$dose_1 <- 1
reference_grid$dose_2 <- ifelse(is.na(reference_grid$drug_2), NA_real_, 1)
reference_grid <- reference_grid[
    c("cell", "drug_1", "dose_1", "drug_2", "dose_2")
]
```

The reference grid is neither training data nor a prior. It identifies the
response surface on which disagreement between posterior draws matters.
Prediction and scoring data may rearrange fitted entities, but cannot introduce
cells or compound-dose treatments absent from the fit. Under `categorical()`, a
new dose of a known drug is also a new treatment entity.

The two candidate plates come directly from the demonstration screen. Plate 6
transfers the known `X/Y` and `Z/W` combinations partly into cell C, while
plate 7 tests new cross-pair combinations in cells A and B. A third, deliberately
redundant plate repeats one observed combination four times.

```{r candidate-plates}
candidate_row <- function(cell, drug_1, drug_2) {
    data.frame(
        cell = cell, drug_1 = drug_1, dose_1 = 1,
        drug_2 = drug_2, dose_2 = 1,
        stringsAsFactors = FALSE
    )
}

prediction_columns <- c("cell", "drug_1", "dose_1", "drug_2", "dose_2")
candidate_plates <- list(
    confirmatory_plate_6 = screen[screen$plate == "6", prediction_columns],
    discovery_plate_7 = screen[screen$plate == "7", prediction_columns],
    redundant_replicates = do.call(
        rbind,
        replicate(
            4L, candidate_row("A", "X", "Y"), simplify = FALSE
        )
    )
)
candidate_plates
```

```{r score-plates}
set.seed(20260804L)
plate_scores <- score_plate_pdbal(
    fit,
    candidate_data = candidate_plates,
    reference_grid = reference_grid,
    response_transform = plogis,
    max_triplets = 1000L
)
plate_scores
```

`plate` is the candidate-list name, `n_experiments` is its row count, `score`
is the ranking-equivalent log PDBAL objective, and `rank` orders plates from
best to worst. Lower scores are preferred. Plate 7 is selected: its cross-pair
combinations better distinguish posterior hypotheses across the reference surface
than the confirmatory or redundant configurations.

The `plogis` transformation applies only when distances between reference-grid
posterior expectations are calculated. Candidate likelihoods and their overlap
remain Gaussian on the fitted logit scale.

## Limitations and next steps

This introduction uses one dose, IID structures, fixed synthetic truth, and a
short sampler run. Hierarchical and other prior adaptations are reserved for a
follow-up vignette. The package also provides other posterior interfaces, but
non-Gaussian response families and a public prior-predictive simulation
interface are not currently available. See
`?combo_model`, `?combo_gaussian_component`, `?fit_combo`,
`?posterior_predictions`, and `?score_plate_pdbal` for API details.

## References

Tosh, C. et al. (2025). A Bayesian active learning platform for scalable
combination drug screens. *Nature Communications*, 16, 156.
doi: [10.1038/s41467-024-55287-7](https://doi.org/10.1038/s41467-024-55287-7).

## Session information

```{r session-information}
sessionInfo()
```
