---
title: "Checking combination-response models"
author: "Teo Sakel"
date: "2026-08-04"
format:
  html:
    toc: true
    toc-depth: 2
    page-layout: article
    html-math-method: mathml
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Checking combination-response models}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4.5
)
```

## Introduction

A fitted `batchieR` model works with familiar R tools for inspecting and sharing
Bayesian results. This vignette follows a small example through:

- [`posterior`](https://mc-stan.org/posterior/) for draw summaries,
- [`bayesplot`](https://mc-stan.org/bayesplot/) for diagnostic plots,
- [`loo`](https://mc-stan.org/loo/) for predictive comparisons, and
- [`broom`](https://broom.tidymodels.org/)-compatible methods for tidy tables.

```{r}
#| label: load-package
#| message: false
library(batchieR)
library(posterior)
library(bayesplot)
library(loo)
library(ggplot2)

set.seed(20260804)
```

## Fit two models

We will use a toy combination screen to demonstrate the package interfaces.
The screen contains three cells (`A`, `B`, `C`) and three compounds (`X`, `Y`, `Z`),
each tested at two doses. It includes single treatments and all three compound
pairs, with three replicate measurements for each condition.

We will fit two simple models to the same responses:

- The **additive model** combines individual treatment effects and a cell offset.
- The **full model** adds an interaction term, allowing the effect of a treatment
  to depend on its partner and the cell.

By design, we generate the responses from the full model, including its
interaction effects and measurement noise. The additive model therefore omits
structure present in the data. This gives us a concrete example for reading the
diagnostic plots and comparison tables produced by the companion packages.

First, define the two models. Both use the same simple Gaussian priors for
their shared components and a fixed intercept of 0.7.

```{r}
#| label: define-models
iid_component <- combo_gaussian_component(
    structure = iid(),
    shrinkage = gamma_precision(shape = 3, rate = 3)
)
full_model <- combo_model(
    rank = 1,
    mean = fixed_mean(0.7),
    cell_offset = iid_component,
    cell_factors = iid_component,
    treatment_offset = iid_component,
    treatment_main_factors = NULL,
    treatment_interaction_factors = iid_component
)
full_model

additive_model <- combo_model(
    rank = 1,
    mean = fixed_mean(0.7),
    cell_offset = iid_component,
    treatment_offset = iid_component,
    cell_factors = NULL,
    treatment_main_factors = NULL,
    treatment_interaction_factors = NULL
)
additive_model
```

Next, construct the screen and use `prior_predict()` to generate one set of
responses from the full model's prior.

```{r}
#| label: diagnostic-data
#| eval: false
cells <- c("A", "B", "C")
drugs <- c("X", "Y", "Z")
drug_pairs <- c("X:Y", "X:Z", "Y:Z")
doses <- 1:2

screen <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    dose_1 = doses,
    drug_2 = c(NA_character_, drugs),
    dose_2 = c(NA_real_, doses),
    replicate = 1:3,
    stringsAsFactors = FALSE
)
screen <- subset(
    screen,
    (is.na(drug_2) & is.na(dose_2)) |
        (sprintf("%s:%s", drug_1, drug_2) %in% drug_pairs & !is.na(dose_2))
)
screen$type <- ifelse(is.na(screen$drug_2), "single", "combination")
screen$plate <- paste0(screen$type, "_", screen$replicate)
screen$response <- prior_predict(full_model, screen, draws = 1)[1, ]
```

For reproducibility of this vignette, we use a pre-generated copy of this screen
included with the package. The following chunk replaces `screen` with that saved
dataset, so the examples below use the same observations on every machine.

```{r}
#| label: load-pregenerated-screen
screen <- read.csv(
    system.file("extdata", "model_checking.csv", package = "batchieR", mustWork = TRUE),
    stringsAsFactors = FALSE
)
doses <- sort(unique(c(screen$dose_1, screen$dose_2[!is.na(screen$dose_2)])))
```

Finally, fit both models to these same simulated observations.

```{r}
#| label: diagnostic-fits
full_fit <- fit_combo(
    full_model,
    screen,
    chains = 4,
    iter_warmup = 500,
    iter_sampling = 500,
    refresh = 0
)
additive_fit <- fit_combo(
    additive_model,
    screen,
    chains = 4,
    iter_warmup = 500,
    iter_sampling = 500,
    refresh = 0
)
```

## Inspect sampling with posterior and bayesplot

Printing a fit reports the largest R-hat and smallest effective sample sizes
(ESS) across its public posterior variables. Start here for both models.

```{r}
#| label: convergence-overview
print(full_fit)
print(additive_fit)
```

- **R-hat** compares sampling chains. Values close to one are desirable;
  elevated values suggest that chains have not explored the same distribution.
- **Bulk and tail ESS** indicate how much independent sampling information is
  available for central summaries and tail estimates. Larger values mean more
  precise summaries; small values call for more sampling or investigation.
- **Unavailable diagnostics** need inspection. Fixed parameters can have `NA`
  diagnostics, so these should not be counted as evidence of convergence.

The `full_fit`'s overview has an elevated maximum R-hat and low minimum ESS.

### Summarize selected parameters

For more fine-grained inspection, `summary()` and `as_draws_array()` provide
posterior summaries and diagnostics for selected parameters. `summary()` reports
estimates, uncertainty, and diagnostics by parameter. Use `parameter_map()` to
find names; here we select the observation standard deviation (`sigma`) and the
cell offsets as an example.

```{r}
#| label: convergence-summary
full_parameters <- parameter_map(full_fit)
cell_offset_params <- with(
    full_parameters,
    variable[component == "cell_offset" & parameter == "value"]
)
diagnostic_variables <- c("sigma", cell_offset_params)
summary(full_fit, select = diagnostic_variables)
```

- `mean` and `median` summarize each parameter,
- `sd` describes its posterior uncertainty, and
- `q5` to `q95` delimit a central 90% credible interval.

Read `rhat`, `ess_bulk`, and `ess_tail` alongside these estimates to assess
sampling precision.

As an alternative to `summary()`, the broom-compatible `tidy()` method returns
parameter summaries and diagnostics in a tibble that can be filtered, plotted,
or included in reports. `batchieR` exports the generic, so loading `broom` is
optional for this call.

```{r}
#| label: tidy-summaries
parameter_summary <- tidy(full_fit, select = diagnostic_variables)
parameter_summary[, c(
    "term", "estimate", "std.error", "conf.low", "conf.high",
    "rhat", "ess_bulk", "ess_tail"
)]
```

`estimate` is the posterior mean by default, and `conf.low` and `conf.high`
are the bounds of a 95% credible interval. Here `std.error` is the posterior
standard deviation, not the Monte Carlo error of the estimate. The diagnostic
columns have the same meanings as above; the full tibble also carries
parameter metadata. The [broom documentation](https://broom.tidymodels.org/)
introduces the tidy-output conventions, and the
[introduction vignette](introduction.html) covers `glance()` and `augment()`.

For a standard `posterior` object, call `as_draws_array()`. It preserves
iterations and chains and can be passed directly to `summarize_draws()`:

```{r}
#| label: posterior-summaries
selected_draws <- as_draws_array(full_fit, select = diagnostic_variables)
summarize_draws(selected_draws, default_convergence_measures())
```

See the `posterior` documentation for
[customizing summaries](https://mc-stan.org/posterior/reference/draws_summary.html)
and [available diagnostics](https://mc-stan.org/posterior/reference/diagnostics.html).

### Plot the sampling chains

The same draw object works with `bayesplot::mcmc_trace()`.

```{r}
#| label: trace-full
#| fig-alt: "Sampling traces for the observation scale and cell offsets, colored by chain"
mcmc_trace(selected_draws)
```

Each color represents a chain. Look for

- overlapping traces moving around a stable range
- persistent separation
- drift, or
- long flat stretches

These plots cover only the selected parameters and exclude warmup.
They do not override problems reported elsewhere in the fit's overview.

## Check agreement in prediction space

Standard R-hat and ESS summaries are necessary, but latent-factor models require
one additional distinction:

- **Factor non-identifiability** changes raw factor coordinates without changing
  the response surface.
- **Posterior uncertainty** is variation among draws within the distribution a
  chain is exploring.
- **Within-chain drift** means successive parts of a chain have not stabilized.
- **Predictive separation** means stable chains imply different expected
  responses. This is the disagreement that ordinary pooling can conceal.

The public `chain_projections()` helper compares chains through expected
responses rather than raw factor coordinates. The example below uses a complete
24-cell, eight-drug, three-dose factorial design with four replicates. Its truth
is bounded, all six dimensions are active, and the response model is Gaussian.
The favorable design deliberately removes sparse coverage and absent
replication as explanations for disagreement.

### Reproduce the fitted example

The initial run uses ten chains with 2,000 warmup and 32,000 post-warmup
transitions each. For a tractable installed vignette, the shipped teaching fit
keeps three chains from the largest predictive basin and two stable departures,
with 400 evenly spaced draws from each full trajectory. The package ships only
the resulting synthetic data and `combo_fit`; all projection, clustering, and
plotting code that follows is evaluated while the vignette is rendered.
is evaluated while the vignette is rendered.

```{r}
#| label: generate-prediction-space-fit
#| eval: false
#| code-fold: true
#| code-summary: "Generate and fit the controlled example"

rank <- 6L
dimension_weights <- 0.85^(0:(rank - 1L))
treatment_key <- function(drug, dose) {
  paste(drug, sprintf("%.17g", dose), sep = "\034")
}
bounded_direction <- function(n, dimensions = rank) {
  value <- matrix(pmax(-0.8, pmin(0.8, rnorm(n * dimensions))),
                  nrow = n)
  value <- sweep(value, 2, colMeans(value), "-")
  sweep(value, 2, sqrt(colMeans(value^2)), "/")
}
truth_terms <- function(data, truth) {
  cell <- match(data$cell, truth$cells)
  a <- match(treatment_key(data$drug_1, data$dose_1),
             truth$treatments$key)
  has_b <- !is.na(data$drug_2)
  b <- rep(NA_integer_, nrow(data))
  b[has_b] <- match(
    treatment_key(data$drug_2[has_b], data$dose_2[has_b]),
    truth$treatments$key
  )
  W <- truth$cell_factors[cell, , drop = FALSE]
  V1a <- truth$treatment_main_factors[a, , drop = FALSE]
  V1b <- V2b <- matrix(0, nrow(data), rank)
  V1b[has_b, ] <- truth$treatment_main_factors[b[has_b], ,
                                                drop = FALSE]
  V2a <- truth$treatment_interaction_factors[a, , drop = FALSE]
  V2b[has_b, ] <- truth$treatment_interaction_factors[b[has_b], ,
                                                       drop = FALSE]
  cell_offset <- truth$cell_offset[cell]
  treatment_offset <- truth$treatment_offset[a]
  treatment_offset[has_b] <- treatment_offset[has_b] +
    truth$treatment_offset[b[has_b]]
  list(
    cell_offset = cell_offset,
    treatment_offset = treatment_offset,
    main_dimension = W * (V1a + V1b),
    interaction_dimension = W * V2a * V2b
  )
}

cells <- sprintf("cell_%02d", 1:24)
treatments <- expand.grid(
  drug = sprintf("drug_%02d", 1:8), dose = 1:3,
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
treatments$key <- treatment_key(treatments$drug, treatments$dose)
single_index <- expand.grid(
  cell = cells, treatment = seq_len(nrow(treatments)),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
singles <- data.frame(
  cell = single_index$cell,
  drug_1 = treatments$drug[single_index$treatment],
  dose_1 = treatments$dose[single_index$treatment],
  drug_2 = NA_character_, dose_2 = NA_real_
)
edges <- combn(seq_len(nrow(treatments)), 2)
edges <- edges[, treatments$drug[edges[1, ]] !=
                 treatments$drug[edges[2, ]]]
pair_index <- expand.grid(
  cell = cells, edge = seq_len(ncol(edges)),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
a <- edges[1, pair_index$edge]
b <- edges[2, pair_index$edge]
pairs <- data.frame(
  cell = pair_index$cell,
  drug_1 = treatments$drug[a], dose_1 = treatments$dose[a],
  drug_2 = treatments$drug[b], dose_2 = treatments$dose[b]
)
design <- rbind(singles, pairs)
stopifnot(nrow(design) == 6624L)

set.seed(20260922L)
truth <- list(
  cells = cells,
  treatments = treatments,
  weights = dimension_weights,
  cell_offset = as.numeric(bounded_direction(length(cells), 1L)),
  cell_factors = sweep(
    bounded_direction(length(cells)), 2, sqrt(dimension_weights), "*"
  ),
  treatment_offset = as.numeric(
    bounded_direction(nrow(treatments), 1L)
  ),
  treatment_main_factors = sweep(
    bounded_direction(nrow(treatments)), 2,
    sqrt(dimension_weights), "*"
  ),
  treatment_interaction_factors = sweep(
    bounded_direction(nrow(treatments)), 2,
    dimension_weights^0.25, "*"
  )
)
targets <- c(
  cell_offset = 0.3, treatment_offset = 0.6,
  main = 0.8, interaction = 0.8, sigma = 0.5
)
initial <- truth_terms(design, truth)
truth$cell_offset <- truth$cell_offset * targets["cell_offset"] /
  sd(initial$cell_offset)
truth$treatment_offset <- truth$treatment_offset *
  targets["treatment_offset"] / sd(initial$treatment_offset)
truth$treatment_main_factors <- truth$treatment_main_factors *
  targets["main"] / sd(rowSums(initial$main_dimension))
truth$treatment_interaction_factors <-
  truth$treatment_interaction_factors *
  sqrt(targets["interaction"] /
       sd(rowSums(initial$interaction_dimension)))
scaled <- truth_terms(design, truth)
truth$intercept <- median(
  -scaled$cell_offset - scaled$treatment_offset -
    rowSums(scaled$main_dimension) -
    rowSums(scaled$interaction_dimension)
) + 0.15426520414136
truth$precision <- 1 / targets["sigma"]^2

truth_mean <- function(data, truth) {
  terms <- truth_terms(data, truth)
  truth$intercept + terms$cell_offset + terms$treatment_offset +
    rowSums(terms$main_dimension) +
    rowSums(terms$interaction_dimension)
}
generate_responses <- function(data, truth, seed,
                               rng = "Mersenne-Twister") {
  RNGkind(rng, "Inversion", "Rejection")
  set.seed(seed)
  result <- data
  mean <- truth_mean(data, truth)
  result$response <- rnorm(nrow(data), mean, 1 / sqrt(truth$precision))
  result
}
training_seeds <- c(41100L, 80001L, 80002L, 80003L)
training <- do.call(rbind, lapply(seq_along(training_seeds), function(i) {
  result <- generate_responses(design, truth, training_seeds[i])
  result$replicate <- i
  result
}))
holdout <- generate_responses(
  design, truth, 20260924L, rng = "L'Ecuyer-CMRG"
)

treatment_prior <- combo_gaussian_component(
  shrinkage = horseshoe(global_scale = 1, local_scale = 1)
)
model <- combo_model(
  rank = rank,
  mean = empirical_mean(),
  cell_factors = combo_gaussian_component(
    shrinkage = multiplicative_gamma(
      shape = c(2, rep(3, rank - 1L)), rate = 1
    )
  ),
  treatment_offset = treatment_prior,
  treatment_main_factors = treatment_prior,
  treatment_interaction_factors = treatment_prior
)
full_fit <- fit_combo(
  model, training, chains = 10, parallel_chains = 4,
  iter_warmup = 2000, iter_sampling = 32000, thin = 40,
  seed = 81401
)
subset_combo_fit <- function(object, keep) {
  result <- object
  result$draws <- result$draws[keep]
  result$chain_id <- result$chain_id[keep]
  result$draw_id <- result$draw_id[keep]
  result$iteration <- result$iteration[keep]
  result$sampling$chains <- length(unique(result$chain_id))
  result
}
selected_chains <- c(1L, 3L, 4L, 6L, 10L)
keep <- full_fit$chain_id %in% selected_chains &
  full_fit$draw_id %% 2L == 0L
fit <- subset_combo_fit(full_fit, keep)
fit$draw_id <- ave(fit$draw_id, fit$chain_id, FUN = seq_along)
saveRDS(
  list(
    fixture_id = "rank6-horseshoe-bounded-truth-v1",
    design = design, training = training, holdout = holdout,
    truth = truth, truth_mean = truth_mean(design, truth), model = model,
    settings = list(
      transform = "plogis", blocks = 4L, holdout_seed = 20260924L,
      fixture_chains = selected_chains, retained_per_chain = 400L,
      source_retained_per_chain = 800L
    )
  ),
  "inst/extdata/predictive-mixture-inputs-v1.rds",
  compress = "xz", version = 3
)
saveRDS(
  fit, "inst/extdata/predictive-mixture-fit-v1.rds",
  compress = "xz", version = 3
)
```

Load those two inputs, then perform the diagnostic itself. Nothing below this
point is a saved QC result.

```{r}
#| label: load-prediction-space-fit
input_path <- system.file(
  "extdata", "predictive-mixture-inputs-v1.rds",
  package = "batchieR", mustWork = TRUE
)
fit_path <- system.file(
  "extdata", "predictive-mixture-fit-v1.rds",
  package = "batchieR", mustWork = TRUE
)
inputs <- readRDS(input_path)
fit <- readRDS(fit_path)
design <- inputs$design
stopifnot(
  inherits(fit, "combo_fit"),
  identical(sort(unique(fit$chain_id)), c(1L, 3L, 4L, 6L, 10L)),
  all(table(fit$chain_id) == 400L)
)
```

```{r}
#| label: compute-prediction-space-qc
projection <- chain_projections(
  fit, reference_grid = design, response_transform = plogis
)

basin_labels <- function(group, chains) {
  groups <- unique(group)
  counts <- vapply(groups, function(g) sum(group == g), integer(1))
  first <- vapply(groups, function(g) min(chains[group == g]), numeric(1))
  groups <- groups[order(-counts, first)]
  paste0("B", match(group, groups))
}
cluster_projections <- function(projection, threshold = 3) {
  distance <- as.matrix(dist(projection$mean)) /
    sqrt(ncol(projection$mean))
  tree <- hclust(as.dist(distance), method = "average")
  heights <- tree$height
  ratios <- if (length(heights) == 1L) {
    heights / projection$noise_floor
  } else {
    heights[-1L] /
      pmax(heights[-length(heights)], projection$noise_floor)
  }
  gap <- which.max(ratios)
  cut <- Inf
  if (ratios[gap] >= threshold) {
    lower <- if (length(heights) == 1L) {
      projection$noise_floor
    } else {
      max(heights[gap], projection$noise_floor)
    }
    upper <- if (length(heights) == 1L) heights else heights[gap + 1L]
    cut <- sqrt(lower * upper)
  }
  group <- if (is.finite(cut)) {
    cutree(tree, h = cut)
  } else {
    rep(1L, length(projection$chains))
  }
  list(
    tree = tree, distance = distance, cut = cut,
    gap_ratio = ratios[gap], noise_floor = projection$noise_floor,
    membership = data.frame(
      chain = projection$chains,
      basin = basin_labels(group, projection$chains)
    )
  )
}
qc <- list(clustering = cluster_projections(projection))
membership <- qc$clustering$membership
```

### Compare chains with their own block variation

For chain $c$, retained draw $s$, and reference configuration $g$, define

$$
z_{cg} =
\frac{1}{S_c}\sum_{s=1}^{S_c}T(\mu_{csg}),
\qquad
d(c,c') =
\sqrt{\frac{1}{G}\sum_{g=1}^{G}
  \left(z_{cg}-z_{c'g}\right)^2}.
$$

Here $T=\operatorname{logit}^{-1}$ puts expected responses on the viability
scale. Each chain is also divided into four contiguous retained-draw blocks.
The pooled 95th percentile of within-chain block distances is an empirical
noise floor. It can reflect both Monte Carlo variation and drift; it is not a
formal standard error.

```{r}
#| label: prediction-space-dendrogram
#| fig-cap: "Average-linkage clustering of chain-average expected responses. The dotted line is the empirical within-chain block floor and the dashed line is the selected separation cut."
tree <- qc$clustering$tree
plot(tree, xlab = "Chain", sub = "",
     ylab = "RMS distance between prediction projections")
abline(h = qc$clustering$noise_floor, lty = 3, col = "grey40")
if (is.finite(qc$clustering$cut)) {
  abline(h = qc$clustering$cut, lty = 2, col = "#B24C3E")
  rect.hclust(
    tree, k = length(unique(membership$basin)),
    border = grDevices::hcl.colors(
      length(unique(membership$basin)), "Dark 3"
    )
  )
}
```

The split is accepted only when the largest adjacent merge-height ratio is at
least three after bounding the lower height by the block noise floor. In this
fixture the rule finds **`r length(unique(membership$basin))` predictive
basins**. Basin labels summarize chain similarity; they are not posterior mode
probabilities.

A dendrogram alone cannot distinguish stable separation from chains that are
still moving. Compare every block with the full-run center of the largest
basin.

```{r}
#| label: prediction-block-stability
#| fig-cap: "Distance from each contiguous chain block to the largest-basin prediction center. Persistent separated trajectories indicate prediction-relevant non-mixing."
largest_basin <- names(sort(table(membership$basin),
                            decreasing = TRUE))[1L]
largest_chains <- membership$chain[
  membership$basin == largest_basin
]
center <- colMeans(
  projection$mean[
    projection$chains %in% largest_chains, , drop = FALSE
  ]
)
block_stability <- do.call(rbind, lapply(
  seq_along(projection$chains),
  function(i) {
    blocks <- matrix(projection$block_mean[i, , ], nrow = 4L)
    data.frame(
      chain = factor(projection$chains[[i]]),
      block = 1:4,
      distance = sqrt(rowMeans(sweep(blocks, 2, center, "-")^2)),
      basin = membership$basin[
        match(projection$chains[[i]], membership$chain)
      ]
    )
  }
))
ggplot(
  block_stability,
  aes(block, distance, group = chain, colour = basin)
) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:4) +
  theme_minimal() +
  labs(
    x = "Contiguous retained-draw block",
    y = "RMS distance to largest-basin center",
    colour = "Predictive basin"
  )
```

Mean agreement can conceal different uncertainty. The same projection records
the variability of expected responses and observation noise separately.

```{r}
#| label: predictive-spread-by-basin
#| fig-cap: "Chain-level predictive and observation variances by predictive basin. Points are chains; boxes summarize basins with multiple chains."
spread <- data.frame(
  chain = projection$chains,
  basin = membership$basin[
    match(projection$chains, membership$chain)
  ],
  predictive_variance = rowMeans(projection$predictive_variance),
  observation_variance = projection$noise_variance
)
spread_long <- rbind(
  data.frame(
    chain = spread$chain, basin = spread$basin,
    diagnostic = "Mean predictive variance",
    value = spread$predictive_variance
  ),
  data.frame(
    chain = spread$chain, basin = spread$basin,
    diagnostic = "Observation-noise variance",
    value = spread$observation_variance
  )
)
ggplot(spread_long, aes(basin, value, colour = basin)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.25) +
  geom_jitter(width = 0.08, height = 0, size = 2) +
  facet_wrap(~ diagnostic, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Predictive basin", y = "Variance")
```

### Interpret the result

Use the following sequence:

1. Check reconstructed main effects, interaction effects, and other public
   quantities with R-hat and ESS.
2. Compare chain projections with their within-chain blocks.
3. If chains drift across blocks, extend sampling before interpreting clusters.
4. If blocks are stable and all chains occupy one predictive region, ordinary
   pooling is reasonable even when raw factor coordinates disagree.
5. If stable predictive regions remain separated, ordinary equal-chain pooling
   assigns weights according to initialization counts. Continue with the
   [predictive-mixtures vignette](predictive_mixtures.html).

This clustering rule is a deterministic diagnostic heuristic, not a
convergence test. A single detected group cannot exclude unvisited regions,
and agreement of chain means can conceal different predictive spreads or
distribution shapes.

## Check predictions with bayesplot

`bayesplot::pp_check()` accepts a fitted `batchieR` model directly.
It generates replicated responses from that model and compares them with observations.
The `type` argument selects the plot; there is no need to extract predictions
manually for these checks.

### Start with a pooled view

An empirical cumulative distribution function (ECDF) shows the fraction of
responses at or below each value. The dark curve represents observations;
the lighter curves represent replicated datasets.

```{r}
#| label: pooled-ecdf-full
#| fig-alt: "Observed and replicated response ECDFs for the full model"
pp_check(full_fit, type = "ecdf_overlay")
```

```{r}
#| label: pooled-ecdf-additive
#| fig-alt: "Observed and replicated response ECDFs for the additive model"
pp_check(additive_fit, type = "ecdf_overlay")
```

Look for the observed curve to sit among the replicated curves. Consistent
separation indicates a mismatch in location, spread, or tails. Much of the
pooled distribution looks similar here, so these plots alone do not clearly
show which combinations the additive model misses.

### Inspect a combination more closely

Use `newdata` to select observations and `group` to split a plot by a data
column. We focus on `X:Z` in cell `C`, where this example has a clear discrepancy,
and keep all three replicates at each dose combination.

```{r}
#| label: select-combination
combination_data <- subset(screen, cell == "C" & drug_1 == "X" & drug_2 == "Z")
```

In the interval plots, `y` marks observed responses and `yrep` marks predictive
medians. Thick and thin intervals contain 50% and 95% of the replicated
responses, respectively, including measurement noise. Panels distinguish Z's
two doses; within each panel, X's dose is on the horizontal axis. Replicates
share the same dose, so their predictive intervals overlap.

```{r}
#| label: combination-intervals-additive
#| fig-alt: "Additive predictive intervals miss the three low responses at X dose 1 and Z dose 2 in cell C"
pp_check(
    additive_fit, type = "intervals", newdata = combination_data,
    x = "dose_1", group = "dose_2", prob = 0.5, prob_outer = 0.95
) +
    labs(title = "Additive model: X:Z in cell C", x = "X dose", y = "Response") +
    scale_x_continuous(breaks = doses) +
    coord_cartesian(ylim = c(-9, 5))
```

```{r}
#| label: combination-intervals-full
#| fig-alt: "Full-model predictive intervals cover the low responses at X dose 1 and Z dose 2 in cell C"
pp_check(
    full_fit, type = "intervals", newdata = combination_data,
    x = "dose_1", group = "dose_2", prob = 0.5, prob_outer = 0.95
) +
    labs(title = "Full model: X:Z in cell C", x = "X dose", y = "Response") +
    scale_x_continuous(breaks = doses) +
    coord_cartesian(ylim = c(-9, 5))
```

At X dose 1 and Z dose 2, all three observations fall below the additive
model's 95% predictive intervals. The full model reproduces those low responses
while also capturing the higher responses at X dose 2. This is the missing
interaction in practical terms: X's dose effect depends on its partner's dose.
The wider additive intervals do not resolve the mismatch.

See the [bayesplot predictive-check tutorial](https://mc-stan.org/bayesplot/articles/graphical-ppcs.html)
for more plot types and examples of interpreting them.

## Compare predictive performance with loo

`loo::loo()` also accepts a fit directly. It estimates how well a model
predicts each observation when that observation is left out, using
Pareto-smoothed importance sampling (PSIS) to approximate refitting.

```{r}
#| label: compute-loo
full_loo <- loo(full_fit)
full_loo
additive_loo <- loo(additive_fit)
additive_loo
```

The main score is `elpd_loo`, the estimated expected log predictive density;
higher values indicate better predictive performance on the same observations.
`looic` expresses that score as a deviance, so lower values are better.
`p_loo` summarizes effective model complexity. Check the reported standard
errors and Pareto-k diagnostics before interpreting differences.

### Inspect the approximation diagnostics

In these plots, each point is an observation. High Pareto-k values flag rows
for which the approximate leave-one-out calculation may be unreliable.

```{r}
#| label: pareto-k-full
#| fig-alt: "Pareto-k diagnostics by observation for the full model"
plot(full_loo, label_points = TRUE)
```

```{r}
#| label: pareto-k-additive
#| fig-alt: "Pareto-k diagnostics by observation for the additive model"
plot(additive_loo, label_points = TRUE)
```

Inspect observations above the plotted thresholds and read the printed
diagnostic categories; exact leave-one-out refits may be needed for flagged rows.
A high Pareto-k value is a warning about the approximation, not by itself
evidence that one model is better. The
[loo diagnostic reference](https://mc-stan.org/loo/reference/pareto-k-diagnostic.html)
explains thresholds and possible remedies.

### Read the comparison table

```{r}
#| label: compare-loo
loo_compare(list(full = full_loo, additive = additive_loo))
```

The model with the highest estimated ELPD appears first and has `elpd_diff = 0`.
Other rows show their score minus that model's score: a negative `elpd_diff`
means worse estimated performance. `se_diff` is the standard error of that **difference**;
a difference small relative to its standard error gives little reason to prefer one model.

The full model ranks first here, consistent with the predictive plots. However,
the sampling diagnostics make this ranking provisional. Resolve sampling
problems and any Pareto-k warnings before relying on the comparison.  Also, each
held-out unit is one measurement: the other replicates remain in training. This
does not measure prediction for entirely new cells or drug pairs.  See the
[loo walkthrough](https://mc-stan.org/loo/articles/loo2-example.html) and
[`loo_compare()` reference](https://mc-stan.org/loo/reference/loo_compare.html)
for more detail.

In this example, the standard interfaces reveal a combination the additive
model misses and a provisional predictive advantage for the full model. They
also expose sampling issues that need attention before a firm comparison. The
same calls can be used with your own `batchieR` fits.
