---
title: "Prediction from chains in different basins"
author: "Teo Sakel"
format:
  html:
    toc: true
    toc-depth: 2
    page-layout: article
    html-math-method: mathml
    embed-resources: true
    code-fold: false
vignette: >
  %\VignetteIndexEntry{Prediction from chains in different basins}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 8,
    fig.height = 4.5
)
```

```{r}
#| label: libraries

library(batchieR)
library(ggplot2)

artifact <- function(name) {
    fname <- paste0("predictive-mixture-", name, "-v2.rds")
    path <- system.file("extdata", fname, package = "batchieR", mustWork = TRUE)
    readRDS(path)
}
```

## When chains disagree about predictions

MCMC chains can remain in different posterior regions and give different
predictions for the same experiment. When this happens, pooling their draws
can make the combined prediction depend on how many chains reached each region.
We investigate how to recognize this disagreement, assess whether it persists
within each chain, and combine predictions using predictive performance.
Combining predictions does not establish posterior convergence.

We build on the introductory and model-checking vignettes, focusing here on
the consequences of incomplete exploration for prediction.

## Generate synthetic data

We use a rank-6 model with horseshoe treatment shrinkage and an intercept fixed
at zero. Horseshoe shrinkage allows many coefficients to remain near zero
while others are large. Combined with the model's interacting latent factors,
this gives a reason to investigate whether chains settle into different
explanations of the data. Multiple predictive basins are a possibility to
check, rather than a guaranteed consequence of this prior.

We cross 24 cells, eight compounds, and three positive dose levels, including
singles and every pair of distinct compounds. Four replicates provide 26,496
training observations at 6,624 configurations; a fifth supplies independent
holdout responses. Complete coverage and replication let us investigate this
sampling difficulty without missing combinations or a lack of repeated
measurements. Fitting the generating model also removes model misspecification
as an explanation.

Responses come directly from one prior predictive draw, without clipping,
censoring, or screening the generating parameters.

<!-- Fixture provenance: this example retains all ten chains from the completed
horseshoe experiment. The input and fit metadata retain the original
sampling settings and chain selection. -->


```{r}
#| label: specify-model
treatment_prior <- combo_gaussian_component(
    shrinkage = horseshoe(global_scale = 1, local_scale = 1)
)
model <- combo_model(
    rank = 6,
    mean = fixed_mean(0),
    cell_factors = combo_gaussian_component(
        shrinkage = multiplicative_gamma(shape = c(2, rep(3, 5)), rate = 1)
    ),
    treatment_offset = treatment_prior,
    treatment_main_factors = treatment_prior,
    treatment_interaction_factors = treatment_prior
)
```

```{r}
#| label: construct-design

cells <- sprintf("cell_%02d", 1:24)
drugs <- sprintf("drug_%02d", 1:8)

singles <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    dose_1 = 1:3,
    drug_2 = NA_character_,
    dose_2 = NA_real_,
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
)
singles$plate <- paste0("complete_single_", singles$cell)

pairs <- expand.grid(
    cell = cells,
    drug_1 = drugs,
    dose_1 = 1:3,
    drug_2 = drugs,
    dose_2 = 1:3,
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
)
pairs <- subset(pairs, drug_1 < drug_2)
pairs$plate <- paste0("complete_pair_", pairs$cell)

design <- rbind(singles, pairs)
stopifnot(nrow(design) == 6624L)
```

We replicate the design five times and call `prior_predict()` once with
`draws = 1` and `type = "response"`. This samples one shared parameter set and
five conditionally independent responses per configuration. Using `draws = 5`
would instead sample five parameter sets. We use four replicates for training
and one for holdout.

```{r}
#| label: generate-responses
#| eval: false

generating_model <- model
replicated_design <- do.call(rbind, replicate(5, design, simplify = FALSE))
replicated_design[["replicate"]] <- rep(1:5, each = nrow(design))
simulated_response <- prior_predict(
    generating_model,
    replicated_design,
    draws = 1,
    type = "response",
    seed = 20260924
)
replicated_design[["response"]] <- as.numeric(simulated_response[1, ])
training <- subset(replicated_design, replicate <= 4)
holdout <- subset(replicated_design, replicate == 5)

inputs <- list(design = design, training = training, holdout = holdout)
saveRDS(inputs, "inst/extdata/predictive-mixture-inputs-v2.rds")
```

## Fit the model

We fit the same model to the training responses using ten independently
initialized chains, retaining 800 draws per chain after warm-up. This gives
multiple opportunities to explore different posterior regions. We run
generation and fitting offline and save the inputs and fit so rendering the
vignette does not repeat the sampling.

```{r}
#| label: fit-replicated-screen
#| eval: false

fit <- fit_combo(
    model,
    training,
    chains = 10,
    parallel_chains = 4,
    iter_warmup = 2000,
    iter_sampling = 32000,
    thin = 40,
    seed = 81401
)
saveRDS(fit, "inst/extdata/predictive-mixture-fit-v2.rds")
```

Here we reload the saved inputs and fit so the analysis uses the same simulated
data and posterior draws each time the vignette is rendered.

```{r}
#| label: load-fitted-example

inputs <- artifact("inputs")
fit <- artifact("fit")
design <- inputs$design
training <- inputs$training
holdout <- inputs$holdout
```

## The chains do not all agree

The model-checking vignette explains the full QC analysis that needs to follow
every fit. We start with R-hat and effective sample sizes to identify variables
whose sampling needs further investigation.

```{r}
#| label: basin-mixing-diagnostics
#| eval: false

# Placeholder: inspect the completed experiment before selecting variables
# with high R-hat or low bulk/tail ESS and enabling this chunk.
print(fit)
mixing_variables <- character()  # Fill with public names from parameter_map(fit).
if (length(mixing_variables)) {
    summary(fit, select = mixing_variables)
}
```

High R-hat and low multi-chain effective sample sizes can arise when chains
remain in different posterior regions, even if each is locally stable. These
diagnostics motivate checking for predictive basins, but do not establish
multimodality on their own. We compare chains' mean predictions to investigate
whether the mixing problems also affect predictions.

`chain_projections()` averages expected responses over draws at each reference
configuration, giving one vector per chain. We apply `plogis` before averaging
to compare predictions on a bounded scale, as in the introductory vignette.
The function also estimates within-chain variation from four contiguous blocks.

```{r}
#| label: compute-chain-projections
projection <- chain_projections(
    fit, reference_grid = design, response_transform = plogis
)
str(projection)
```

We cluster these vectors into predictive basins: groups of chains with similar
mean predictions, which need not correspond to distinct posterior modes.
The deterministic average-linkage rule below mirrors the rule used internally
by `predictive_mixture()`: accept the largest merge-height gap above the block
noise floor only when its ratio is at least three.

```{r}
#| label: cluster-chain-projections

basin_labels <- function(group, chains) {
    candidates <- unique(group)
    counts <- vapply(candidates, function(g) sum(group == g), integer(1))
    first <- vapply(candidates, function(g) min(chains[group == g]), numeric(1))
    candidates <- candidates[order(-counts, first)]
    paste0("B", match(group, candidates))
}

cluster_projection <- function(projection, threshold = 3) {
    mu <- projection$mean
    chains <- projection$chains
    noise_floor <- projection$noise_floor

    distance <- dist(mu) / sqrt(ncol(mu))
    tree <- hclust(distance, method = "average")

    height <- tree$height
    ratios <- height[-1L] / pmax(height[-length(height)], noise_floor)
    gap <- which.max(ratios)
    cut <- if (ratios[gap] >= threshold) {
        sqrt(height[gap + 1L] * max(height[gap], noise_floor))
    } else {
        max(height)  # should always be the last
    }
    group <- cutree(tree, h = cut)
    list(
        tree = tree,
        distance = distance,
        cut = cut,
        gap_ratio = ratios[gap],
        membership = data.frame(chain = chains, basin = basin_labels(group, chains))
    )
}

clustering <- cluster_projection(projection)
membership <- clustering$membership
basins <- unique(membership$basin)
core <- membership$chain[membership$basin == "B1"]
basin_colours <- setNames(hcl.colors(length(basins), "Dark 3"), basins)
```

```{r}
#| label: fig-mixture-dendrogram
#| fig-cap: "Prediction-space clustering computed from the shipped combo_fit. The dotted line is the within-chain block floor; the dashed line is the selected cut."

plot(
    clustering$tree, xlab = "Chain", sub = "",
    ylab = "RMS distance between mean-response projections"
)
abline(h = projection$noise_floor, lty = 3, col = "grey40")
if (is.finite(clustering$cut)) {
    abline(h = clustering$cut, lty = 2, col = "#B24C3E")
    rect.hclust(
        clustering$tree,
        k = length(unique(membership$basin)),
        border = basin_colours
    )
}
```

The rule identifies **`r length(unique(membership$basin))` predictive basins**.
The largest contains chains **`r paste(core, collapse = ", ")`**.

### Basin stability

Having identified distinct predictive basins, we next ask whether each
chain's mean predictions are stable over the sampled interval. We assess
this by comparing mean-response projections across contiguous parts of each chain.

`chain_projections()` divides each chain into four contiguous blocks of equal
size (up to one draw). With 800 retained draws, these are four blocks of 200.
We compare all six pairs of block mean-response projections on the same
reference grid, using their root-mean-square (RMS) differences. Direct block
comparisons can reveal movement that a distance to a fixed basin center would
miss, including movement around that center at a constant distance.

```{r}
#| label: fig-block-stability
#| fig-cap: "Within-chain block-pair RMS differences relative to the minimum basin-center distance. Points show all six comparisons per chain, coloured by chain; boxes are descriptive summaries. The reference line at 1 is shown only when within the data-driven plotting range."

block_stability <- projection$block_distances
block_stability$basin <- factor(
    membership$basin[match(block_stability$chain, membership$chain)],
    levels = basins
)
chain_basin <- membership$basin[match(projection$chains, membership$chain)]
basin_mean <- t(vapply(
    basins,
    function(basin) colMeans(projection$mean[chain_basin == basin, , drop = FALSE]),
    numeric(ncol(projection$mean))
))
min_basin_dist <- min(dist(basin_mean)) / sqrt(ncol(basin_mean))
block_stability$plot_distance <- block_stability$distance / min_basin_dist

p <- ggplot(block_stability, aes(basin, plot_distance)) +
    geom_boxplot(width = .5, outlier.shape = NA, fill = "grey95") +
    geom_point(
        aes(colour = factor(chain)),
        position = position_jitter(width = .12, height = 0),
        size = 2, alpha = .8
    ) +
    scale_colour_brewer(palette = "Set3") +
    labs(
        x = "Predictive basin",
        y = "RMS difference / minimum basin-center distance",
        colour = "Chain"
    ) +
    theme_minimal()
if (1 <= 1.05 * max(block_stability$plot_distance)) {
    p <- p + geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40")
}
p
```

We average the chain projection vectors within each basin to define its center,
then divide the block distances by the minimum RMS distance between basin
centers. This expresses temporal variation relative to the observed separation
between basins.

Each box summarizes the six within-chain block comparisons per chain in
that basin; point colours identify chains. Values near zero indicate small
temporal differences relative to the minimum basin-center separation. Small
values across all six comparisons for a chain support local stability of its
mean predictions over the sampled interval.

This diagnostic assesses stability of mean predictions, not the full predictive
distributions. Four-block summaries can miss changes within blocks, and RMS
averages can conceal changes at individual configurations. Agreement does not
establish convergence or rule out later transitions between basins.

## Basin stacking

Distinct basins give different predictions for the same configurations, leaving
us to decide how much each should contribute to a combined prediction. With
equally long chains, pooling draws weights basins according to how many chains
reached them. Those counts reflect initialization and sampling dynamics and
need not represent posterior basin probabilities or predictive performance.

For prediction, we can instead treat each basin's predictive distribution as
a candidate component and choose mixture weights by predictive performance.
Stacking chooses weights jointly to maximize the mixture's leave-one-out log
predictive score, allowing basins to contribute where their predictions are
useful. Here we approximate the leave-one-out predictive densities within each
basin using Pareto-smoothed importance sampling (PSIS) on the training observations.

For basin $b$, let $\widehat p_{b,-i}(y_i)$ denote its conditional PSIS
estimate. The weights solve

$$
\widehat w = \arg\max_{w_b\geq0,\;\sum_b w_b=1}
    \sum_{i=1}^{n}\log\left\{\sum_b w_b\widehat p_{b,-i}(y_i)\right\}.
$$

`predictive_mixture()` performs three steps:

1. Clusters chains using agglomerative clustering as shown above.
2. Computes basin-specific PSIS-LOO predictive densities.
3. Optimizes the stacking weights.

It returns a weighted mixture of basin predictive distributions, represented by
`ndraws` selected draws from the fitted chains. The returned object records
which basin each chain belongs to in `membership`, the fitted and represented
weights in `weights`, and the leave-one-out diagnostics in `psis`. We inspect
those weights and diagnostics below.

```{r}
#| label: construct-predictive-mixture
mixture <- predictive_mixture(
    fit,
    reference_grid = design,
    response_transform = plogis,
    ndraws = 400
)
```

The weights table has one row per basin. `fitted` is its optimized predictive
weight; `count` is the number of selected draws and `realized` is
`count / ndraws`, the weight used by the returned empirical mixture.
`allocation_error` is `realized - fitted`; values near zero indicate little
rounding error. `omitted_positive` flags a positive fitted weight that received
no draws. These weights describe contributions to the combined prediction,
not posterior basin probabilities or a ranking of standalone basin performance.

```{r}
#| label: tbl-mixture-weights
#| tbl-cap: "PSIS-LOO stacking weights and their representation by selected draws."
knitr::kable(mixture$weights, digits = 6, row.names = FALSE)
```

The PSIS table reports whether the leave-one-out approximation is dependable
for each basin. `observations` and `draws` give the training-row count and the
draw count used for that basin. `maximum_k` is the largest Pareto-k diagnostic;
`threshold` is the draw-dependent cutoff used here. `flagged` counts rows above
that cutoff or with missing diagnostics. `count_ge_1` and `fraction_over_0_7`
highlight large Pareto-k values; fewer problematic rows is better.
`elpd_loo` is the basin's total estimated leave-one-out log predictive density:
higher is better when the PSIS approximation is reliable. It scores the basin
alone, whereas stacking optimizes the combined prediction.

```{r}
#| label: tbl-mixture-psis
#| tbl-cap: "PSIS diagnostics computed from every observed training row."
knitr::kable(mixture$psis$diagnostics, digits = 4, row.names = FALSE)
```

The PSIS status is **`r if (mixture$psis$reliable) "reliable under the implemented thresholds" else "flagged as unreliable"`**.
A failure weakens the stacking recommendation even when finite weights are
available. Passing does not prove global convergence or discover an unvisited
basin.

## Evaluate prediction on untouched data

The returned `combo_predictive_mixture` supports `predict()`,
`posterior_predict()`, and `log_lik()` for evaluating combined predictions.
Parameter extraction is unsupported because stacking weights target predictive
performance rather than posterior basin probabilities.

The following evaluated helper computes holdout log predictive density, error,
and predictive coverage from the original pooled fit and the fitted mixture.
`log_lik()` returns a draw-by-observation matrix, so predictive density is a
stable log-mean-exp over draws.

```{r}
#| label: tbl-holdout-live
#| tbl-cap: "Fresh-holdout performance, recomputed from the combo_fit and predictive mixture."

log_mean_exp <- function(value) {
    maximum <- apply(value, 2, max)
    maximum + log(colMeans(exp(sweep(value, 2, maximum, "-"))))
}

evaluate_prediction <- function(object, label, holdout, seed = 20260925) {
    resid <- predict(object, newdata = holdout, type = "mean") - holdout$response
    lpd <- log_mean_exp(log_lik(object, newdata = holdout))
    set.seed(seed)
    interval <- object |>
        posterior_predict(newdata = holdout) |>
        apply(2, quantile, probs = c(.05, .95))
    data.frame(
        method = label,
        draws = length(object$draws),
        total_elpd = sum(lpd),
        mean_elpd = mean(lpd),
        rmse = sqrt(mean(resid^2)),
        mae = mean(abs(resid)),
        coverage_90 = mean(
            holdout$response >= interval[1, ] &
                holdout$response <= interval[2, ]
        )
    )
}
metrics <- rbind(
    evaluate_prediction(fit, "Naive equal-chain pooling", holdout),
    evaluate_prediction(mixture, "PSIS-LOO basin stacking", holdout)
)

knitr::kable(metrics, digits = 4, row.names = FALSE)
```

The holdout table compares equal-chain pooling with the stacked mixture on
responses not used to fit the weights.

- `draws` reports the number of draws in each predictive representation.
- Higher `total_elpd` and `mean_elpd` mean better holdout log predictive density.
- Lower `rmse` and `mae` mean smaller prediction errors.
- `coverage_90` is the fraction of responses inside the 90% predictive intervals,
  i.e. 0.90 is the nominal target.


## Guidelines

1. Fit multiple independently initialized chains and perform the usual model
   checks before assessing disagreement in mean-response projections.
2. When predictive basins appear, compare contiguous blocks within each chain
   to assess stability of its mean predictions. Investigate drift before
   treating a basin as a persistent predictive component.
3. When locally stable predictive basins remain, inspect basin-specific PSIS
   diagnostics before interpreting stacking weights, then compare the combined
   predictions with equal-chain pooling on untouched data.
4. Interpret stacking weights as contributions to a predictive mixture, not
   posterior basin probabilities. Stacking does not repair mixing or establish
   convergence.
5. Retain the original `combo_fit` for parameter analyses, subject to its
   convergence limitations; use the mixture to evaluate combined predictions.
